---
name: resampling, tuning, and workflows
---

## Resampling, Tuning, and Validating

We have omitted a big part of the [ML Process]{.hi}: Tuning and Validating models.

As always, we begin with the setup (mostly in case this is not the same week as the other parts)

``` {r setup}
# Packages
library(pacman)
p_load(tidyverse, modeldata, skimr, janitor, tidymodels, magrittr)
# Load credit data
data(credit_data)
credit_df <- credit_data
# Clean up names
credit_df %<>% clean_names()
# Create a dummy variable for our outcome (good credit status)
credit_df %<>% mutate(status_good = 1 * (status == "good")) %>%
    # drop old status variable
    select(-status)
```

---

## Simple Sample Splitting

First simple task: Split the full `credit_df` dataset into a [training (sub)set]{.h} and a [testing (sub)set]{.h}

To split the sample, lets use the `initial_split()` function from the `rsample` package. 

- It ***splits*** a proportion (`prop` argument) of your data into a [training subset]{.h} and leaves the rest for the [testing subset]{.h}. 
    - If you wanted to sample by `strata` you could as well

- It is another tool for you to use (we had used `sample_frac()` previously)
- The output is an `rsplit` object, not a dataframe
    - We will pass the object given by `initial_split()` to the `training()` and `testing()` functions to get the actual subsets as dataframes

[Note: It can be helpful to set a seed before you split the data for reproducing results]

---

## Simple Sample Splitting

``` {r sample-splitting-setup}
# Set the seed
set.seed(123)
# Create the split (80-20 split)
credit_split <- credit_df %>%
    initial_split(prop = 0.8)
# Check the output
credit_split
```

Now we can [impose the split]{.h} by using the `training()` and `testing()` functions. 

---

## Simple Sample Splitting

After we grab the proper split data, we can check the dimensions to make sure it makes sense

``` {r subsetting}
# Grab the training data (Larger Chunk)
credit_train <- credit_split %>% training()
# Grab the testing data (Smaller Chunk)
credit_test <- credit_split %>% testing()
# Check dimensions
dim(credit_df); dim(credit_train); dim(credit_test)
```

---

## Resampling

As you've seen in class, the goal of prediction exercises is good [out-of-sample]{.h} performance

To avoid overfitting, while still allowing for "optimal" flexibility, we use resampling methods. 

We will use the `vfold_cv()` function to set up 5-fold cross validation for our model

`vfold_cv()` takes the following arguments:

- `data`
- `v`, the number of folds
- `repeats` the number of times you want to [repeat]{.h} the cross validation (default is 1)
- `strata` is an optional variable to use for stratified sampling in the folds

Creating the folds looks like:

``` {r folds}
# set seed
set.seed(123)
# 5-fold CV on the training dataset
credit_cv <- credit_train %>% vfold_cv(v = 5)
# Check the output 
credit_cv %>% tidy()
```

---

## Define the Data-Processing Recipe

For each split, we want to [process the data]{.h} (`recipes`) and [fit the model]{.h} (`parsnip`)

[Important!]{.hp} We want to preprocess the training subset [in each split]{.h} separately from the split's subset. This helps us avoid data leakage. 

In everything we have done before, we impute using the entire data. We are supposed to be keeping them totally [separate]{.h}.

What we ***really*** want is to process the datasets independently, but using the same `recipe` and steps

---

## Define the Recipe

Only the `recipe`, no `prep()`, `juice()`, or `bake()`

``` {r recipe-def-final, message=TRUE}
# Data-processing recipe
    # IMPORTANT! NOTICE I AM USING THE TRAINING DATA HERE NOT THE FULL SAMPLE
credit_recipe <- recipe(status_good ~ ., data = credit_train) %>%
    # Mean imputation for numeric predictors
    step_impute_mean(all_predictors() & all_numeric()) %>% 
    # KNN imputation for categorical predictors
    step_impute_knn(all_predictors() & all_nominal(), neighbors = 5) %>%
    # Create dummies for categorical variables
    step_dummy(all_predictors() & all_nominal()) %>%
    # Interactions
    step_interact(~income:starts_with("home"))
# Check the result
credit_recipe
```

---

## Define the Model

Tell it both the [type of model and the desired engine]{.h}

Keeping it simple with `linear_reg()` and the standard `"lm"`

- Feel free to try whatever you want here in the code to test things out

Although not necessary for `"lm"`, we will set the `mode` using `set_mode()`. Just good to get into the habit of things

``` {r model-def-final, message=TRUE}
model_lm <- 
    # Set my model
    linear_reg() %>%
    # Set my mode
    set_mode("regression") %>%
    # Set my engine
    set_engine("lm")
# Check the result
model_lm
```

---

## Putting it Together in a Workflow

We have three different objects floating around

1. The defined CV splits for the training dataset (`credit_cv`)
2. The recipe for processing the data (`credit_recipe`)
3. The model that we want to fit on the trianing data (folds)

Now we need to put them together.

- Repeating the preprocessing for each training and validation split ***could*** be annoying if you did it all by hand. 
- Luckily, we will be uwing the `workflows` package within `tidymodels`

How? Literally by defining a workflow with the `workflow()` function and then pipe in your model and then recipe

``` {r workflow, message=TRUE}
workflow() %>% 
    add_model(model_lm) %>%
    add_recipe(credit_recipe)
```

---

## Putting it Together in a Workflow

The workflow we just wrote defines the model, the data processing, and the engine. 

We will have a couple of options: 

[_1._ Fit a Single Model Without CV]{.hp}

[_2._ Fit a Single Model With CV]{.hp}

[_3._ Tune your model (using CV)]{.hp}

---

## Putting it Together in a Workflow

The workflow we just wrote defines the model, the data processing, and the engine. 

[_1._ Fit a Single Model Without CV]{.hp}

If you [only]{.h} want to fit the model (with no CV), you can pipe the workflow to `fit(data = dataframe)`

- This is not what we will ultimately want as we want to cross-validate using the folds we created

``` {r fit-01}
# Workflow and model fit
workflow() %>%
    add_model(model_lm) %>%
    add_recipe(credit_recipe) %>%
    fit(data = credit_df)
```

---

## Putting it Together in a Workflow

The workflow we just wrote defines the model, the data processing, and the engine. 

[_2._ Fit a Single Model With CV]{.hp}

If you want to estimate your model's out-of-sample performance, then you can use `fit_resamples()`
And we can use `collect_metrics()` function to assess the performance across the resamples

- By default, it will average across the folds (notice it says `n = 5`, which is our folds).

``` {r fit-02}
fit_lm_cv <- 
    workflow() %>%
    add_model(model_lm) %>%
    add_recipe(credit_recipe) %>%
    fit_resamples(credit_cv)
# Check performance
fit_lm_cv %>% collect_metrics()
```

---

## Putting it Together in a Workflow

The workflow we just wrote defines the model, the data processing, and the engine. 

[_2._ Fit a Single Model With CV]{.hp}

- If you wanted to see how the model performs on [each fold]{.h} then you can set `summarize = F` inside `collect_metrics()`

``` {r collect-alt}
fit_lm_cv %>% collect_metrics(summarize = F)
```

---

## Putting it Together in a Workflow

RMSE and $R^{2}$ are the default metrics. We can change them using the `metrics` argument inside of `fit_resamples()`

You can pass whichever metric you want to use from the `yardstick` package. 

- Suppose we want Mean Absolute Error (MAE) as well

``` {r add-MAE}
# Define Workflow, then fit model on folds
fit_lm_cv <- 
    workflow() %>%
    add_model(model_lm) %>%
    add_recipe(credit_recipe) %>%
    fit_resamples(credit_cv, metrics = metric_set(rmse, rsq, mae))
# Check performance
fit_lm_cv %>% collect_metrics()
```

---

## Putting it Together in a Workflow

The workflow we just wrote defines the model, the data processing, and the engine. 

[_3._ Tune your model (using CV)]{.hp}

Instead of "just" estimating out-of-sample model performance for a specified model and pre-processing operations, we will often want to [tune a model's hyperparameter(s)]{.h}

Suppose we are using $k$ nearest neighbors instead of a linear regression. Then we either need to specify a $k$ [or]{.h} we need to [tune k]{.hp}. 

By [tune]{.hp} we mean estimate the out-of-sample performance for various values of $k$ and then choose the "best" $k$ (based on some metric that defines "best").

To begin, let's define a KNN model using `nearest_neighbor()`. We will tell it to [tune]{.hp} by specifiying we want to `tune()` as our neighbors value.

``` {r tune-knn}
# Define our KNN Model
model_knn <- 
    nearest_neighbor(neighbors = tune()) %>%
    set_mode("regression") %>%
    set_engine("kknn") 
# Check the Model
model_knn
```

---

## Tuning KNN

Almost ready to fit the model, but we still need to let R know which values to try for the parameter we are tuning

To tune our hyperparameter(s), we will use the `tune_grid()` function (instead of `fit()` or `fit_resamples()` functions)

`tune_grid()` works similarly to `fit_resamples()` except it takes an additional argument: `grid`

- We will pass the possible values of our hyperparameter(s) to the `grid` argument and it will evaluate each fold of our samples on each set of hyperparameter(s) we fed it

---

## Tuning KNN

KNN only needs one hyperparameter: `neighbors`. 
We can give the `grid` argument a `data.frame` of all the values we would like to try, where the [column's name matches the hyperparameter we are trying to tune]{.h} `neighbors`

- If, like me, you don't know what may be "reasonable" values, you can see what is suggested by the function with the same name as the hyperparameter (`neighbors()` in this case)

``` {r check-neighbors}
# Check "suggested" value of neighbors
neighbors()
```

---

## Tuning KNN

[Caution]{.hi-red} 

As you add more values of hyperparameters, you fit more models.

More models means more computation. 

More computation means more time.

Keep in mind that you are already fitting a model on each fold (5 in our example) so if you want to test 100 values of `neighbors` then you need to fit $100 \times 5$ models

Let's try 1 `neighbor` (surely not the best) and then 5, and then 10 to 100 by 10s

---

## Tuning KNN

Let's try 1 `neighbor` (surely not the best) and then 5, and then 10 to 100 by 10s

``` {r workflow-tuning-knn}
# Define the workflow
workflow_knn <- workflow() %>%
    add_model(model_knn) %>%
    add_recipe(credit_recipe)

# Fit the workflow on our predefined folds and hyperparameters
fit_knn_cv <- workflow_knn %>%
    tune_grid(
        # CV data
        credit_cv, 
        # Grid where we put our values
        grid = data.frame(neighbors = c(1, 5, seq(10,100,10))),
        metrics = metric_set(rmse, rsq, mae)
    )

# Check performance
fit_knn_cv %>% collect_metrics()
```

---

## Tuning KNN

Recognize what we just did was pretty cool: We changed very little to move from a cross-validated regression model to a CV-tuned KNN model.

The output is also cool, but overwhelming. 

We can use the `show_best()` function to have it tell us which value(s) of the hyperparameters performed best according to the metrics we asked for.
We either need to give `show_best()` the name of the `metric` to use, or it will choose one for us. 

- `show_best()` also defaults to showing the top 5 models but you can change that with the `n` argument
- `select_best()` allows us to select the best model (based on the metrics)

``` {r show-best-01}
# Show the best model in terms of RMSE
fit_knn_cv %>% show_best(metric = "rmse",, n = 5)
```

``` {r show-best-02}
# Show the best model in terms of RMSE
fit_knn_cv %>% show_best(metric = "rsq", n = 3)
```

---

## We Can Visualize It 

We can use the `autoplot()` function for a nice and quick picture (we can improve on it too).

We just need to tell it what metric we want to map.

``` {r autoplot-01}
# Quick plot
autoplot(fit_knn_cv, metric = "rmse")
```

---

## Visualize it Part 02

``` {r plot-improve-02}
# Plot metrics' averages across values of 'k'
fit_knn_cv %>% 
    collect_metrics(summarize = T) %>%
    ggplot(aes(x = neighbors, y = mean)) +
    geom_line(size = 0.7, alpha = 0.6) +
    geom_point(size = 2.5) +
    facet_wrap(~ toupper(.metric), scales = "free", nrow = 1) +
    scale_x_continuous("Neighbors (k)", labels = scales::label_number()) +
    scale_y_continuous("Estimate") +
    scale_color_viridis_d("CV Folds:") +
    theme_minimal() + 
    theme(legend.position = "bottom")
```

---

## Visualize it Part 03

``` {r plot-improve-03}
# Show each fold's metrics across values of 'k'
fit_knn_cv %>% 
    collect_metrics(summarize = F) %>%
    ggplot(aes(x = neighbors, y = .estimate, color = id)) +
    geom_line(size = 0.7, alpha = 0.6) +
    geom_point(size = 2.5) +
    facet_wrap(~ toupper(.metric), scales = "free", nrow = 1) +
    scale_x_continuous("Neighbors (k)", labels = scales::label_number()) +
    scale_y_continuous("Estimate") +
    scale_color_viridis_d("CV Folds:") +
    theme_minimal() + 
    theme(legend.position = "bottom")
```