---
name: prediction
---

## Lastly: Prediction!

We are just about finished 

We will wrap it up by taking our selected model (the best model), fitting it on [all]{.h} of the training data, and then predict onto the test data

- Recall we pick our "best" model (using whatever `metric` we prefer) using the `select_best()` function

Once we found the "best" model, we wrap up our `workflow()` using `finalize_workflow()`

- This function wants (1) your initial workflow [and]{.h} (2) your best model

``` {r prediction-01}
# The final workflow for our KNN model
final_knn <-
    workflow_knn %>% 
    finalize_workflow(select_best(fit_knn_cv, metric = "rmse"))
# Check out the final workflow object
final_knn
```

---

## Prediction

Now we fit this finalized workflow to the [full training]{.h} dataset (`credit_train`). 

Enter `fit()`.

``` {r prediction-02}
# Fitting our final workflow
final_fit_knn <- final_knn %>% fit(data = credit_train)
# Examine the final workflow
final_fit_knn
```

---

## Prediction

Finally, we predict onto the [testing data]{.h} by using `predict()`

``` {r prediction-03}
# Predict onto the test data
test_hat <- final_fit_knn %>% predict(new_data = credit_test)
# Check out predictions
test_hat %>% head()
```

---

## Prediction but Easier

Once we have a final workflow (`final_knn` is the workflow plus the selected best model), you can pass it to the `last_fit()` function

Along with the initial split (`credit_split` in our case) to both:

- Fit your final model on your full training dataset

- Make predictions onto the testing dataset (defined by the initial split object)

This `last_fit()` approach streamlines your work and also lets you easily collect metrics using the `collect_metrics()` function from before

``` {r last-fit-01}
# Write over 'final_fit_knn' with this last_fit() approach
final_fit_knn <- final_knn %>% last_fit(credit_split)
# Collect metrics on the test data!
final_fit_knn %>% collect_metrics()
```

---

## Prediction but Easier

If instead you want predictions, just use the `collect_predictions()` on the output from `last_fit()`

``` {r last-fit-02}
# Get test-sample predictions
final_fit_knn %>% collect_predictions() %>% head()
```