---
name: modeling
---

## Fitting Models

Now that we have a nice clean dataset, we can begin to fit a model

In the `tidymodels` universe, we need to do (at least) three things:

[_1_.]{.hi} [Define the Model Specification]{.hp} 

- The [Type]{.hp} of model we want (linear regression, KNN, or random forest)

[_2_.]{.hi} [Define the Engine]{.hp} 

- The [function/package]{.hp} that R will use to fit your model 

[_3_.]{.hi} [Define the data and relationships to fit]{.hp}

- Point R to your data and define output/input

::: {.notes}
If the model specification defines a general class of models that you are using, then engine defines the specific functions/algorithms R will use to fit these models. 

Using the output(s) and input(s) in the dataset
:::

---

## Specifying the Type of Model

Your first decision is the type of model you want to use (in `tidymodels` "type" = "specification", but that may be confusing for the econometric mind)

- You have many choices for the type of model you want to specify. Each comes with a different `parsnip` [function](){target="_blank"}

    - Linear Regression: `linear_reg()`
    - K-Nearest Neighbor: `nearest_neighbor()`
    - Random Forest: `rand_forest()`

- We can check what the linear regression model function `linear_reg()` tells us:

``` {r linear-reg-check}
linear_reg()
```

---

## Setting an Engine

Before we actually fit/train the model, we need to tell R which [engine]{.h} we want to use. 

The [engine]{.h} tells R what algorithms to use, and some algos may differ greatly from others

You define the [engine]{.h} using the `set_engine()` [function](){target="_blank"}

You can see that we can fit a [linear regression]{.h} model using [any of five engines]{.h}

:::: {.columns}

::: {.column}
- `"lm"`
- `"glmnet"`
- `"stan"`
:::

::: {.column}
- `"spark"`
- `"keras"`
:::

::::

You can use `show_engines()` function to see which engines are available for your model type

``` {r show-engines}
# Show which engines are available for linear_reg()
show_engines("linear_reg")
```

---

## Setting an Engine

We will use the `"lm"` engine to fit a traditional linear regression model. Fancy stuff comes later. 

``` {r set-engine}
# Define the "lm" engine for `linear_reg()`
linear_reg() %>% set_engine("lm")
```

---

## Setting an Engine

We can also use `translate()` to see what is going on under the hood a bit more

``` {r set-engine-translate}
# Translate our regression model
linear_reg() %>% set_engine("lm") %>% translate()
```

---

## Fitting (Training) Your Model

After defining the type of model and the engine, we can now begin to `fit()` the model

- The `fit()` [function](){target="_blank"} wants a few arguments

    - The `model_spec` object comes from setting the [type]{.h} and [engine]{.h} (form before)
    - A `formula` the defines the outputs and inputs (just like using `lm()`)
    - The `data`

- In our case, we will have:

    - Model type is `linear_reg()` and our Engine is `"lm"`
    - The `formula` is our linear regression where `status_good` is our outcome and the other vars. `(.)` are the predictors
    - Our data is our cleaned dataset `credit_clean`


This is a simple linear regression but the `tidymodels` way.

``` {r first-tidymodel-reg}
est_reg <- 
    # Model Type
    linear_reg() %>%
    # Engine 
    set_engine("lm") %>%
    # Formula 
    fit(status_good ~ ., data = credit_clean)
```

---

## `tidymodels` Linear Regression

We can use `tidy()` to see the results in a more usual way

``` {r tidy-model}
est_reg %>% tidy()
```

We can also use this information and use `predict()` as we have before. 

---

## Changing The Type of Model

Okay, so we have just gone through a complex way to run a simple regression model that you likely knew how to do in a simpler way

But the power of `tidymodels` is coming from its flexibility. 

- It is very easy to change any of our decisions along the way

Let's say instead we wanted to fit a [KNN]{.h} model. We just need to change the type of model and engine. 

- [Note]{.note .hi} Some methods like [KNN]{.h} can be used for either `"regression"` tasks and `"classification"` tasks. We want to se the [mode]{.h} for these methods, either within the `nearest_neighbor()` function (`nearest_neighbor(mode = "regression)`) or using the `set_mode()` function.

We can finally choose a value for our parameters (tuning and CV later). In [KNN]{.h} that is just setting a value of $k$. I'll arbitrarily choose 6. 

``` {r knn-tidymodel}
est_knn <- nearest_neighbor(mode = "regression", neighbors = 6) %>%
    set_engine("kknn") %>%
    fit(status_good ~ ., data = credit_clean)

est_knn %>% class()
```