---
name: slow and steady
---

## Setup

Let's begin by loading in some packages

``` {r packages}
library(pacman)
p_load(tidyverse, modeldata, skimr, janitor, kknn, magrittr, tidymodels)
```

<br>

For these examples we will use data from the `modeldata` package. 

The dataset is (amazingly) named `credit_data`

``` {r load-data}
# Load credit dataset
data(credit_data)
# (Optional) Assign it a new name
credit_df <- credit_data
```

Remember you should always familiarize yourself with the data you are working with

``` {r data-skim}
# Glimpse at it
credit_df %>% glimpse()
```

---

## Look at Your Data

Ugly data should be cleaned:

``` {r clean-data}
# "Fix" variable names
credit_df %<>% clean_names()
```

And we can take a deeper peak in the data [(we are looking for NAs and distributions)]{.h}

``` {r data-skmir}
credit_df %>% skim()
```

---

## Create Our Outcome Variable

Suppose we want to create an indicator for individuals' whose credit `status` is `"good"`.
Easy enough, can use `tidyverse` to do that.

::: {.solution .fragment}
``` {r credit-good-basic}
# Create a dummy variable for "good" credit status
credit_df %<>% 
    # There are many ways to do this. You could also use ifelse() or case_when().
    mutate(status_good = 1 * (status == "good")) %>% 
    # Drop the old status variable
    select(-status)
```
:::

::: {.fragment}
That's great! We have an outcome that we want to test our model on

<br>

[But we still have some missing data that could be of use]{.hp}

<br>

Here's where `tidymodels` can come in handy in the [preprocessing part]{.h}
:::

::: {.notes}
It is very useful in the modeling and processing part
:::
