---
name: tidymodels
---

## How Does `tidymodels` Help Us?

Imagine we want to clean [each variable]{.h}, then our `tidyverse` approach will take a lot of typing/copying/pasting which may become difficult to track. 

For example, you might want to:

- Standardize your numeric variables
- Create dummies for your categorical variables
- Remove variables that are perfectly determined by other variables (multicollinearity)
- Imput missing values using KNN

> `tidymodels` is a better way! We will use `recipes` and `parsnip` later on to help our modeling

---

## Using `tidymodels` to Preprocess Data {.small}

[Recipes]{.h}

The `tidymodels` workflow begins with [defining a]{.h} [recipe]{.hp}

- This tells R which variable is your [outcome]{.h} and which variable(s) are your [explanatory]{.hi} ones

- Think of it as a defining the formula and dataset within `lm(y ~ x, data = df)` without running the regression 

::: {.fragment}
What we will be doing is try to predict which individuals have `status_good = 1` $\rightarrow$ where `status_good` is our outcome

Two important things before we move on:

[_1._]{.hp} [No functions in your recipe:]{.hp} We are only defining the [roles]{.h} of the data. We will be able to make transformations later

[_2._]{.hp} [All other variables:]{.hp} You can use `.` on the right-hand side of the `~` to say "the rest of the variables". Saves you some time from writing everything down
:::

---

## Recipe

Here is our first `recipe`. It'll help us understand further.

```{r recipe-01, message=TRUE}
# Define the recipe: status_good predicted by all other variables in the dataframe
recipe_all <- recipe(status_good ~ ., data = credit_df)
# What does recipe do?
recipe_all
```

::: {.notes}
Note: this is our training data. This is how the future model will inform itself. 
:::

---

## Recipe

Here is our first `recipe`. It'll help us understand further.

```{r recipe-02, eval=FALSE}
# Define the recipe: status_good predicted by all other variables in the dataframe
recipe_all <- recipe(status_good ~ ., data = credit_df)
# What does recipe do?
recipe_all %>% str()
```

<br>

::: {.callout-tip}
[Note:]{.note} If you have an "ID" variable, you can give it an "ID" role. 

`recipe(...) %>% update_role(id_var, new_role = "ID")`
:::

---

## Recipe

::: {.tiny}
R now ["knows"]{.h} the roles each variable will play in our model. 

Now we can start cleaning/transforming our dataset.

We can apply whatever steps we want to clean/pre-process/manipulate our data. 
The [`tidymodels` universe](https://recipes.tidymodels.org/reference/index.html){target="_blank"} has many possible steps for:
:::

:::: {.columns}
::: {.column .small .align-center}
[Imputation]{.hp} 

- `step_impute_mean()` 
- `step_impute_mode()` 
- `step_impute_knn()` 
- `step_impute_bag()`
:::
::: {.column .small .align-center}
[Transformation]{.hp}

- `step_log()` 
- `step_poly()` 
- `step_mutate()` 
- `step_interact()`
:::
::: {.column .small .align-center}
[Dummies & Discretization]{.hp}

- `step_dummy()` 
- `step_num2factor()` 
- `step_discretize()` 
- `step_cut()`
:::
::: {.column .small .align-center}
[Many More]{.hp}

- `step_center()` 
- `step_normalize()` 
- `step_date()` 
- `step_lag()`
:::
::::

---

## Recipe

To apply any of these, you need to also tell the function `step_*()` [which variables]{.h} you want it to target. 

:::: {.columns}
::: {.column .small}
- [All:]{.hp} `all_vars()`
- [Role:]{.hp} `all_predictors(), all_outcomes(), has_role()`
:::

::: {.column .small}
- [Type:]{.hp} `all_nominal(), all_numeric()`
- [Variable Names:]{.hp} Listing the name of the variables or [using selectors](https://recipes.tidymodels.org/reference/selections.html){target="_blank"} `starts_with(), contains()`, etc.
:::
::::

::: {.align-center}
Try writing a [simple mean imputation for all variables that are]{.h} <br> [(1) predictors AND (2) numeric]{.hp}
:::

::: {.solution .fragment}
``` {r imputation-ex ,message=TRUE}
# Mean imputation for all numeric predictors

recipe_all %>% step_impute_mean(all_predictors() & all_numeric())
```
:::

---

## Recipe

To apply any of these, you need to also tell the function `step_*()` [which variables]{.h} you want it to target. 

:::: {.columns}
::: {.column .small}
- [All:]{.hp} `all_vars()`
- [Role:]{.hp} `all_predictors(), all_outcomes(), has_role()`
:::

::: {.column .small}
- [Type:]{.hp} `all_nominal(), all_numeric()`
- [Variable Names:]{.hp} Listing the name of the variables or [using selectors](https://recipes.tidymodels.org/reference/selections.html){target="_blank"} `starts_with(), contains()`, etc.
:::
::::

::: {.small}
Notice this [is not]{.hp} a processed dataframe. 

To get a processed dataframe, we need to add two more things to our recipe

- `prep()`: estimates the parameters defined by the recipes' preprocessing
- `juice()`: applies the preprocessing to the [training]{.h} dataset (we will talk about CV in `tidymodels` soon)
:::

---

## `prep()` and `juice()`

``` {r prep-ex-01 ,message=TRUE}
# Applying prep()
recipe_all %>%
    step_impute_mean(all_predictors() & all_numeric()) %>%
    prep()
```

The output is still a `recipe`, but now it has [trained]{.h} the preprocessing operations on the [training data]{.hi} and knows the number of observations and number of incomplete rows

---

## `prep()` and `juice()`

Now we add (a dash of) `juice()` to get the preprocessed (cleaned) dataset

``` {r prep-ex-02, message=TRUE}
# Add juice()
credit_clean <- recipe_all %>%
    step_impute_mean(all_predictors() & all_numeric()) %>%
    prep() %>%
    juice()
# Skim it 
credit_clean %>% skim()
```

---

## Success! 🤔

No more [missing values for our numeric predictors]{.h}

However, we still have [some missing values in our]{.hi} [categorical predictors]{.hp}

We just gotta fix those too!

<br>

Let's define our [preprocessing]{.h} as: 

[_1._]{.hp} Impute the numeric predictors using the [means](https://recipes.tidymodels.org/reference/step_meanimpute.html){target="_blank"}

[_2._]{.hp} Impute the categorical predictors using [knn](https://recipes.tidymodels.org/reference/step_knnimpute.html){target="_blank"} (k = 5)

[_3._]{.hp} Create [dummy variables](https://recipes.tidymodels.org/reference/step_dummy.html){target="_blank"} for all categorical predictors

- [Note: by default, `step_dummy()` drops the original column]{.note .tiny}

[_4._]{.hp} [Interact](https://recipes.tidymodels.org/reference/step_interact.html){target="_blank"} the outputted `home_` dummies with `income`

---

## All Together Now

``` {r preprocessing-all}
# Starting from scratch
recipe_all <- recipe(status_good ~ ., data = credit_df)
# Putting it all together
credit_clean <- recipe_all %>%
    # 1. Impute the numerical predictors using the means
    step_impute_mean(all_predictors() & all_numeric()) %>%
    # 2. Impute the categorical variables using KNN (k = 5)
    step_impute_knn(all_predictors() & all_nominal(), neighbors = 5) %>%
    # 3. Create dummies for categorical variables
    step_dummy(all_predictors() & all_nominal()) %>%
    # 4. Interact the created dummies with `income`
    step_interact(~income:starts_with("home")) %>%
    # Prep
    prep() %>%
    # And Juice
    juice()

credit_clean %>% skim()
```

---

## Great Success!

If we did it correctly, after imputing missing values from the training data, we should have:

- No more missing values

- Created dummies 

- Created interactions between `home` and `income`

<br>

[Next, we will explore modeling with]{.align-center .hp}   

[`tidymodels`]{.align-center}
