---
title: 'Lecture .mono[010]'
subtitle: 'Unsupervised learning, dimensionality reduction, and image classification'
author: 'Edward Rubin'
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
exclude: true

```{r, setup, include = FALSE}
library(pacman)
p_load(
  tidyverse,
  data.table,
  tidymodels,
  embed,
  ClusterR,
  future,
  future.apply,
  patchwork,
  scales,
  collapse
)

red_pink   = '#e64173'
turquoise  = '#20B2AA'
orange     = '#FFA500'
red        = '#fb6107'
blue       = '#3b3b9a'
green      = '#8bb174'
grey_light = 'grey70'
grey_mid   = 'grey50'
grey_dark  = 'grey20'
purple     = '#6A5ACD'
slate      = '#23373b'

knitr::opts_chunk$set(
  comment = '#>',
  fig.align = 'center',
  fig.height = 6.75,
  fig.width = 10.5,
  warning = FALSE,
  message = FALSE
)
knitr::opts_chunk$set(dev = 'svg')
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(knitr.table.format = 'html')
xaringanExtra::use_xaringan_extra(c('tile_view', 'fit_screen'))

Sys.setenv('_R_CHECK_LIMIT_CORES_' = FALSE)

workers_n = future::availableCores() |>
  min() |>
  as.integer()
workers_n = max(1L, workers_n - 1L)

if (workers_n > 1L) {
  future::plan(future::multisession, workers = workers_n)
} else {
  future::plan(future::sequential)
}

set.seed(524)

train_dt = fread('train.csv')
test_dt = fread('test.csv')

all_dt = rbindlist(list(train_dt, test_dt), use.names = TRUE, fill = TRUE)
i_train = seq_len(nrow(train_dt))
i_test = (nrow(train_dt) + 1):nrow(all_dt)
kaggle_split = make_splits(
  x = list(analysis = i_train, assessment = i_test),
  data = all_dt
)

train_dt = training(kaggle_split)
test_dt = testing(kaggle_split)

# Use a moderate sample because repeated clustering comparisons are expensive.
analysis_n = min(6000L, nrow(train_dt))
analysis_dt = copy(train_dt[sample(.N, analysis_n)])

pixel_cols = names(analysis_dt)[grepl('^pixel', names(analysis_dt))]
x_pixels = analysis_dt[, ..pixel_cols]

plot_digit = function(row_df, title = NULL) {
  expand_grid(y = 28:1, x = 1:28) |>
    mutate(value = as.numeric(row_df[, pixel_cols, with = FALSE])) |>
    ggplot(aes(x = x, y = y, fill = value)) +
    geom_tile() +
    scale_fill_viridis_c('', option = 'magma', direction = -1) +
    coord_equal() +
    theme_void(base_size = 16, base_family = 'Fira Sans Condensed') +
    labs(title = title) +
    theme(legend.position = 'bottom', plot.title = element_text(hjust = 0.5))
}

kmeans_modal_accuracy = function(x, label, k, seed = 524) {
  set.seed(seed + k)
  km = ClusterR::KMeans_arma(data = x, clusters = k, verbose = FALSE)
  cl = ClusterR::predict_KMeans(data = x, CENTROIDS = km)
  dt = data.table(label = label, cl = cl)
  dt[, pred := collapse::fmode(label), by = cl]
  dt[, mean(label == pred)]
}

wcss = function(x, cl, centers) {
  mu = centers[cl, , drop = FALSE]
  sum(rowSums((as.matrix(x) - mu)^2))
}

future_rbind = function(x, fun) {
  future.apply::future_lapply(
    X = x,
    FUN = fun,
    future.seed = TRUE
  ) |>
    rbindlist()
}
```

---
name: admin
# Admin
### Today

.b[Topics:] We've got a lot (coming to the end)!

- K-means clustering & unsupervised learning;
- PCA and UMAP for dimension reduction;
- why geometry matters for image classification.

### Upcoming

.b[Projects] due this week and next! (Also: Final exam.)

.b[Readings]

- .note[Today:] .it[ISL] Ch. 12 (*Unsupervised Learning*)
- .note[Next:] [NNs and Deep Learning](http://neuralnetworksanddeeplearning.com/) Ch. 1-2; .it[ISL] Ch. 10 (*Deep Learning*).

---
layout: true
# Predicting digits

---

```{r, show-examples-0, echo = FALSE, cache = TRUE}
# Show some examples of the digits
example_idx = c(2, 1, 17, 8, 4, 9, 22, 7, 11)
# Make a plot for each example
plots = lapply(
  X = sample(seq_len(nrow(train_dt)), 96),
  FUN = function(i) {
    plot_digit(train_dt[i])
  }
)
# Wrap the plots into a grid
wrap_plots(plots, ncol = 12) +
  plot_layout(guides = 'collect') &
  theme(
    legend.position = 'none',
    legend.key.width = unit(2, 'cm'),
    plot.title = element_text(size = 15, hjust = 0.5)
  )
```

---
name: intro

## Why digits?

Handwritten digits are a useful .attn[machine-vision sandbox]:

- .hi-pink[784 pixels] per image (28 × 28),
- .hi-orange[10 labels], so this is multiclass prediction,
- .b.purple[easy] to .purple[for us] visualize—and an easy task for humans;
- .b.purple[difficult] to separate .purple[for many ML methods].

.note[Bonus:] Connects .attn[unsup. learning], .attn[dim. reduction], .attn[image/multi-class pred].

We use Kaggle MNIST files: `train.csv` and `test.csv`.

---
name: vision
## Image classification (in one slide)

For .b[MNIST].super.pink[†]

.footnote[.pink[†] MNIST is a classic benchmark for ML. It contains handwritten digits from the US National Institute of Standards and Technology (written by Census employees and HS students).]

- .hi-pink[predictors:] pixel intensities, .grey-vlight[(each x.sub[j]∈[0–255])]
- .hi-orange[target] = the digit label, .grey-vlight[(y∈{0, 1, ..., 9})]

$\implies$ each image is one point in .hi-purple[784-dimensional] predictor space.

--

The target is simple. The .attn[geometry] is the hard part.

.note[Why?] .attn[Combinations] of pixels matter. .grey-vlight[(Not so much individual pixels.)]

---
layout: false
class: clear, middle, center

.b[What do the digits look like?]

```{r, show-examples, echo = FALSE}
example_idx = c(2, 1, 17, 8, 4, 9, 22, 7, 11)
plots = lapply(example_idx, function(i) {
  plot_digit(train_dt[i], title = paste('Label =', train_dt[i, label]))
})

wrap_plots(plots, ncol = 3) +
  plot_layout(guides = 'collect') &
  theme(
    legend.position = 'bottom',
    legend.key.width = unit(2, 'cm'),
    plot.title = element_text(size = 15, hjust = 0.5)
  )
```

---
layout: true

# Predicting digits

---
## Kaggle's Setup

As usual, Kaggle gives us two files

- .b.pink[`train.csv`] has images and labels
- .b.pink[`test.csv`] has images only

Today we use the labeled sample to

1. fit the unsupervised methods,
2. check what structure those methods recover.

---
class: inverse, middle
name: unsupervised

## Supervised *vs.* unsupervised learning

---
## What's the difference?

<br>

.col-left[
.hi-pink[Supervised learning]

- .b[train] with $X$ and $Y$
- .b[goal:] predict $Y$
- .ex[ex:] lasso, RFs, SVMs
]

.col-right[
.hi-purple[Unsupervised learning]

- .b[train] with $X$ only
- .b[goal:] find structure in $X$
- .ex[ex:] clustering & dim. reduction
]

.clear-up[
Today we start on the .b.purple[unsupervised] side, but we will keep asking:<br>.it.pink[Does the structure help prediction?]
<br><br>
.tran25[(In reality, we have a .pink[supervised] task.)]
]

---
class: inverse, middle
name: kmeans

## *K*-means clustering

---
## The general idea

*K*-means is a simple clustering method groups our data into $K$ groups:

- choose a number of .b.pink[clusters], *K*; .grey-vlight[*clusters* = groups of similar obs.]
- assign each observation to (exactly) one cluster $k$; .grey-vlight["nearest" cluster]
- summarize each cluster by its .b.pink[centroid]. .grey-vlight[center cluster's obs.]

*K*-means compresses a messy cloud into $\color{#e64173}{K}$ .attn[representative centers].

--

Now we just need to an algorithm to find the .it[best] clustering...
<br>
which probably means defining .it[best] (.it[i.e.,] an .attn[objective function]).

---
## *K*-means: Objective function

Let cluster $k$ have centroid $\mu_k$. .grey-vlight.it[(The mean location of its obs. in .slab[p] dim.)]

*K*-means chooses clusters $C_1, \ldots, C_K$ to minimize .attn[within-cluster variation]

$$\min_{C_1,\ldots,C_K} \sum_{k = 1}^K \sum_{i \in C_k} \lVert x_i - \mu_k \rVert^2$$

.ex[E.g.,] .attn[within-cluster sum of squares (WCSS)] above.

Interpretation

- reward .hi-pink[tight, compact] clusters, .grey-vlight.it[(short distances add little to the sum)]
- "penalize" observations far from their centroids. .grey-vlight.it[(or "bad" clusters.)]

---
## *K*-means: Algorithm

A general algorithm for clustering:

1. pick $K$ initial .pink[centroids]; .grey-vlight.it[(randomly, or with a heuristic)]
1. .pink[assign] each observation to its nearest centroid; .grey-vlight.it[(Euclidean distance)]
1. .pink[recompute] the centroids; .grey-vlight.it[(mean of the obs. in each cluster)]
1. .pink[repeat] until assignments stop changing. .grey-vlight.it[("convergence")]

  .b.orange[+]  Fast and intuitive.
<br>
  .b.purple[-]  Only guarantees (approx.) a local optimum.

---
## K-means on pixels

.ex[Recall:] Each image (obs.) is one point in .purple[784 dimensions].

There are .b.pink[10 labels]. $\color{#e64173}{K = 10}$ seems like a natural first guess.

... however, $K = 10$ is a guess; not a theorem.

.qa[Q:] Why might $K = 10$ be a bad choice?

```{r, ex-kmeans, include = FALSE, cache = TRUE}
set.seed(524)
kmeans_train = ClusterR::KMeans_arma(
  data = x_pixels,
  clusters = 10,
  verbose = FALSE
)

kmeans_dt = copy(analysis_dt)
kmeans_dt[, cl_k := ClusterR::predict_KMeans(
  data = x_pixels,
  CENTROIDS = kmeans_train
)]

kmeans_dt[, pred_k10 := collapse::fmode(label), by = cl_k]
k10_accuracy = kmeans_dt[, mean(label == pred_k10)]
```

---
## Turning Clusters into Predictions

We have .b.orange[clusters]. .grey-vlight.it[(Groups of "close" images in pixel space.)]

We *want* .b.pink[predictions]. .grey-vlight.it[(Predicted digit labels.)]

- Cluster labels, *i.e.*, $1, 2, ..., K$ are arbitrary. .grey-vlight.it[(they just index the clusters)]
- .note[Solution:] Relabel each cluster by its .attn[modal label]; .grey-vlight.it[(like RF: majority vote)]

.b[Result:] With $K = 10$, this classifier is approx. `r k10_accuracy |> scales::percent(0.1)` accurate..super[.pink[†]]

.footnote[.pink[†] This measure of accuracy is still *in-sample*.]

--

.qa[Q].sub[1] Is this level of accuracy good?
<br>
.qa[Q].sub[2] What would random guessing get us?

---
class: clear, middle, center

.b[Confusion matrix], $K = 10$

```{r, kmeans-conf-mat, echo = FALSE}
# Summarize the confusion matrix for K-means with K = 10
cm_k10 = kmeans_dt[, .N, by = .(label, pred_k10)]
# Fill in the missing combinations of label and pred_k10 with N = 0
cm_k10 = cm_k10[CJ(label = 0:9, pred_k10 = 0:9), on = .(label, pred_k10)]
cm_k10[is.na(N), N := 0]

# Row density: P(prediction | truth)
cm_k10[, dens := N / sum(N)]

ggplot(cm_k10, aes(x = pred_k10, y = label, fill = dens)) +
  geom_tile() +
  scale_fill_viridis_c(
    name = 'Density',
    direction = -1,
    option = 'magma',
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_x_continuous(breaks = 0:9) +
  scale_y_continuous(breaks = 0:9) +
  coord_equal() +
  theme_minimal(base_size = 18, base_family = 'Fira Sans Condensed') +
  labs(
    x = 'Prediction',
    y = 'Truth'
  ) +
  theme(panel.grid = element_blank())
```

---
## Why aren't we doing better?

What is it about the task—and our approach—that isn't working?

We may need more than 10 clusters to capture .b.purple[within-label variation]

- one .orange[label] can contain several .purple[handwriting styles];
- the same .orange[digit] can shift .purple[left, right, up, or down];
- strokes can be .purple[thicker, thinner, or slanted];
- some characters are .purple[quite ambiguous].

So Euclidean distance in raw pixel space is a rough measure of similarity.

---
layout: false
class: clear, middle, center

**Accuracy by label** for $K = 10$ (sensitivity)
```{r, kmeans10-accuracy-by-label, echo = FALSE}
kmeans_dt[, .(
  accuracy = fmean(label == pred_k10) |> scales::percent(accuracy = .1),
  .N |> scales::comma()
), keyby = label] |>
knitr::kable(
  col.names = c('Digit', 'Accuracy', 'Count'),
  align = 'c',
)
```
---
layout: false
class: clear, middle, center

**Separating 7 and 9 is difficult:** *Correct* guesses vs. *incorrect* guesses
```{r, kmeans10-correct7, echo = FALSE}
# Take mean pixel for all the correctly classified 7s
correct_7s = kmeans_dt[label == 7 & pred_k10 == 7, ..pixel_cols]
false_7s = kmeans_dt[label == 7 & pred_k10 != 7, ..pixel_cols]
# Take mean pixel for all the correctly classified 9s
correct_9s = kmeans_dt[label == 9 & pred_k10 == 9, ..pixel_cols]
false_9s = kmeans_dt[label == 9 & pred_k10 != 9, ..pixel_cols]
# Plot the means
p_7true =
  correct_7s |>
  colMeans() |>
  t() |>
  as.data.table() |>
  plot_digit('Correct')
p_7false =
  false_7s |>
  colMeans() |>
  t() |>
  as.data.table() |>
  plot_digit('Mistakes')
p_9true =
  correct_9s |>
  colMeans() |>
  t() |>
  as.data.table() |>
  plot_digit()
p_9false =
  false_9s |>
  colMeans() |>
  t() |>
  as.data.table() |>
  plot_digit()
# Put it all together
wrap_plots(
  list(p_7true, p_7false, p_9true, p_9false),
  ncol = 2
) +
  plot_layout(guides = 'collect') &
  theme(
    legend.position = 'none',
    legend.key.width = unit(2, 'cm'),
    plot.title = element_text(size = 18, hjust = 0.5)
  )
```

---
layout: true
# Predicting digits

---
name: choose-k

## What's $K$?

As you might have guessed, .b.slate[the choice of] $\color{#23373b}{K}$ is a big deal for *K*-means.

- The number of (useful) clusters is usually .b.pink[unknown].
- .note[Common:] The .b.orange["elbow"] & downstream .b.orange[usefulness]. .grey-vlight.it[(marginal gains)]
- The same .b.purple[hyper-parameter tuning] logic we use throughout the course.
  
--

So... we should probably *tune* it.

---
layout: false
class: clear, middle, center

**Finding the elbow.** .grey-vlight.it[You see it, right?]

```{r, elbow-data, include = FALSE, cache = TRUE}
try_k = 1:50

elbow_dt = future_rbind(
  try_k,
  function(k) {
    set.seed(524 + k)
    km = ClusterR::KMeans_arma(data = x_pixels, clusters = k, verbose = FALSE)
    cl = ClusterR::predict_KMeans(data = x_pixels, CENTROIDS = km)
    data.table(k = k, wcss = wcss(x_pixels, cl, km))
  }
)
```

```{r, elbow-plot, echo = FALSE}
ggplot(elbow_dt, aes(x = k, y = wcss)) +
  geom_line(linewidth = .25, color = red_pink, alpha = .8) +
  geom_point(size = 2.4, color = red_pink) +
  theme_minimal(base_size = 18, base_family = 'Fira Sans Condensed') +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(
    x = 'Number of clusters (K)',
    y = 'Within-cluster sum of squares',
  )
```

---
layout: true
# Predicting digits

---
## Elbows and clusters 

.qa[Q] What's the *elbow*?
<br>
.qa[A] The bendy part of the curve, of course!

- WCSS mechanically falls as $K$ rises.
- We're looking for the point of .attn[diminishing marginal gains].
- .note[Caveat] If the bend is weak, the choice of $K$ stays ambiguous.

.grey-light[Elbows are useful (subjective) heuristics—not *proper* decision rules.]

---
layout: false
class: clear, middle, center

**Modal cluster-label accuracy by $K$.** .grey-vlight.it[(Still in-sample.)]

```{r, find-k-pixels, include = FALSE, cache = TRUE}
find_k = future_rbind(
  1:50,
  function(k) {
    data.table(
      k = k,
      accuracy = kmeans_modal_accuracy(x_pixels, analysis_dt$label, k)
    )
  }
)

best_pixel = find_k[which.max(accuracy)]
```

```{r, find-k-pixels-plot, echo = FALSE}
ggplot(find_k, aes(x = k, y = accuracy)) +
  geom_hline(yintercept = 0, linewidth = .25) +
  geom_line(linewidth = .25, color = red_pink, alpha = .8) +
  geom_point(size = 2.5, color = red_pink) +
  theme_minimal(base_size = 18, base_family = 'Fira Sans Condensed') +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    x = 'K',
    y = 'Accuracy',
  )
```

---
layout: true

# Predicting digits

---
## $K$-means takeaways

**Accuracy** depends on $K$, varies by label, and is *meh*.

- With raw pixels, $K = 10$ gives about `r scales::percent(k10_accuracy, accuracy = 0.1)`.
- The best value is around $K$ = `r best_pixel$k` with `r scales::percent(best_pixel$accuracy, accuracy = 0.1)`.
- .attn[Key] .b.pink[10 labels] does .it[not] imply .b.orange[10 clusters].

More clusters can help.

- One true .b.pink[digit] can split into several .hi-purple[styles].
- Extra clusters can improve the modal-label benchmark.<br>.grey-vlight.it[(But can overfit too.)]

---
class: inverse, middle
name: pca

## Now with PCA!

---
## PCA: Intuition

.attn[Principal component analysis (PCA)] is a .orange[dimension-reduction] method that finds new axes to summarize the data..super[.orange[†]]

Same data; different coordinates.

- The **first** PC explains the **most** variation...
- later PCs explain the most remaining *orthogonal* variation.
- We can use PCA to replace *many* pixel-based features with fewer .attn[linear combinations] of pixels..super[.pink[††]]

A relatively .orange[small subset] of PCs can give a compact summary of data.
<br>
$\implies$ PCA *can* .b.orange[reduce dimensions] while .purple[preserving information]!

.footnote[
  .orange[†] PCA is applied in other contexts beyond dimension reduction.
  <br>
  .pink[††] Linear combinations of pixels are like "superpixels" that capture common patterns.
]

---
## PCA: A little math

The first principal component is

$$Z_{1} = \phi_{1,1}X_{1} + \phi_{2,1}X_{2} + \cdots + \phi_{p,1}X_{p}$$

with analogous definitions for the other $p-1$ PCs $Z_2, Z_3, \ldots, Z_p$.

Don't be overwhelmed by the notation.
<br>Each PC is just a weighted combo. of the original features. .grey-vlight.it[(Just like OLS.)]

**Key ideas:**

- PCs are .pink[.b[linear combinations] of the original variables],
- PC.sub[i] is .b.purple[orthogonal] to all other PCs,.super[.purple[†]]
- we usually .b.slate[normalize] variables before PCA.

.footnote[.purple[†] Links: [Nice visualizations](https://setosa.io/ev/principal-component-analysis/) plus another [resource](https://www.nature.com/articles/nmeth.4346).]

---
layout: false
class: clear, middle

**Updating our *K*-means workflow to include PCA**

1. .pink[normalize] the pixels,
1. keep enough PCs to .pink[explain 90% of variance],
1. .pink[run *K*-means] in the PC space.

```{r, ex-pca-wf, eval = FALSE}
# Define a simple PCA recipe
simple_recipe_pca =
  recipe(
    label ~ .,
    data =
      analysis_dt |>
      select(label, starts_with('pixel'))
  ) |>
  step_mutate(label = as.factor(label)) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_predictors(), threshold = .9) |>
  step_rm(starts_with('pixel'))
```

```{r, ex-pca, include = FALSE, cache = TRUE}
simple_recipe_pca =
  recipe(
    label ~ .,
    data = analysis_dt |>
      select(label, starts_with('pixel'))
  ) |>
  step_mutate(label = as.factor(label)) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_predictors(), threshold = 0.9) |>
  step_rm(starts_with('pixel'))

pca_prep = prep(simple_recipe_pca)
train_pc = juice(pca_prep)
n_pc = train_pc |>
  select(starts_with('PC')) |>
  ncol()

set.seed(524)
kmeans_pc10 = ClusterR::KMeans_arma(
  data = train_pc |>
    select(starts_with('PC')),
  clusters = 10,
  verbose = FALSE
)

pc_dt = data.table(
  label = train_pc$label,
  cl = ClusterR::predict_KMeans(
    data = train_pc |>
      select(starts_with('PC')),
    CENTROIDS = kmeans_pc10
  )
)
pc_dt[, pred := collapse::fmode(label), by = cl]
pca_k10_accuracy = pc_dt[, mean(label == pred)]
```

---
class: clear, middle, center

Should we have standardized the pixels with plain *K*-means clustering?

---
layout: true

# Predicting digits

---
## PCA: What is it good for?

So what does PCA do for us here? Subsetting to the PCs that represent 90% of the variance

- leaves us with .hi-orange[`r n_pc` PCs], .grey-vlight.it[(≪ 784)]
- has .b.purple[`r scales::percent(pca_k10_accuracy, accuracy = 0.1)`] accuracy with $K = 10$ clusters. .grey-vlight.it[(< `r scales::percent(k10_accuracy, accuracy = 0.1)`)]
  
--

Wait. So... it's worse?

1. Yes, if we only consider .pink[*training* accuracy].
2. Maybe, if we consider the .orange[predictor space] was `r scales::percent(n_pc / 784)` of the original size.
3. TBD, if we actually check the .purple[test-set accuracy].
   
We have a more *compact* representation, but we *did* lose some information.

---
## What are we doing?

Let's think about what we're actually doing here.

- We are **still running *K*-means**.
- Only the .pink[coordinates] change with PCA.
- Comparable accuracy with far .purple[fewer dimensions] is a decent trade.

We *could* add more PCs. That road eventually leads to the original space.

--

What if we tried a .b.purple[nonlinear] dimension reduction method instead?

---
class: inverse, middle
name: umap

## UMAP .slate[No, *U*MAP!]

---
## UMAP: Intuition

.attn[UMAP].super.pink[†] is a popular .b.purple[nonlinear dimension-reduction method] that can capture more complex structures than PCA.

.footnote[.pink[†] UMAP stands for *Uniform Manifold Approximation and Projection*.<br>Examples [here](https://pair-code.github.io/understanding-umap/); paper's repo [here](https://github.com/lmcinnes/umap).]

- .b.purple[Start:] a high-dim. graph of the data—edges connect nearby obs.;
- .b.pink[Then:] optimized low-dim. graph that preserves local structure.
- .b.purple[Nonlinear:] can uncover *curved* structure that PCA misses.

--

The axes help with geometry—not direct interpretation

- strong for .hi-orange[visualization]; useful for .hi-orange[clustering];
- .b.slate[less interpretable] and more .b.slate[tuning-sensitive] than PCA.

---
layout: false
class: clear, middle

**Updating our *K*-means workflow to include UMAP**

1. .pink[normalize] the pixels
1. embed the data in .pink[.b[2] dimensions] via UMAP
1. cluster with .pink[*K*-means] in the embedded space
   
```{r, ex-umap-wf, eval = FALSE}
simple_recipe_umap =
  recipe(
    label ~ .,
    data = analysis_dt |>
      select(label, starts_with('pixel'))
  ) |>
  step_mutate(label = as.factor(label)) |>
  step_normalize(all_numeric_predictors()) |>
  step_umap(all_predictors(), num_comp = 2) |>
  step_rm(starts_with('pixel'))
```

```{r, ex-umap, include = FALSE, cache = TRUE}
simple_recipe_umap =
  recipe(
    label ~ .,
    data = analysis_dt |>
      select(label, starts_with('pixel'))
  ) |>
  step_mutate(label = as.factor(label)) |>
  step_normalize(all_numeric_predictors()) |>
  step_umap(all_predictors(), num_comp = 2) |>
  step_rm(starts_with('pixel'))

umap_prep = prep(simple_recipe_umap)
train_umap = juice(umap_prep)

set.seed(524)
kmeans_umap10 = ClusterR::KMeans_arma(
  data = train_umap |>
    select(starts_with('UMAP')),
  clusters = 10,
  verbose = FALSE
)

umap_dt = data.table(
  label = train_umap$label,
  cl = ClusterR::predict_KMeans(
    data = train_umap |>
      select(starts_with('UMAP')),
    CENTROIDS = kmeans_umap10
  )
)
umap_dt[, pred := collapse::fmode(label), by = cl]
umap_k10_accuracy = umap_dt[, mean(label == pred)]
```

---
layout: false
class: clear, middle

So what does the UMAP embedding look like?

---
layout: false
class: clear, middle, center

.b[2-dimensional UMAP embedding] of 784-dimensional MNIST pixel space

```{r, umap-plot, echo = FALSE}
ggplot(train_umap, aes(x = UMAP1, y = UMAP2, color = label)) +
  geom_point(alpha = 0.28, size = 0.7) +
  theme_minimal(base_size = 18) +
  labs(
    x = expression(UMAP[1]),
    y = expression(UMAP[2]),
  ) +
  # Set color scale alpha to 1
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 3))) +
  scale_color_viridis_d('', option = 'magma', end = 0.9)
```

---
layout: false
class: clear, middle

Cool. So how does this 2-D UMAP embedding perform?

---
# Predicting digits
## K-means in the UMAP space

With only .attn[2 dimensions].grey-vlight[(!!)], *K*-means accuracy is .b.pink[`r scales::percent(umap_k10_accuracy, accuracy = 0.1)`] (*K*=10).

- This result *does not* mean UMAP always wins.
- It *does* mean a .b.purple[nonlinear embeddings] can preserve useful structures.
- UMAP's .b.purple[nonlinearity] can capture complex relationships.

Reading handwriting likely involves such complex/nonlinear relationships.

---
layout: false
class: clear, middle

**Comparing *K*-means accuracy** across representations and values of $K$

```{r, compare-k-data, include = FALSE, cache = TRUE}
find_k_pca = future_rbind(
  1:30,
  function(k) {
    data.table(
      k = k,
      space = 'PCA',
      accuracy = kmeans_modal_accuracy(
        train_pc |>
          select(starts_with('PC')),
        train_pc$label,
        k
      )
    )
  }
)

find_k_umap = future_rbind(
  1:30,
  function(k) {
    data.table(
      k = k,
      space = 'UMAP',
      accuracy = kmeans_modal_accuracy(
        train_umap |>
          select(starts_with('UMAP')),
        train_umap$label,
        k
      )
    )
  }
)

find_k_pixels_labeled = copy(find_k)[, space := 'Pixels']
acc_dt = rbindlist(
  list(find_k_pixels_labeled, find_k_pca, find_k_umap),
  use.names = TRUE
)
acc_dt[, accuracy := as.numeric(accuracy)]
acc_dt[, space := factor(space, levels = c('Pixels', 'PCA', 'UMAP'))]
```

```{r, compare-k-plot, echo = FALSE}
ggplot(acc_dt, aes(x = k, y = accuracy, color = space, shape = space)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_line(linewidth = 0.45) +
  geom_point(size = 2.1) +
  theme_minimal(base_size = 18, base_family = 'Fira Sans Condensed') +
  scale_color_manual(values = c('Pixels' = red_pink, 'PCA' = orange, 'UMAP' = purple)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    x = 'K',
    y = 'Accuracy',
    color = '',
    shape = '',
  ) +
  theme(legend.position = 'bottom') +
  guides(color = guide_legend(override.aes = list(size = 3)))
```

---
# Predicting digits
## Representation tradeoffs

What do these results mean?

- .hi-pink[Pixels] Simplest description, roughest geometry
- .hi-orange[PCA] Linear, compact, mathematically clean
- .hi-purple[UMAP] Nonlinear, visual, less interpretable

In-sample accuracy at $K = 10$

- .hi-pink[Pixels] `r scales::percent(k10_accuracy, accuracy = 0.1)`
- .hi-orange[PCA] `r scales::percent(pca_k10_accuracy, accuracy = 0.1)`
- .hi-purple[UMAP] `r scales::percent(umap_k10_accuracy, accuracy = 0.1)`

Representation choice can change clustering performance a lot.

---
layout: true

# Wrapping up

---
class: inverse, middle
name: wrap

---
## Structure and prediction

Unsupervised tools help depicts whether the features contain signal.

- they can reduce dimension before supervised learning,
- they can reveal where classes overlap

If the goal is prediction, we usually move back to .attn[supervised] methods.

.grey-vlight[(We could probably beat *K*-means with a more direct supervised method.)]

---
## Basic image-classification intuition

Our approach today for image classification

1. learn a good .attn[representation],
1. use that representation to make a prediction.

.note[Next step] Neural networks start learning both pieces together.

--

.note[Reminder] We have been treating the .b.purple[representation] as a .b[preprocessing step], but it is really part of the model. Cross validation should reflect that.

---
name: final-thoughts
## Final thoughts

What did we see?

- .b.pink[*K*-means] is simple; it depends heavily on geometry. <br>Other clustering methods have different assumptions and objectives.
- Raw, disjoint .b.pink[pixel space] is high-dimensional and awkward.<br>Convolutions and nonlinear embeddings can help.
- .hi-orange[PCA] gives a linear low-dimensional summary.<br>Reliance on linearity can be too restrictive.
- .hi-purple[UMAP] gives a nonlinear, low-dimensional summary.

.note[Next stop] Neural networks and a more serious classification pipeline.

---
layout: false
# Table of contents

.col-left[
.small[
#### Opening
- [Admin](#admin)
- [Why digits?](#intro)
- [Image classification](#vision)
- [Supervised vs. unsupervised](#unsupervised)

#### Clustering
- [K-means](#kmeans)
- [Choosing $K$](#choose-k)
]
]

.col-right[
.small[
#### Dimension reduction
- [PCA](#pca)
- [UMAP](#umap)

#### Closing
- [Structure and prediction](#wrap)
- [Final thoughts](#final-thoughts)

]
]
