---
title: "Problem Set 1"
subtitle: "`tidymodels` example"
format:
  html:
    toc: true
    number-sections: true
    self-contained: true
    theme:
    - cosmo
    - ed.scss
editor:
  render-on-save: true
params:
  showanswers: false
---

## Setup

First, let's load the packages and data...

```{r}
#| label: setup

# Load necessary packages
library(pacman)
p_load(here, readr, skimr, fst, tidymodels, glmnet)
# Load the data
raw_df =
  here('projects', 'project-000', 'sales-data.fst') |> 
  read_fst(as.data.table = TRUE)
  # read_fst(as.data.table = TRUE, from = 1, to = 2e4)
  # read_fst(as.data.table = TRUE) |> sample_frac(0.05)
```

**Note:** When working with a relatively large dataset like this one, it can help to use a subset first. If you don't care about representativeness, you could just read the first few rows of the dataset, e.g., the commented-out `from = 1, to = 1e4` above. If you want a representative sample, you could `sample_n` or `sample_frac` from the full dataset (also commented out above). This approach can save you *a lot* of time when you're testing out your code. For example, the `prep` step of the was 20-times faster when I used a 5% sample.

## Data cleaning and preprocessing

Before we write the *official* `recipe`, which is more about *engineering* features, I'm going to fix any strangeness in the dataset—e.g., the variables that are incorrectly `class`-ed. Examples include variables that should be numeric but were read as character due to `$` or `,` and codes that should be factors but were read as character. 

```{r}
#| label: fix

# Recode variables
clean_df =
  raw_df |>
  # Parse characters to numbers
  mutate(
    sale_price = sale_price |> parse_number(),
    land_square_feet = land_square_feet |> parse_number(),
    building_square_feet = building_square_feet |> parse_number(),
    number_of_units = number_of_units |> parse_number(),
    percent_ownership = percent_ownership |> parse_number(),
    other_improvements = other_improvements |> parse_number(),
  ) |>
  # Convert numeric variables to factors
  mutate(
    across(
      c(
        ends_with('material'),
        ends_with('finish'),
        contains('class'),
        pin, use, type_of_residence, design_plan, multi_code
      ),
    as.character
    )
  ) |>
  # Remove unwanted variables :(
  select(!starts_with('square_root')) |>
  select(!contains('squared')) |>
  select(!starts_with('estimate')) |>
  select(!c(
    deed_no, census_tract,
    town_code, town_and_neighborhood,
    neighborhood_code, neigborhood_code_mapping,
    total_building_square_feet, property_address,
    sale_date, pure_market_filter
  ))
``` 

I'm also going to remove observations with a sales price less than or equal to $1. Ideally, we would have a flag for "non-arm's-length" sales. We don't, so this crude filter is the best we can do to remove non-arm's-length sales.

```{r}
#| label: filter

# Filter out non-arm's-length sales
clean_df = clean_df |> filter(sale_price > 1)

# Check it out
clean_df |> skim()
```

Let's define a recipe for cleaning and preprocessing the data. As discussed in class, we'll *at least* need to convert categorical predictors to dummies and standardize the numeric predictors.

```{r}
#| label: recipe

# Define a recipe for preprocessing the data
a_recipe =
  # Start the recipe
  recipe(sale_price ~ ., data = clean_df) |> 
  # Update ID variable role
  update_role(pin, new_role = "id") |>
  # Drop variables with too many missing values
  step_filter_missing(all_predictors(), threshold = 0.01) |>
  # Impute missing values in numeric predictors with the median
  step_impute_median(all_numeric_predictors()) |>
  # Interact numeric predictors with each other
  step_interact(~ all_numeric_predictors():all_numeric_predictors()) |>
  # Normalize numeric predictors
  step_normalize(all_numeric_predictors()) |>
  # Impute missing values in categorical predictors with the most common value
  step_impute_mode(all_nominal_predictors()) |>
  # Collapse infrequent factor levels into 'other'
  step_other(all_nominal_predictors(), threshold = 0.05) |>
  # Create dummy variables for categorical predictors
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>
  # Remove near-zero-variance predictors
  step_nzv(all_predictors()) |>
  # Remove predictors that are highly correlated with other predictors
  step_corr(all_predictors(), threshold = 0.9) |>
  # Remove variables that are linear combinations of other predictors
  step_lincomb(all_predictors())

# Let's see the recipe
a_recipe
```

Notice that executing this recipe *once* takes quite some time.

```{r}
#| label: prep

# Time the execution of the recipe
tictoc::tic()
baked_df = a_recipe |> prep() |> bake(new_data = NULL)
tictoc::toc()

# Check out the baked data
baked_df |> skim()
```

So if you are then asking R to run this recipe *each* time you fit a model (for each fold of cross-validation, for each value of the tuning parameters), that can add up to *a lot* of time. 

One more note: I used `step_other` to collapse infrequent factor levels to an "other" category. This step will help you avoid creating a huge number of dummy variables that are mostly `0`s. However, it *could* cause you to lose important information.

## Elasticnet

Now we can set up our **workflow**. We will need to specify and model. Then we can add our recipe.

```{r}
#| label: net-setup

# Define the model specification
net_model =
  linear_reg(penalty = tune(), mixture = tune()) |>
  set_engine('glmnet')
# Define the workflow
net_wf =
  workflow() |>
  add_model(net_model) |>
  add_recipe(a_recipe)

# Print the workflow
net_wf
```

We're finally ready to tune the model... after we create a resampling object (define the folds) and define the set of tuning parameters we wish to evaluate.

This may take some time. I'm going to use `tictoc` to time the execution of this code chunk again.

Note that parallelization can help here—to a degree. Tuning 5 levels of `lambda` took 99 seconds to run sequentially on my computer. When I parallelized (with 5 workers in `future`), the same code took approxiately 30 seconds.

```{r}
#| label: tune
#| cache: true

# Start timer
tictoc::tic()

# Define the resampling object
set.seed(12)
cv_folds = vfold_cv(clean_df, v = 5)
# Parallelize the tuning process
library(future)
plan(multisession, workers = parallelly::availableCores() - 1)
# Tune the model
tune_results =
  tune_grid(
    net_wf,
    resamples = cv_folds,
    grid = expand_grid(
      penalty = 10^seq(-5, 3, length.out = 50),
      mixture = seq(0, 1, length.out = 11)
    ),
    metrics = metric_set(rmse, mae, rsq),
    control = control_grid(parallel_over = 'everything')
  )

# Stop timer
tictoc::toc()
```

```{r}
#| label: tune-results

# Check out the tuning results: RMSE
tune_results |> show_best(metric = 'rmse', n = 10)
# Check out the tuning results: MAE
tune_results |> show_best(metric = 'mae', n = 10)
# Plot the tuning results
tune_results |> autoplot()
```