---
name: project setup
---

## Setup

[_0._]{.hp} [Create a new directory for EC524, somewhere (e.g., Desktop, Documents, iCloud, Dropbox, etc.)]{.h}

::: {.faded}
[_1._]{.hp} Within this directory, create subdirectories

[_2._]{.hp} Open RStudio

[_3._]{.hp} Click on **File** \> **New Project...** \> **Existing Directory** and navigate to the separate project folder under "lab" \> "001-projects" and click **Create Project**. RSudio will open a new session with the working directory set to the project folder

[_4._]{.hp} Move the data files and Quarto document into the project folder

[_5._]{.hp} Open the `doc001.qmd` file in the project folder to get started
:::

::: {.notes}
Before we can start any analysis, we need to load the data in RStudio. 
There are many ways to do this, but just a little forward thinking on project work flow will save us headaches later.

As discussed last week, we will be using RStudio projects to manage our work. 
This is a great way to keep organized between separate projects (assignments) and simplify the process of loading data and saving files. 
Here are the steps to create a new project:
:::

---

## Setup

::: {.faded .small}
[_0._]{.hp} Create a new directory for EC524, somewhere (e.g., Desktop, Documents, iCloud, Dropbox, etc.)
:::

[_1._]{.hp} [Within this directory, create subdirectories]{.h}

```
      EC524W25              # Class folder
      └── lab               # Lab folder
          └── 02-projects   # Separate folder for the lab 02 project
```

::: {.faded .small}
[_2._]{.hp} Open RStudio

[_3._]{.hp} Click on **File** \> **New Project...** \> **Existing Directory** and navigate to the separate project folder under "lab" \> "001-projects" and click **Create Project**. RSudio will open a new session with the working directory set to the project folder

[_4._]{.hp} Move the data files and Quarto document into the project folder

[_5._]{.hp} Open the `w02-lab.qmd` file in the project folder to get started
:::

---

## Setup

::: {.faded}
[_0._]{.hp} Create a new directory for EC524, somewhere (e.g., Desktop, Documents, iCloud, Dropbox, etc.)

[_1._]{.hp} Within this directory, create subdirectories
:::

[_2._]{.hp} Open RStudio

[_3._]{.hp} Click on **File** \> **New Project...** \> **Existing Directory** and navigate to the separate project folder under "lab" \> "001-projects" and click **Create Project**. 
RSudio will open a new session with the working directory set to the project folder

[_4._]{.hp} Move the data files and Quarto document into the project folder

[_5._]{.hp} Open the `w02-lab.qmd` file in the project folder to get started

---

## Setup

After completing these steps, here is an visualization of a well organized directory structure:

```

   EC524W25                                # Class folder  
   └──lab                                  # Lab folder
      ├── 002-projects                     # Project folder
      │   ├── 002-projects.Rproj           # Rproj file (we are here)
      │   ├── data                         # Data folder
      │   │   ├── data_description.txt     
      │   │   ├── sample_submission.csv    
      │   │   ├── test.csv                 
      │   │   └── train.csv                
      │   └── w02-lab.qmd                  # Quarto document
      └── 003-...                          # Future lab projects
      
```

Download the zip file under lab 02 on the course GitHub repo. 

Unzip the file in your "lab" folder. 

---

## Getting Started

First, install `pacman`, a package manager that will help us install and load packages. Recall we only need to install packages once. Ex.

``` r
install.packages("pacman")
```

However, we always have to load them in each new R session. We so by running the following code:

```{r setup-pacman, echo=TRUE}
library(pacman)
```

Sometimes we have to load a bunch of packages at once, which can be wordy. Ex.

``` r
# Load ALL packages
library(tidyverse)
library(tidymodels)
library(broom)
library(magrittr)
library(janitor)
library(here)
```

---

## Getting Started

The `pacman` package can simplify our code with the `p_load` function. 
This function allows us to load mulitple pacakges at once. 
Further, if you do not have it installed, it will install it for you. 

```{r setup-pload, echo=TRUE}
# Load ALL packages
pacman::p_load(tidyverse, data.table, broom, magrittr, janitor, skimr)
```

You can use either method to load packages. 
The `p_load` function is a bit more concise and I use it often.

::: {.fragment}
::: callout-tip 
New packages can be scary. To learn more, read the vignette.

```{r vignette-tip, eval=FALSE}
browseVignettes(package = 'here')
```

Vignettes are a short introduction to the package and its functions. 
Great for a quick overview and reference. 
But not all packages have them.
:::
:::