---
name: resampling methods
---

## Randomization seeds

Since we are using some randomization, let's set a seed so we all get the same results. A random seed is a number used to initialize a pseudorandom number generator. Ex.

```{r set-seed}
# Set seed
set.seed(123)
```

<br> 

These functions are used to ensure that the results of our code are reproducible. 
This is important for debugging and sharing code and it is a good practice to include them at the beginning of your script whenever you are using randomization.

---

## Data 

Load both the `train.csv` and `test.csv` data sets. We will be using both.

```{r read-data, message=FALSE}
# Clean Environment
rm(list = ls())

# Load training data
house_df = read_csv("data/train.csv")
```

Double check that everything is loaded correctly. Ex.

```{r check-data, eval=FALSE}
# Print first 10 observations
head(house_df, 10)
```

These data have really crappy column names. Let's clean them up using the `clean_names` function from the `janitor` package.

```{r}
# Clean column names
house_df = house_df %>% clean_names()
```

---

## Data {.small}

For today, we only need a subset. Let's trim down our data set to four columns:

-   `id`: house id
-   `sale_price`: sale price of the house
-   `age`: age of the house at the time of the sale (difference between year sold and year built)
-   `area`: the non-basement square-foot area of the house

We can do this in a few different ways but `dplyr::transmute()` is a very convenient function to use here

```{r subset-data}
# Subset our data and create new columns
house_df %<>% transmute(
    id = id,
    sale_price = sale_price / 10000,
    age = yr_sold - year_built,
    area = gr_liv_area
)
```

---

## Data {.small}

It's always a good idea to look at the new dataframe and make sure it is exactly what we expect it to be. Any of the following is a great way to check:

- `summary()`: provides a quick overview of the data
- `glimpse()`: provides a more detailed overview of each column
- `skimr::skim()`: provides a "nicer" looking overview of the data
- `View()`: opens spreadsheet view of the data

```{r subset-data-check}
summary(house_df)

glimpse(house_df)
```

---

## Resampling

::: {.vertical-center}
- Today we are introducing resampling via [holdout sets]{.h} 
    - A very straightforward method that is used to estimate the performance

- The key concept here is that we are partitioning our data into two sets: 
    - a training set 
    - a validation set

- The training set is used to fit the model, while the validation set is used to evaluate the model's performance
:::

---

## Create a Validation Set

Start by creating a single validation set composed of 30% of our training data. 

- We can draw the validation sample randomly using the `dplyr` function `sample_frac()`. 

- The argument `size` will allow us to choose the desired sample of 30%.

- `dplyr`'s function `setdiff()` will give us the remaining, non-validation observation from the original training data.

```{r split}
# Draw validation set
validation_df = house_df %>% sample_frac(size = 0.3)
# Find remaining training set
train_df = setdiff(house_df, validation_df)
```

---

## Verify We Did It Right

Finally, let's check our work and make sure that

[`training_df` + `validation_df` = `house_df`]{.align-center}

- To do so, we can use `nrow()`

<br>

::: {.solution .fragment}
```{r check-split}
# Check that dimensions make sense
nrow(house_df) == nrow(validation_df) + nrow(train_df)
```
:::

---

## Model fit

With training and validation sets we can start train a learner on the training set and evaluate its performance on the validation set. We will use a linear regression model to predict `sale_price` using `age` and `area`. We will give some flexibility to the model by including polynomial terms for `age` and `area`.

We proceed with the following steps (algorithm):

[_1._]{.hp} Train a regression model with various degrees of flexibility

[_2._]{.hp}  Calculate MSE on the `training_df`

[_3._]{.hp}  Determine which degree of flexibility minimizes validation MSE

---

## Model specification

We will fit a model of the form:

\begin{align*}
Price_i = &\beta_0 + \beta_1 * age_i^2 + \beta_2 * age_i + \beta_3 * area_i^2 +  \\ 
          &\beta_4 * area_i + \beta_5 * age_i^2 \cdot area_i^2 + \beta_6 * age_i^2 \cdot area_i + \\
          & \beta_7 * age_i \cdot area_i^2 + \beta_8 * area_i \cdot age_i
\end{align*}

Often in programming, we want to automate the process of fitting a model to different specifications. 
We can do this by defining a function.

---

## Model Fit: Creating a Function

We define a function that will fit a model to the training data and return the validation MSE. 
The function takes two arguments: 

:::: {.columns}

::: {.column .align-center}
`deg_age` 
:::
::: {.column .align-center}
`deg_area`
:::
::::

These arguments represent the degree of polynomial for [age]{.h} and [area]{.h} that we want to fit our model. 

```{r}
# Define function
fit_model = function(deg_age, deg_area) {
    # Estimate the model with training data
    est_model = lm(
      sale_price ~ poly(age, deg_age, raw = T) * poly(area, deg_area, raw = T),
      data = train_df)
    # Make predictions on the validation data
    y_hat = predict(est_model, newdata = validation_df, se.fit = F)
    # Calculate our validation MSE
    mse = mean((validation_df$sale_price - y_hat)^2)
    # Return the output of the function
    return(mse)
}
```

---

## Model Fit

First let's create a dataframe that is 2 by 4x6 using the `expand_grid()` function. We will attach each model fit MSE to an additional column.

```{r}
# Take all possible combinations of our degrees
deg_df = expand_grid(deg_age = 1:6, deg_area = 1:4)
deg_df
```

---

## Function

Now let's iterate over all possible combinations (4x6) of polynomial specifications and see which model fit produces the smallest MSE.

```{r}
# Iterate over set of possibilities (returns a vector of validation-set MSEs)
mse_v = mapply(
    FUN = fit_model,
    deg_age = deg_df$deg_age,
    deg_area = deg_df$deg_area
)
```

---

## Funtcion

Now that we have a 1 by 24 length vector of all possible polynomial combinations, lets attach this vector as an additional column to the `deg_df` dataframe we assigned a moment ago and arrange by the smalled MSE parameter.

```{r mse-result}
# Add validation-set MSEs to 'deg_df'
deg_df$mse = mse_v
# Which set of parameters minimizes validation-set MSE?
arrange(deg_df, mse)
```

---

## Plot the MSE (Code)

Now let's turn this table into a more visual appealing plot. I used `ggplot2` to make the following heat map. Recall, less MSE is better.

```{r mse-plot-code, eval=FALSE, fig.width=8, fig.height=6}
#| fig-align: center
# Plot
ggplot(data = deg_df, aes(x = deg_age, y = deg_area, fill = mse)) +
  geom_tile() +
  #hrbrthemes::theme_ipsum(base_size = 12) +
  scale_fill_viridis_c("MSE", option = "inferno", begin = 0.1, direction = -1) +
  scale_x_continuous(breaks = seq(1, 6, 1)) +
  scale_y_continuous(breaks = seq(1, 6, 1)) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(
    title = 'Model fit MSE',
    subtitle = 'Fill values denote MSE',
    x = "Degrees of age",
    y = 'Degrees of area',
    colour = 'Log of MSE')
```

---

## Plot the MSE (Graph)

```{r mse-plot, echo=FALSE, fig.width=8, fig.height=6}
#| fig-align: center
# Plot
ggplot(data = deg_df, aes(x = deg_age, y = deg_area, fill = mse)) +
  geom_tile() +
  #hrbrthemes::theme_ipsum(base_size = 12) +
  scale_fill_viridis_c("MSE", option = "inferno", begin = 0.1, direction = -1) +
  scale_x_continuous(breaks = seq(1, 6, 1)) +
  scale_y_continuous(breaks = seq(1, 6, 1)) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(
    title = 'Model fit MSE',
    subtitle = 'Fill values denote MSE',
    x = "Degrees of age",
    y = 'Degrees of area',
    colour = 'Log of MSE')
```
