---
name: coding practice
---

##  Loading the Data

Now that we have our project set up, let's load the data. 
First, let's see where we are on our computers

```{r eval=FALSE}
# The following function prints the current working directory
getwd()
```

You should find the output of the `getwd()` function to be the path to the project folder. 
This is our **working directory**, or `.` for short. 
This is where R will look for files when we use relative paths.

<br>

::: {.fragment}
Let's load the data in the code chunk below using the `read_csv` function. 
For quick reference for this function type `?read_csv` in the console.

```{r}
# Load train.csv 
train_df = read_csv("data/train.csv")
```
:::

---

## Loading the Data

If the read worked, the following code should print out the first 6 rows in the console:

```{r eval=FALSE}
# Print the first 6 rows of the data
head(train_df)
```

If this worked, we are ready to move on. 
If not, double check your work or wait for me to come around and help.

---

## `dplyr` questions

If you have managed to load the "train.csv" data as the `train_df` object, the following questions will test your knowledge of the `dplyr` package. There are three questions (easy, medium, and hard), each with a code chunk to fill in your answer.

<br>

Before starting the questions, I would clean things up first:

```{r}
# Load convenience packages
library(magrittr)
library(janitor)

# Clean names
train_df %<>% clean_names()

# Rename first and second floor columns
train_df %<>%
  rename(first_floor_sqft = x1st_flr_sf,
         second_floor_sqft = x2nd_flr_sf)

```

---

## Question 01 *(easy)* {.small}


**Task**: Filter the dataset to only include rows with the two story houses that were built since 2000. Return only the following columns in the output: `Id`, `YearBuilt`, `HouseStyle`. Order the result by year built in descending order.

[Expected Result]{.hi}

```{r eval=FALSE}
# A tibble: 143 × 3
|   Id | year_built | house_style |
|------|------------|-------------|
|   88 |       2009 | 2Story      |
|  158 |       2009 | 2Story      |
|  213 |       2009 | 2Story      |
|  461 |       2009 | 2Story      |
|  573 |       2009 | 2Story      |
|  763 |       2009 | 2Story      |
```

::: {.solution .fragment}
```{r eval=FALSE}
# Your answer here
train_df %>%
  filter(house_style == "2Story" & year_built > 2000) %>%
  select(id, year_built, house_style) %>%
  arrange(desc(year_built))
```
:::

---

## Question 02 *(medium)* {.small}

**Task**: Create a new column called `TotalSF` that is the sum of the `1stFlrSF` and `2ndFlrSF` columns. Filter the dataset to only include rows with a total square footage greater than 3,000. Return only the `Id` and `TotalSF` columns in the output.

[Expected output]{.hi}

```{r eval=FALSE}
# A tibble: 12 × 2
|   id | total_sqft |
|------|------------|
|  119 |       3222 |
|  186 |       3036 |
|  305 |       3493 |
|  497 |       3228 |
|  524 |       4676 |
```

::: {.solution .fragment} 

```{r eval=FALSE}
# Your answer here
train_df %>%
  mutate(total_sqft = first_floor_sqft + second_floor_sqft) %>%
  filter(total_sqft > 3000) %>%
  select(id, total_sqft)
```

:::

---

## Question 03 *(hard)* {.small}

**Task**: From the dataset, find the average `YearBuilt` for each `HouseStyle`, but only include `HouseStyle`s where more than 20 houses were built after the year 2000. Sort the resulting data frame by `YearBuilt` in descending order.

[Expected output]{.hi}

```{r eval=FALSE}
# A tibble: 3 × 3
| house_style | average_year_built |   n   |
|-------------|--------------------|-------|
| 1Story      | 2005.0             |  211  |
| 2Story      | 2005.0             |  143  |
| SLvl        | 2004.0             |    6  |
```

::: {.solution .fragment}
```{r eval=FALSE}
# Your answer here
train_df %>%
  filter(year_built > 2000)%>%
  group_by(house_style) %>%
  summarise(average_year_built = mean(year_built), 
            n = n()) %>%
  filter(n > 5) %>%
  arrange(desc(average_year_built))
```
:::


