
---
## Common activation functions

```{r, activation-functions2, echo = FALSE}
activation_df |>
  filter(activation %in% c('Identity', 'Sigmoid', 'Tanh', 'ReLU')) |>
  ggplot(aes(x = x, y = y, color = activation)) +
  geom_hline(yintercept = 0, linewidth = 0.25) +
  geom_vline(xintercept = 0, linewidth = 0.25) +
  geom_line(linewidth = 1.1, show.legend = FALSE) +
  facet_wrap(~activation, scales = 'free_y') +
  scale_color_manual(values = c(
    Identity = slate,
    Sigmoid = red_pink,
    Tanh = orange,
    ReLU = blue
  )) +
  theme_minimal(base_size = 17) +
  labs(x = 'z', y = 'g(z)')
```

---
## The derivatives matter too

```{r, activation-derivatives, echo = FALSE}
activation_deriv_df |>
  ggplot(aes(x = x, y = y, color = derivative)) +
  geom_hline(yintercept = 0, linewidth = 0.25) +
  geom_line(linewidth = 1.0, show.legend = FALSE) +
  facet_wrap(~derivative, scales = 'free_y') +
  scale_color_manual(values = c(
    'sigmoid\'(z)' = red_pink,
    'tanh\'(z)' = orange,
    'ReLU\'(z)' = blue
  )) +
  theme_minimal(base_size = 17) +
  labs(x = 'z', y = 'g\'(z)')
```

---
## Why those derivatives matter

The key derivatives are:

$$
\sigma'(z) = \sigma(z)\big(1 - \sigma(z)\big)
$$

$$
\tanh'(z) = 1 - \tanh^2(z)
$$

$$
\text{ReLU}'(z) = \mathbf{1}\{z > 0\}
$$

Why do we care?

- gradient-based training keeps multiplying by these derivatives
- if the derivative is tiny, updates can shrink
- if the derivative is zero, learning can stall locally

---
## Output activations should match the task

.col-left[
.hi-pink[Regression]

- often use identity output
- prediction is unrestricted
]

.hi-slate[Binary classification]

- often use sigmoid output
- prediction is a probability
]

.hi-orange[Multiclass classification]

- often use softmax output
- prediction is a vector of class probabilities
]

---
## Forward propagation

At layer $\ell$:

$$
z^{(\ell)} = W^{(\ell)}a^{(\ell-1)} + b^{(\ell)}
\qquad\text{and}\qquad
a^{(\ell)} = g^{(\ell)}\big(z^{(\ell)}\big)
$$

So the network repeats the same recipe:

1. weighted sums
1. activations
1. pass the result forward

---
## A dimension check helps

If:

- $a^{(\ell-1)}$ is $m_{\ell-1} \times 1$
- $W^{(\ell)}$ is $m_\ell \times m_{\ell-1}$
- $b^{(\ell)}$ is $m_\ell \times 1$

then:

$$
z^{(\ell)} = W^{(\ell)}a^{(\ell-1)} + b^{(\ell)}
$$

is also $m_\ell \times 1$.

That bookkeeping matters once the network gets deep.

---
## Softmax for multiclass output

If there are $K$ classes, the output layer produces

$$
\hat p_k = \frac{\exp(z_k)}{\sum_{j=1}^K \exp(z_j)}
$$

for $k = 1, \ldots, K$.

Properties:

- each $\hat p_k$ is between 0 and 1
- the probabilities sum to 1
- the largest one is the predicted class

---
## Gradient descent

Once we have $C(\theta)$, we update

$$
\theta \leftarrow \theta - \eta \nabla_\theta C(\theta)
$$

where:

- $\theta$ collects all weights and biases
- $\eta$ is the learning rate

Same idea as before in the course, but now the parameter vector is much larger.

---
## Mini-batches and epochs

In practice we usually update on small batches:

- split the data into .attn[mini-batches]
- compute the gradient on one mini-batch
- update the weights
- repeat

One full pass through the training data is an .attn[epoch].

---
## Where do the gradients come from?

A weight in an early layer affects the loss through a chain:

- change a weight
- that changes a weighted sum
- that changes an activation
- that changes later activations and eventually the loss

So computing one gradient means following a .attn[chain of effects] through the network.

---
## One hidden weight affects loss indirectly

For one hidden-layer weight $w$,

$$
\frac{\partial C}{\partial w}
=
\frac{\partial C}{\partial \hat y}
\cdot
\frac{\partial \hat y}{\partial a^{(1)}}
\cdot
\frac{\partial a^{(1)}}{\partial z^{(1)}}
\cdot
\frac{\partial z^{(1)}}{\partial w}
$$

Deeper networks have longer chains, but the idea is the same.

---
## Backprop in words

- run the .attn[forward pass] and compute the loss
- start at the output and measure how wrong it is
- pass that error signal backward, one layer at a time
- combine those signals with local derivatives
- use the result to update weights and biases

Backprop is just an efficient way to organize that repeated chain rule.

---
## Backpropagation reuses the chain rule

```{r, backprop-plot, echo = FALSE}
make_backprop_plot()
```

---
## Start at the output layer

Write the output-layer error signal as:

$$
\delta^{(L)} =
\frac{\partial C}{\partial a^{(L)}} \odot g'\big(z^{(L)}\big)
$$

For .attn[softmax + cross-entropy], this simplifies nicely:

$$
\delta^{(L)} = \hat p - y
$$

That simplification is one reason cross-entropy is so common in classification.

---
## Then move backward through the hidden layers

For earlier layers,

$$
\delta^{(\ell)} =
\left(W^{(\ell+1)}\right)' \delta^{(\ell+1)}
\odot g'\big(z^{(\ell)}\big)
$$

So each layer receives an error signal from the layer after it, then adjusts it using its own derivative.

That is backpropagation in one sentence.

---
## Then turn those signals into gradients

Once we have the layer errors, the gradients are local:

$$
\frac{\partial C}{\partial W^{(\ell)}} =
\delta^{(\ell)} \left(a^{(\ell-1)}\right)'
\qquad\text{and}\qquad
\frac{\partial C}{\partial b^{(\ell)}} = \delta^{(\ell)}
$$

Then update:

$$
W^{(\ell)} \leftarrow W^{(\ell)} - \eta \frac{\partial C}{\partial W^{(\ell)}}
$$

and similarly for the biases.

---
## What activation are we using here?

For the MLP slide example we now use:

- `mlp(..., activation = 'relu') |> set_engine('brulee')`

That means:

- the .attn[`activation`] argument applies to the hidden layer, not the output layer
- for a multiclass factor response, the .attn[output layer] is still chosen by the task
- here the hidden units use .attn[ReLU], which is closer to a modern neural-net default than the old `nnet` setup

So the MLP example is now a .b[ReLU hidden-layer network], while the multinomial logit baseline remains the no-hidden-layer comparison.

---
## There's a lot more to training

Once the basic setup is in place, training quality also depends on:

- cross-entropy versus squared error
- regularization
- weight initialization
- architectural choices

So the story is not just .it[add a hidden layer].

The details of optimization matter.

---
## What a convolution does

- choose a small filter, often `3 x 3`
- place it over one local patch of pixels
- multiply and add to get one score
- slide the .attn[same filter] across the whole image

So one filter asks the same question everywhere:

- is there a vertical stroke here?
- is there an edge or a corner here?

In practice, the filter weights are .attn[learned from data].

---
## One tiny convolution example

Take a local patch

$$
X =
\begin{bmatrix}
0 & 1 & 1 \\
0 & 1 & 1 \\
0 & 1 & 1
\end{bmatrix}
\qquad\text{and filter}\qquad
K =
\begin{bmatrix}
-1 & 0 & 1 \\
-1 & 0 & 1 \\
-1 & 0 & 1
\end{bmatrix}
$$

Then the score at that location is

$$
\sum_{u,v} X_{uv} K_{uv} = 3
$$

So this filter responds when the patch is dark on the left and bright on the right:
a simple .attn[vertical-edge detector].

---
## A filter sliding over an image

The highlighted `3 x 3` patch on the left generates the highlighted output cell on the right.

```{r, convolution-demo, echo = FALSE}
make_convolution_demo_plot()
```

---
## A more ambitious workflow

To make the example more realistic, we will:

1. build a train / validation / test split
1. compare multinomial logit to a starter MLP
1. tune the MLP's capacity and preprocessing
1. see whether a very simple CNN does better on raw images
1. refit the better models on train + validation
1. evaluate on the held-out test set

This is still small, but it is now a real modeling workflow.

---
## The starter and tuned models

```{r, model-code, eval = FALSE}
# Baseline recipe: normalize and keep 40 PCs.
rec_base =
  recipe(label ~ ., data = mnist_train) |>
  step_zv(all_predictors()) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 40)

# Tuned recipe: keep 60 PCs and add 8 UMAP coordinates.
rec_tuned =
  recipe(label ~ ., data = mnist_train) |>
  step_zv(all_predictors()) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 60) |>
  embed::step_umap(
    all_predictors(),
    num_comp = 8,
    neighbors = 30,
    min_dist = 0.05,
    keep_original_cols = TRUE,
    prefix = 'umap'
  )

# Starter MLP: small network with ReLU hidden units.
mlp_start_spec =
  mlp(hidden_units = 15, penalty = 0.01, epochs = 50, activation = 'relu') |>
  set_engine('brulee') |>
  set_mode('classification')

# Tuned MLP: more hidden units, lighter penalty, more epochs.
mlp_tuned_spec =
  mlp(hidden_units = 50, penalty = 0.001, epochs = 150, activation = 'relu') |>
  set_engine('brulee') |>
  set_mode('classification')

# Simple CNN: two small conv layers on the raw 28 x 28 image.
# conv -> relu -> pool -> conv -> relu -> pool -> dense -> 10 logits
```

---
## A very simple CNN

- keep the raw 28 x 28 image instead of PCA or UMAP features
- use small .attn[3 x 3 filters] to scan local patches
- reuse the same filter weights across the image
- then flatten and map to 10 output classes

That is enough to exploit .attn[spatial structure] that the MLP does not see directly.

---
## Validation comparison

```{r, validation-compare-plot, echo = FALSE}
ggplot(validation_compare, aes(x = model, y = accuracy, fill = model)) +
  geom_col(width = 0.62, show.legend = FALSE) +
  geom_text(
    aes(label = scales::percent(accuracy, accuracy = 0.1)),
    vjust = -0.35,
    family = 'Zilla Slab'
  ) +
  scale_fill_manual(values = c(
    'Multinomial logit' = slate,
    'Starter MLP' = grey_mid,
    'Tuned MLP' = red_pink,
    'Simple CNN' = turquoise
  )) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = NULL, y = 'Validation accuracy', title = 'A simple CNN improves again') +
  theme_minimal(base_size = 18)
```

---
## Test comparison

```{r, test-compare-plot, echo = FALSE}
ggplot(test_compare, aes(x = model, y = accuracy, fill = model)) +
  geom_col(width = 0.62, show.legend = FALSE) +
  geom_text(
    aes(label = scales::percent(accuracy, accuracy = 0.1)),
    vjust = -0.35,
    family = 'Zilla Slab'
  ) +
  scale_fill_manual(values = c(
    'Multinomial logit' = slate,
    'Tuned MLP' = red_pink,
    'Simple CNN' = turquoise
  )) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = NULL, y = 'Test accuracy', title = 'A simple CNN does better still') +
  theme_minimal(base_size = 18)
```

---
## A single image, many probabilities

```{r, multi-output-example, echo = FALSE}
plot_digit(example_case, title = paste('True digit =', example_case$label[[1]])) +
  ggplot(example_prob_plot_df, aes(x = digit, y = probability, fill = digit == example_case$label[[1]])) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c('TRUE' = red_pink, 'FALSE' = slate)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    x = 'Output node',
    y = 'Predicted probability',
    title = 'Softmax output for one CNN prediction'
  ) +
  theme_minimal(base_size = 18)
```

---
## Confusion matrix for the simple CNN

```{r, cnn-cm, echo = FALSE}
ggplot(cm_cnn_test, aes(x = .pred_class, y = label, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c('') +
  coord_equal() +
  theme_minimal(base_size = 18) +
  labs(
    x = 'Predicted digit',
    y = 'True digit',
    title = 'Held-out test confusion matrix'
  )
```

---
## What this example does and does not show

It does show:

- real multiclass output
- real train / validation / test logic
- real improvement from a stronger architecture

It does not show:

- deep networks
- state-of-the-art MNIST performance

This is still a .attn[teaching example], just a better one.

