---
title: 'Lecture .mono[011]'
subtitle: 'Neural networks: intuition, math, and MNIST'
author: 'Edward Rubin'
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
exclude: true

```{r, setup, include = FALSE}
library(pacman)
p_load(
  tidyverse,
  data.table,
  tidymodels,
  patchwork,
  scales,
  torch,
  coro
)

# Lecture color palette.
red_pink   = '#e64173'
turquoise  = '#20B2AA'
orange     = '#FFA500'
red        = '#fb6107'
blue       = '#181485'
navy       = '#150E37FF'
green      = '#8bb174'
yellow     = '#D8BD44'
grey_light = 'grey70'
grey_mid   = 'grey50'
grey_dark  = 'grey20'
purple     = '#6A5ACD'
slate      = '#314f4f'

knitr::opts_chunk$set(
  comment = '#>',
  fig.align = 'center',
  fig.height = 6.75,
  fig.width = 10.5,
  warning = FALSE,
  message = FALSE
)
knitr::opts_chunk$set(dev = 'svg')
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(knitr.table.format = 'html')
xaringanExtra::use_xaringan_extra(c('tile_view', 'fit_screen'))

set.seed(111)

# Load MNIST data from the local lecture folder.
train_dt = fread('train.csv') |> as_tibble()
test_dt = fread('test.csv') |> as_tibble()
pixel_cols = names(train_dt)[grepl('^pixel', names(train_dt))]

# Plot one 28 x 28 digit image from a single-row data frame.
plot_digit = function(row_df, title = NULL) {
  ggplot(
    expand_grid(y = 28:1, x = 1:28) |>
      mutate(value = as.numeric(row_df[, pixel_cols])),
    aes(x = x, y = y, fill = value)
  ) +
    geom_raster() +
    scale_fill_viridis_c('') +
    coord_equal() +
    theme_void() +
    theme(legend.position = 'bottom') +
    labs(title = title)
}

# Activation functions for intuition and comparison plots.
activation_df = tibble(x = seq(-5, 5, length.out = 500)) |>
  mutate(
    Identity = x,
    Sigmoid = 1 / (1 + exp(-x)),
    Tanh = tanh(x),
    ReLU = pmax(0, x),
    Step = if_else(x >= 0, 1, 0)
  ) |>
  pivot_longer(-x, names_to = 'activation', values_to = 'y')

# Common derivatives that appear inside backpropagation.
activation_deriv_df = tibble(x = seq(-5, 5, length.out = 500)) |>
  mutate(
    `sigmoid''(z)` = (1 / (1 + exp(-x))) * (1 - 1 / (1 + exp(-x))),
    `tanh''(z)` = 1 - tanh(x)^2,
    `ReLU''(z)` = if_else(x > 0, 1, 0)
  ) |>
  pivot_longer(-x, names_to = 'derivative', values_to = 'y')

# Draw a simple fully connected network for lecture visuals.
make_single_neuron_plot = function() {
  nodes = tribble(
    ~name,    ~x,  ~y,  ~label, ~fill_color, ~text_color, ~size,
    'bias',   1.2, 5.0, '1',    'white',     slate,       5.0,
    'x1',     1.2, 4.0, 'x1',   'white',     slate,       5.0,
    'x2',     1.2, 3.0, 'x2',   'white',     slate,       5.0,
    'xdots',  1.2, 2.0, '...',  'white',     slate,       5.0,
    'xp',     1.2, 1.0, 'xp',   'white',     slate,       5.0,
    'neuron', 4.2, 3.0, 'a',    red_pink,    'white',     5.3,
    'output', 7.2, 3.0, 'yhat', orange,      'white',     4.7
  )

  edges = tribble(
    ~from,   ~to,      ~weight,
    'bias',  'neuron', 'b',
    'x1',    'neuron', 'w1',
    'x2',    'neuron', 'w2',
    'xdots', 'neuron', '',
    'xp',    'neuron', 'wp',
    'neuron','output', ''
  ) |>
    left_join(nodes |> select(name, x_from = x, y_from = y), by = c('from' = 'name')) |>
    left_join(nodes |> select(name, x_to = x, y_to = y), by = c('to' = 'name')) |>
    mutate(
      label_x = 0.58 * x_from + 0.42 * x_to,
      label_y = 0.58 * y_from + 0.42 * y_to
    )

  ggplot() +
    geom_segment(
      data = edges,
      aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
      color = grey_light,
      linewidth = 0.55,
      arrow = arrow(length = unit(0.16, 'inches'))
    ) +
    geom_point(
      data = nodes,
      aes(x = x, y = y, fill = fill_color),
      shape = 21,
      size = 11,
      color = slate,
      stroke = 1.1
    ) +
    geom_text(
      data = nodes,
      aes(x = x, y = y, label = label, color = text_color, size = size),
      family = 'Zilla Slab',
      fontface = 'bold',
      show.legend = FALSE
    ) +
    geom_label(
      data = edges |> filter(weight != ''),
      aes(x = label_x, y = label_y, label = weight),
      family = 'Zilla Slab',
      fill = 'white',
      color = slate,
      label.size = 0,
      size = 4.1
    ) +
    annotate('text', x = 1.2, y = 5.75, label = 'Inputs', family = 'Zilla Slab', color = slate, size = 5.2) +
    annotate('text', x = 4.2, y = 5.75, label = 'Neuron', family = 'Zilla Slab', color = red_pink, size = 5.2) +
    annotate('text', x = 7.2, y = 5.75, label = 'Output', family = 'Zilla Slab', color = orange, size = 5.2) +
    annotate('text', x = 4.2, y = 4.15, label = "z = b + w'x", family = 'Zilla Slab', color = slate, size = 5.1) +
    annotate('text', x = 4.2, y = 1.9, label = 'a = g(z)', family = 'Zilla Slab', color = blue, size = 5.1) +
    annotate('text', x = 8.25, y = 3.0, label = 'prediction', family = 'Zilla Slab', color = slate, hjust = 0, size = 5.0) +
    scale_fill_identity() +
    scale_color_identity() +
    scale_size_identity() +
    coord_cartesian(xlim = c(0.4, 9.2), ylim = c(0.4, 6.1), clip = 'off') +
    theme_void()
}

make_network_plot = function() {
  nodes = tribble(
    ~layer,   ~name, ~x,  ~y,  ~label,     ~type,
    'input',  'x1',  1.0, 4.5, 'pixel 1',  'input',
    'input',  'x2',  1.0, 3.5, 'pixel 2',  'input',
    'input',  'x3',  1.0, 2.5, 'pixel 3',  'input',
    'input',  'x4',  1.0, 1.5, 'pixel 784','input',
    'hidden', 'h1',  3.2, 5.0, '',         'hidden',
    'hidden', 'h2',  3.2, 4.0, '',         'hidden',
    'hidden', 'h3',  3.2, 3.0, '',         'hidden',
    'hidden', 'h4',  3.2, 2.0, '',         'hidden',
    'hidden', 'h5',  3.2, 1.0, '',         'hidden',
    'output', 'o1',  5.5, 4.8, '0',        'output',
    'output', 'o2',  5.5, 3.9, '1',        'output',
    'output', 'o3',  5.5, 3.0, '2',        'output',
    'output', 'o4',  5.5, 2.1, '...',      'output',
    'output', 'o5',  5.5, 1.2, '9',        'output'
  )

  edges = expand_grid(
    from = nodes |> filter(layer == 'input') |> pull(name),
    to = nodes |> filter(layer == 'hidden') |> pull(name)
  ) |>
    bind_rows(
      expand_grid(
        from = nodes |> filter(layer == 'hidden') |> pull(name),
        to = nodes |> filter(layer == 'output') |> pull(name)
      )
    ) |>
    left_join(nodes |> select(name, x_from = x, y_from = y), by = c('from' = 'name')) |>
    left_join(nodes |> select(name, x_to = x, y_to = y), by = c('to' = 'name'))

  ggplot() +
    geom_segment(
      data = edges,
      aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
      color = grey_light,
      linewidth = 0.35,
      alpha = 0.75
    ) +
    geom_point(
      data = nodes,
      aes(x = x, y = y, fill = type),
      shape = 21,
      size = 7,
      color = slate,
      stroke = 1
    ) +
    geom_text(
      data = nodes |> filter(layer == 'input'),
      aes(x = x - 0.45, y = y, label = label),
      hjust = 1,
      family = 'Zilla Slab'
    ) +
    geom_text(
      data = nodes |> filter(layer == 'output'),
      aes(x = x + 0.35, y = y, label = label),
      hjust = 0,
      family = 'Zilla Slab'
    ) +
    annotate('text', x = 1.0, y = 5.45, label = 'Input layer', family = 'Zilla Slab', color = slate) +
    annotate('text', x = 3.2, y = 5.85, label = 'Hidden layer', family = 'Zilla Slab', color = red_pink) +
    annotate('text', x = 5.5, y = 5.75, label = 'Output layer', family = 'Zilla Slab', color = orange) +
    scale_fill_manual(values = c(input = 'white', hidden = red_pink, output = orange)) +
    coord_cartesian(xlim = c(0, 6.5), ylim = c(0.5, 6.1), clip = 'off') +
    theme_void() +
    theme(legend.position = 'none')
}

# Draw a forward-pass / backward-pass computational graph.
make_backprop_plot = function() {
  boxes = tribble(
    ~x,  ~y,  ~label,              ~fill,
    1.0, 2.5, 'Inputs\nx',         'white',
    3.0, 2.5, 'Weighted sum\nz',   '#fde5ec',
    5.0, 2.5, 'Activation\na',     '#eef4ff',
    7.1, 2.5, 'Loss\nC',           '#fff2db'
  )

  arrows = tribble(
    ~x,   ~y,   ~xend, ~yend, ~label,
    1.65, 2.5,  2.35,  2.5,   '',
    3.65, 2.5,  4.35,  2.5,   '',
    5.65, 2.5,  6.45,  2.5,   '',
    6.45, 1.55, 5.65,  1.55,  'gradient',
    4.35, 1.55, 3.65,  1.55,  'gradient',
    2.35, 1.55, 1.65,  1.55,  'gradient'
  )

  ggplot() +
    geom_label(
      data = boxes,
      aes(x = x, y = y, label = label, fill = fill),
      size = 5,
      label.size = 0.35,
      color = slate,
      family = 'Zilla Slab'
    ) +
    geom_segment(
      data = arrows,
      aes(x = x, y = y, xend = xend, yend = yend),
      arrow = arrow(length = unit(0.18, 'inches')),
      color = grey_dark,
      linewidth = 0.5
    ) +
    geom_text(
      data = arrows |> filter(label != ''),
      aes(x = (x + xend) / 2, y = y + 0.25, label = label),
      color = red_pink,
      family = 'Zilla Slab'
    ) +
    scale_fill_identity() +
    coord_cartesian(xlim = c(0, 8), ylim = c(1, 3.5), clip = 'off') +
    theme_void()
}

# Visualize a toy convolution on a small image patch.
make_convolution_demo_plot = function() {
  input_mat = matrix(c(
    0, 0, 1, 1, 0,
    0, 0, 1, 1, 0,
    0, 0, 1, 1, 0,
    0, 0, 1, 1, 0,
    0, 0, 1, 1, 0
  ), nrow = 5, byrow = TRUE)

  kernel_mat = matrix(c(
    -1, 0, 1,
    -1, 0, 1,
    -1, 0, 1
  ), nrow = 3, byrow = TRUE)

  feature_mat = matrix(0, nrow = 3, ncol = 3)
  for (i in 1:3) {
    for (j in 1:3) {
      feature_mat[i, j] = sum(input_mat[i:(i + 2), j:(j + 2)] * kernel_mat)
    }
  }

  matrix_df = function(mat) {
    tidyr::crossing(row = seq_len(nrow(mat)), col = seq_len(ncol(mat))) |>
      mutate(
        value = purrr::map2_dbl(row, col, ~ mat[.x, .y]),
        x = col,
        y = nrow(mat) - row + 1
      )
  }

  input_df = matrix_df(input_mat)
  kernel_df = matrix_df(kernel_mat)
  feature_df = matrix_df(feature_mat)

  input_hi = input_df |> filter(row %in% 2:4, col %in% 2:4)
  feature_hi = feature_df |> filter(row == 2, col == 2)

  input_plot =
    ggplot(input_df, aes(x = x, y = y, fill = value)) +
    geom_tile(color = 'white', linewidth = 0.9) +
    geom_text(
      aes(label = value, color = value > 0.5),
      family = 'Zilla Slab',
      size = 6,
      show.legend = FALSE
    ) +
    geom_tile(
      data = input_hi,
      fill = NA,
      color = red_pink,
      linewidth = 1.2
    ) +
    scale_fill_gradient(low = 'grey95', high = slate, guide = 'none') +
    scale_color_manual(values = c('FALSE' = slate, 'TRUE' = 'white')) +
    coord_equal() +
    labs(title = 'Input image') +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5, family = 'Zilla Slab', size = 18))

  kernel_plot =
    ggplot(kernel_df, aes(x = x, y = y, fill = value)) +
    geom_tile(color = 'white', linewidth = 0.9) +
    geom_text(aes(label = value), family = 'Zilla Slab', size = 6, color = slate) +
    scale_fill_gradient2(low = blue, mid = 'white', high = orange, midpoint = 0, guide = 'none') +
    coord_equal() +
    labs(title = '3 x 3 filter') +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5, family = 'Zilla Slab', size = 18))

  feature_plot =
    ggplot(feature_df, aes(x = x, y = y, fill = value)) +
    geom_tile(color = 'white', linewidth = 0.9) +
    geom_text(aes(label = value), family = 'Zilla Slab', size = 6, color = slate) +
    geom_tile(
      data = feature_hi,
      fill = NA,
      color = turquoise,
      linewidth = 1.2
    ) +
    scale_fill_gradient(low = 'grey95', high = turquoise, guide = 'none') +
    coord_equal() +
    labs(title = 'Feature map') +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5, family = 'Zilla Slab', size = 18))

  input_plot + kernel_plot + feature_plot + plot_layout(widths = c(1.25, 1, 1))
}

# Convert flat pixel rows into image tensors for a simple CNN.
make_image_tensor = function(df) {
  n = nrow(df)
  x = array(0, dim = c(n, 1, 28, 28))
  pixel_mat = as.matrix(df[, pixel_cols]) / 255

  for (i in seq_len(n)) {
    x[i, 1, , ] = matrix(pixel_mat[i, ], nrow = 28, byrow = TRUE)
  }

  torch_tensor(x, dtype = torch_float())
}

make_label_tensor = function(df) {
  torch_tensor(as.integer(as.character(df$label)) + 1L, dtype = torch_long())
}

simple_cnn = nn_module(
  'simple_cnn',
  initialize = function() {
    self$conv1 = nn_conv2d(1, 8, kernel_size = 3, padding = 1)
    self$conv2 = nn_conv2d(8, 16, kernel_size = 3, padding = 1)
    self$pool = nn_max_pool2d(kernel_size = 2)
    self$fc1 = nn_linear(16 * 7 * 7, 64)
    self$drop = nn_dropout(p = 0.2)
    self$fc2 = nn_linear(64, 10)
  },
  forward = function(x) {
    x = self$conv1(x)
    x = nnf_relu(x)
    x = self$pool(x)
    x = self$conv2(x)
    x = nnf_relu(x)
    x = self$pool(x)
    x = torch_flatten(x, start_dim = 2)
    x = self$fc1(x)
    x = nnf_relu(x)
    x = self$drop(x)
    self$fc2(x)
  }
)

fit_simple_cnn = function(train_df, epochs = 6, batch_size = 64, learn_rate = 0.003) {
  torch_manual_seed(111)
  net = simple_cnn()
  net$to(dtype = torch_float())
  opt = optim_adam(net$parameters, lr = learn_rate)
  loss_fn = nn_cross_entropy_loss()
  train_ds = tensor_dataset(make_image_tensor(train_df), make_label_tensor(train_df))
  train_dl = dataloader(train_ds, batch_size = batch_size, shuffle = TRUE)

  for (epoch in seq_len(epochs)) {
    net$train()
    coro::loop(for (batch in train_dl) {
      x_batch = batch[[1]]$to(dtype = torch_float())
      y_batch = batch[[2]]$to(dtype = torch_long())
      opt$zero_grad()
      logits = net(x_batch)
      loss = loss_fn(logits, y_batch)
      loss$backward()
      opt$step()
    })
  }

  net
}

predict_simple_cnn = function(net, df, batch_size = 256) {
  label_levels = levels(df$label)
  ds = tensor_dataset(make_image_tensor(df), make_label_tensor(df))
  dl = dataloader(ds, batch_size = batch_size, shuffle = FALSE)
  probs_list = list()

  with_no_grad({
    net$eval()
    coro::loop(for (batch in dl) {
      logits = net(batch[[1]]$to(dtype = torch_float()))
      probs = nnf_softmax(logits, dim = 2)
      probs_list[[length(probs_list) + 1]] = as.matrix(as.array(probs))
    })
  })

  probs = do.call(rbind, probs_list)
  prob_df = as_tibble(probs, .name_repair = 'minimal')
  names(prob_df) = paste0('.pred_', 0:9)

  tibble(
    .pred_class = factor(max.col(probs, ties.method = 'first') - 1L, levels = label_levels)
  ) |>
    bind_cols(prob_df)
}

# Build a manageable MNIST subset for lecture-time fitting.
mnist_small = train_dt |>
  mutate(label = factor(label)) |>
  group_by(label) |>
  slice_sample(n = 500) |>
  ungroup()

# Keep train / validation / test separate so tuning claims are honest.
mnist_split = initial_validation_split(
  mnist_small,
  prop = c(train = 0.6, validate = 0.2),
  strata = label
)

mnist_train = training(mnist_split)
mnist_valid = validation(mnist_split)
mnist_test = testing(mnist_split)
mnist_train_valid = bind_rows(mnist_train, mnist_valid)

# Baseline recipe: modest PCA compression.
rec_base =
  recipe(label ~ ., data = mnist_train) |>
  step_zv(all_predictors()) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 40)

# Tuned recipe: keep richer PCs and append a UMAP embedding.
rec_tuned =
  recipe(label ~ ., data = mnist_train) |>
  step_zv(all_predictors()) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 60) |>
  embed::step_umap(
    all_predictors(),
    num_comp = 8,
    neighbors = 30,
    min_dist = 0.05,
    keep_original_cols = TRUE,
    prefix = 'umap'
  )

# Final refit recipes use train + validation together.
rec_base_tv =
  recipe(label ~ ., data = mnist_train_valid) |>
  step_zv(all_predictors()) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 40)

rec_tuned_tv =
  recipe(label ~ ., data = mnist_train_valid) |>
  step_zv(all_predictors()) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 60) |>
  embed::step_umap(
    all_predictors(),
    num_comp = 8,
    neighbors = 30,
    min_dist = 0.05,
    keep_original_cols = TRUE,
    prefix = 'umap'
  )

# Multinomial logit is the 'no hidden layer' baseline.
logit_spec =
  multinom_reg(penalty = 0) |>
  set_engine('nnet', trace = FALSE, MaxNWts = 20000) |>
  set_mode('classification')

# Starter neural net: explicit ReLU hidden layer.
mlp_start_spec =
  mlp(hidden_units = 15, penalty = 0.01, epochs = 50, activation = 'relu') |>
  set_engine('brulee') |>
  set_mode('classification')

# Tuned neural net: same ReLU activation with a larger network.
mlp_tuned_spec =
  mlp(hidden_units = 50, penalty = 0.001, epochs = 150, activation = 'relu') |>
  set_engine('brulee') |>
  set_mode('classification')
```

---
name: admin
# Admin
## Today

.b[Topics]

- Logistic regression as a one-neuron network
- Activation functions, hidden layers, and forward propagation
- Cross-entropy, gradient descent, and backpropagation
- MNIST with MLPs and a simple CNN

## Reading

- .note[Core] Nielsen, .it[Neural Networks and Deep Learning], Ch. [1](http://neuralnetworksanddeeplearning.com/chap1.html)
- .note[Supplement] Ch. [2](http://neuralnetworksanddeeplearning.com/chap2.html) and [3](http://neuralnetworksanddeeplearning.com/chap3.html)
- .note[Course text] .it[ISL] Ch. 11

## Interactive

- TensorFlow's [Neural Network Playground](https://playground.tensorflow.org/)

---
layout: true
# Neural Networks

---
class: inverse, middle
name: intro

## Neural-net intuition

---
## Logistic regression is already a tiny neural network

For a binary outcome,

$$
\Pr(Y = 1 \mid X = x) = \sigma(\beta_0 + x'\beta)
$$

where $\sigma(z) = \dfrac{1}{1 + e^{-z}}$.

That is:

- a weighted sum of inputs
- a bias term
- an activation function
- one output

So neural networks are not a sharp break from logistic regression. They are a .attn[generalization].

---
## One neuron = weighted sum + activation

```{r, logistic-neuron, echo = FALSE}
make_single_neuron_plot()
```

---
## What do the weights and bias do?

The weighted sum

$$
z = b + \sum_{j = 1}^p w_j x_j
$$

does two jobs:

- the .attn[weights] say which inputs matter and in which direction
- the .attn[bias] shifts the cutoff left or right

So even before hidden layers, a neuron is already learning a scoring rule.

---
## Start with the perceptron

The .attn[perceptron] uses a hard threshold:

$$
a =
\begin{cases}
1 & \text{if } z \ge 0 \\
0 & \text{if } z < 0
\end{cases}
$$

That is intuitive, but not smooth.

The sigmoid neuron keeps the same weighted-sum idea and replaces the hard step with a differentiable activation.

---
## Perceptron vs. sigmoid neuron

```{r, step-v-sigmoid, echo = FALSE}
activation_df |>
  filter(activation %in% c('Step', 'Sigmoid')) |>
  ggplot(aes(x = x, y = y, color = activation)) +
  geom_hline(yintercept = 0, linewidth = 0.25) +
  geom_vline(xintercept = 0, linewidth = 0.25) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c(Step = slate, Sigmoid = red_pink)) +
  labs(
    x = 'z',
    y = 'activation',
    color = '',
    title = 'A sigmoid is a smooth version of a threshold rule'
  ) +
  theme_minimal(base_size = 18)
```

---
## Why stack neurons?

The point is not .it[more coefficients].

It is:

- early neurons can respond to simple patterns
- later neurons can combine those into more complex patterns
- composition can build feature detectors

For images, that can mean:

- pixels
- edges / strokes
- loops / corners
- whole digits

---
## A feed-forward network

```{r, network-plot, echo = FALSE}
make_network_plot()
```

---
## Hidden layers need nonlinearity

Suppose we stack two .b[linear] maps:

$$
a^{(1)} = W^{(1)}x + b^{(1)}
\qquad\text{and}\qquad
\hat y = W^{(2)}a^{(1)} + b^{(2)}
$$

Then

$$
\hat y = W^{(2)}W^{(1)}x + \big(W^{(2)}b^{(1)} + b^{(2)}\big)
$$

which is still just one linear map.

So hidden layers matter only when we add .attn[nonlinear activations].

---
## Common activation functions

```{r, activation-functions, echo = FALSE}
activation_df |>
  filter(activation %in% c('Identity', 'Sigmoid', 'Tanh', 'ReLU')) |>
  ggplot(aes(x = x, y = y, color = activation)) +
  geom_hline(yintercept = 0, linewidth = 0.25) +
  geom_vline(xintercept = 0, linewidth = 0.25) +
  geom_line(linewidth = 1.1, show.legend = FALSE) +
  facet_wrap(~activation, scales = 'free_y') +
  scale_color_manual(values = c(
    Identity = slate,
    Sigmoid = red_pink,
    Tanh = orange,
    ReLU = blue
  )) +
  theme_minimal(base_size = 17) +
  labs(x = 'z', y = 'g(z)')
```

---
## The derivatives matter too

```{r, activation-derivatives, echo = FALSE}
activation_deriv_df |>
  ggplot(aes(x = x, y = y, color = derivative)) +
  geom_hline(yintercept = 0, linewidth = 0.25) +
  geom_line(linewidth = 1.0, show.legend = FALSE) +
  facet_wrap(~derivative, scales = 'free_y') +
  scale_color_manual(values = c(
    'sigmoid\'(z)' = red_pink,
    'tanh\'(z)' = orange,
    'ReLU\'(z)' = blue
  )) +
  theme_minimal(base_size = 17) +
  labs(x = 'z', y = 'g\'(z)')
```

---
## Why those derivatives matter

The key derivatives are:

$$
\sigma'(z) = \sigma(z)\big(1 - \sigma(z)\big)
$$

$$
\tanh'(z) = 1 - \tanh^2(z)
$$

$$
\text{ReLU}'(z) = \mathbf{1}\{z > 0\}
$$

Why do we care?

- gradient-based training keeps multiplying by these derivatives
- if the derivative is tiny, updates can shrink
- if the derivative is zero, learning can stall locally

---
## Output activations should match the task

.col-left[
.hi-pink[Regression]

- often use identity output
- prediction is unrestricted
]

.hi-slate[Binary classification]

- often use sigmoid output
- prediction is a probability
]

.hi-orange[Multiclass classification]

- often use softmax output
- prediction is a vector of class probabilities
]

---
## Forward propagation

At layer $\ell$:

$$
z^{(\ell)} = W^{(\ell)}a^{(\ell-1)} + b^{(\ell)}
\qquad\text{and}\qquad
a^{(\ell)} = g^{(\ell)}\big(z^{(\ell)}\big)
$$

So the network repeats the same recipe:

1. weighted sums
1. activations
1. pass the result forward

---
## A dimension check helps

If:

- $a^{(\ell-1)}$ is $m_{\ell-1} \times 1$
- $W^{(\ell)}$ is $m_\ell \times m_{\ell-1}$
- $b^{(\ell)}$ is $m_\ell \times 1$

then:

$$
z^{(\ell)} = W^{(\ell)}a^{(\ell-1)} + b^{(\ell)}
$$

is also $m_\ell \times 1$.

That bookkeeping matters once the network gets deep.

---
## Softmax for multiclass output

If there are $K$ classes, the output layer produces

$$
\hat p_k = \frac{\exp(z_k)}{\sum_{j=1}^K \exp(z_j)}
$$

for $k = 1, \ldots, K$.

Properties:

- each $\hat p_k$ is between 0 and 1
- the probabilities sum to 1
- the largest one is the predicted class

---
class: inverse, middle
name: training

## How the network learns

---
## The loss function still runs the show

We choose weights and biases to make the loss small.

Common losses:

- regression: mean squared error
- binary classification: log loss
- multiclass classification: multiclass cross-entropy

For one multiclass observation:

$$
C = -\sum_{k = 1}^K y_k \log \hat p_k
$$

where $y_k$ is the one-hot target.

---
## Gradient descent

Once we have $C(\theta)$, we update

$$
\theta \leftarrow \theta - \eta \nabla_\theta C(\theta)
$$

where:

- $\theta$ collects all weights and biases
- $\eta$ is the learning rate

Same idea as before in the course, but now the parameter vector is much larger.

---
## Mini-batches and epochs

In practice we usually update on small batches:

- split the data into .attn[mini-batches]
- compute the gradient on one mini-batch
- update the weights
- repeat

One full pass through the training data is an .attn[epoch].

---
## Where do the gradients come from?

A weight in an early layer affects the loss through a chain:

- change a weight
- that changes a weighted sum
- that changes an activation
- that changes later activations and eventually the loss

So computing one gradient means following a .attn[chain of effects] through the network.

---
## One hidden weight affects loss indirectly

For one hidden-layer weight $w$,

$$
\frac{\partial C}{\partial w}
=
\frac{\partial C}{\partial \hat y}
\cdot
\frac{\partial \hat y}{\partial a^{(1)}}
\cdot
\frac{\partial a^{(1)}}{\partial z^{(1)}}
\cdot
\frac{\partial z^{(1)}}{\partial w}
$$

Deeper networks have longer chains, but the idea is the same.

---
## Backprop in words

- run the .attn[forward pass] and compute the loss
- start at the output and measure how wrong it is
- pass that error signal backward, one layer at a time
- combine those signals with local derivatives
- use the result to update weights and biases

Backprop is just an efficient way to organize that repeated chain rule.

---
## Backpropagation reuses the chain rule

```{r, backprop-plot, echo = FALSE}
make_backprop_plot()
```

---
## Start at the output layer

Write the output-layer error signal as:

$$
\delta^{(L)} =
\frac{\partial C}{\partial a^{(L)}} \odot g'\big(z^{(L)}\big)
$$

For .attn[softmax + cross-entropy], this simplifies nicely:

$$
\delta^{(L)} = \hat p - y
$$

That simplification is one reason cross-entropy is so common in classification.

---
## Then move backward through the hidden layers

For earlier layers,

$$
\delta^{(\ell)} =
\left(W^{(\ell+1)}\right)' \delta^{(\ell+1)}
\odot g'\big(z^{(\ell)}\big)
$$

So each layer receives an error signal from the layer after it, then adjusts it using its own derivative.

That is backpropagation in one sentence.

---
## Then turn those signals into gradients

Once we have the layer errors, the gradients are local:

$$
\frac{\partial C}{\partial W^{(\ell)}} =
\delta^{(\ell)} \left(a^{(\ell-1)}\right)'
\qquad\text{and}\qquad
\frac{\partial C}{\partial b^{(\ell)}} = \delta^{(\ell)}
$$

Then update:

$$
W^{(\ell)} \leftarrow W^{(\ell)} - \eta \frac{\partial C}{\partial W^{(\ell)}}
$$

and similarly for the biases.

---
## What activation are we using here?

For the MLP slide example we now use:

- `mlp(..., activation = 'relu') |> set_engine('brulee')`

That means:

- the .attn[`activation`] argument applies to the hidden layer, not the output layer
- for a multiclass factor response, the .attn[output layer] is still chosen by the task
- here the hidden units use .attn[ReLU], which is closer to a modern neural-net default than the old `nnet` setup

So the MLP example is now a .b[ReLU hidden-layer network], while the multinomial logit baseline remains the no-hidden-layer comparison.

---
## There's a lot more to training

Once the basic setup is in place, training quality also depends on:

- cross-entropy versus squared error
- regularization
- weight initialization
- architectural choices

So the story is not just .it[add a hidden layer].

The details of optimization matter.

---
## Try the playground

TensorFlow's [Neural Network Playground](https://playground.tensorflow.org/) is a very good way to build intuition.

Try changing:

- the activation function
- the number of hidden units
- the learning rate
- the amount of regularization

Watch how the boundary changes as you add nonlinear layers.

---
class: inverse, middle
name: mnist

## MNIST application

---
## Why MNIST again?

MNIST is a natural neural-net example because it has:

- 784 predictors for each image
- 10 output classes
- lots of nonlinear variation in handwriting

So it gives us both .attn[high-dimensional inputs] and .attn[multiclass outputs].

---
## The raw inputs are still images

```{r, mnist-examples, echo = FALSE}
idx = c(2, 1, 17, 8, 4, 9, 22, 7, 11)
plots = lapply(idx, function(i) {
  plot_digit(train_dt[i, ], title = paste('Label =', train_dt$label[i]))
})

wrap_plots(plots, ncol = 3) +
  plot_layout(guides = 'collect') &
  theme(
    legend.position = 'bottom',
    legend.key.width = unit(2, 'cm'),
    plot.title = element_text(size = 15, hjust = 0.5)
  )
```

---
## One image, many outputs

For each image, the output layer should produce:

$$
\big(
\hat p_0,\,
\hat p_1,\,
\ldots,\,
\hat p_9
\big)
$$

That is why MNIST is a .attn[multi-output classification] problem.

If we remove the hidden layer and keep only the softmax output, we are back at .attn[multinomial logistic regression].

---
## What a convolution does

- choose a small filter, often `3 x 3`
- place it over one local patch of pixels
- multiply and add to get one score
- slide the .attn[same filter] across the whole image

So one filter asks the same question everywhere:

- is there a vertical stroke here?
- is there an edge or a corner here?

In practice, the filter weights are .attn[learned from data].

---
## One tiny convolution example

Take a local patch

$$
X =
\begin{bmatrix}
0 & 1 & 1 \\
0 & 1 & 1 \\
0 & 1 & 1
\end{bmatrix}
\qquad\text{and filter}\qquad
K =
\begin{bmatrix}
-1 & 0 & 1 \\
-1 & 0 & 1 \\
-1 & 0 & 1
\end{bmatrix}
$$

Then the score at that location is

$$
\sum_{u,v} X_{uv} K_{uv} = 3
$$

So this filter responds when the patch is dark on the left and bright on the right:
a simple .attn[vertical-edge detector].

---
## A filter sliding over an image

The highlighted `3 x 3` patch on the left generates the highlighted output cell on the right.

```{r, convolution-demo, echo = FALSE}
make_convolution_demo_plot()
```

---
## A more ambitious workflow

To make the example more realistic, we will:

1. build a train / validation / test split
1. compare multinomial logit to a starter MLP
1. tune the MLP's capacity and preprocessing
1. see whether a very simple CNN does better on raw images
1. refit the better models on train + validation
1. evaluate on the held-out test set

This is still small, but it is now a real modeling workflow.

---
## The starter and tuned models

```{r, model-code, eval = FALSE}
# Baseline recipe: normalize and keep 40 PCs.
rec_base =
  recipe(label ~ ., data = mnist_train) |>
  step_zv(all_predictors()) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 40)

# Tuned recipe: keep 60 PCs and add 8 UMAP coordinates.
rec_tuned =
  recipe(label ~ ., data = mnist_train) |>
  step_zv(all_predictors()) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 60) |>
  embed::step_umap(
    all_predictors(),
    num_comp = 8,
    neighbors = 30,
    min_dist = 0.05,
    keep_original_cols = TRUE,
    prefix = 'umap'
  )

# Starter MLP: small network with ReLU hidden units.
mlp_start_spec =
  mlp(hidden_units = 15, penalty = 0.01, epochs = 50, activation = 'relu') |>
  set_engine('brulee') |>
  set_mode('classification')

# Tuned MLP: more hidden units, lighter penalty, more epochs.
mlp_tuned_spec =
  mlp(hidden_units = 50, penalty = 0.001, epochs = 150, activation = 'relu') |>
  set_engine('brulee') |>
  set_mode('classification')

# Simple CNN: two small conv layers on the raw 28 x 28 image.
# conv -> relu -> pool -> conv -> relu -> pool -> dense -> 10 logits
```

---
## A very simple CNN

- keep the raw 28 x 28 image instead of PCA or UMAP features
- use small .attn[3 x 3 filters] to scan local patches
- reuse the same filter weights across the image
- then flatten and map to 10 output classes

That is enough to exploit .attn[spatial structure] that the MLP does not see directly.

---
## Fit the models

```{r, mnist-fit, cache = TRUE}
# Fit each workflow on the training split.
fit_logit_valid =
  workflow() |>
  add_recipe(rec_base) |>
  add_model(logit_spec) |>
  fit(mnist_train)

fit_mlp_start_valid =
  workflow() |>
  add_recipe(rec_base) |>
  add_model(mlp_start_spec) |>
  fit(mnist_train)

fit_mlp_tuned_valid =
  workflow() |>
  add_recipe(rec_tuned) |>
  add_model(mlp_tuned_spec) |>
  fit(mnist_train)

fit_cnn_valid =
  fit_simple_cnn(mnist_train, epochs = 6)

# Collect validation predictions for model comparison.
pred_logit_valid =
  bind_cols(
    mnist_valid |> select(label),
    predict(fit_logit_valid, mnist_valid)
  )

pred_mlp_start_valid =
  bind_cols(
    mnist_valid |> select(label),
    predict(fit_mlp_start_valid, mnist_valid)
  )

pred_mlp_tuned_valid =
  bind_cols(
    mnist_valid |> select(label),
    predict(fit_mlp_tuned_valid, mnist_valid)
  )

pred_cnn_valid =
  bind_cols(
    mnist_valid |> select(label),
    predict_simple_cnn(fit_cnn_valid, mnist_valid)
  )

validation_compare = tibble(
  model = c('Multinomial logit', 'Starter MLP', 'Tuned MLP', 'Simple CNN'),
  accuracy = c(
    pred_logit_valid |> metrics(truth = label, estimate = .pred_class) |> filter(.metric == 'accuracy') |> pull(.estimate),
    pred_mlp_start_valid |> metrics(truth = label, estimate = .pred_class) |> filter(.metric == 'accuracy') |> pull(.estimate),
    pred_mlp_tuned_valid |> metrics(truth = label, estimate = .pred_class) |> filter(.metric == 'accuracy') |> pull(.estimate),
    pred_cnn_valid |> metrics(truth = label, estimate = .pred_class) |> filter(.metric == 'accuracy') |> pull(.estimate)
  )
)

# Refit the strongest models on train + validation.
fit_logit_test =
  workflow() |>
  add_recipe(rec_base_tv) |>
  add_model(logit_spec) |>
  fit(mnist_train_valid)

fit_mlp_tuned_test =
  workflow() |>
  add_recipe(rec_tuned_tv) |>
  add_model(mlp_tuned_spec) |>
  fit(mnist_train_valid)

fit_cnn_test =
  fit_simple_cnn(mnist_train_valid, epochs = 6)

# Evaluate on the held-out test split.
pred_logit_test =
  bind_cols(
    mnist_test |> select(label),
    predict(fit_logit_test, mnist_test)
  )

pred_mlp_tuned_test =
  bind_cols(
    mnist_test |> select(label),
    predict(fit_mlp_tuned_test, mnist_test),
    predict(fit_mlp_tuned_test, mnist_test, type = 'prob')
  )

pred_cnn_test =
  bind_cols(
    mnist_test |> select(label),
    predict_simple_cnn(fit_cnn_test, mnist_test)
  )

test_compare = tibble(
  model = c('Multinomial logit', 'Tuned MLP', 'Simple CNN'),
  accuracy = c(
    pred_logit_test |> metrics(truth = label, estimate = .pred_class) |> filter(.metric == 'accuracy') |> pull(.estimate),
    pred_mlp_tuned_test |> metrics(truth = label, estimate = .pred_class) |> filter(.metric == 'accuracy') |> pull(.estimate),
    pred_cnn_test |> metrics(truth = label, estimate = .pred_class) |> filter(.metric == 'accuracy') |> pull(.estimate)
  )
)

# Build a confusion matrix and one probability-vector example for the CNN.
cm_cnn_test =
  pred_cnn_test |>
  count(label, .pred_class)

example_prob_row =
  pred_cnn_test |>
  mutate(case_id = row_number()) |>
  rowwise() |>
  mutate(max_prob = max(c_across(matches('^\\.pred_[0-9]+$')))) |>
  ungroup() |>
  filter(label == .pred_class) |>
  slice_max(order_by = max_prob, n = 1)

example_case =
  mnist_test |> slice(example_prob_row$case_id)

example_prob_plot_df =
  example_prob_row |>
  select(matches('^\\.pred_[0-9]+$')) |>
  pivot_longer(
    everything(),
    names_to = 'digit',
    values_to = 'probability'
  ) |>
  mutate(digit = gsub('.pred_', '', digit, fixed = TRUE))
```

---
## Validation results

The first pass looks like this:

- multinomial logit: .attn[`r scales::percent(validation_compare$accuracy[validation_compare$model == 'Multinomial logit'], accuracy = 0.1)`]
- starter MLP: .attn[`r scales::percent(validation_compare$accuracy[validation_compare$model == 'Starter MLP'], accuracy = 0.1)`]
- tuned MLP: .attn[`r scales::percent(validation_compare$accuracy[validation_compare$model == 'Tuned MLP'], accuracy = 0.1)`]
- simple CNN: .attn[`r scales::percent(validation_compare$accuracy[validation_compare$model == 'Simple CNN'], accuracy = 0.1)`]

So the lesson is not .it[any hidden layer wins].

The lesson is that .attn[capacity, preprocessing, and architecture all matter].

---
## Validation comparison

```{r, validation-compare-plot, echo = FALSE}
ggplot(validation_compare, aes(x = model, y = accuracy, fill = model)) +
  geom_col(width = 0.62, show.legend = FALSE) +
  geom_text(
    aes(label = scales::percent(accuracy, accuracy = 0.1)),
    vjust = -0.35,
    family = 'Zilla Slab'
  ) +
  scale_fill_manual(values = c(
    'Multinomial logit' = slate,
    'Starter MLP' = grey_mid,
    'Tuned MLP' = red_pink,
    'Simple CNN' = turquoise
  )) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = NULL, y = 'Validation accuracy', title = 'A simple CNN improves again') +
  theme_minimal(base_size = 18)
```

---
## Held-out test results

After refitting on train + validation:

- multinomial logit gets about .attn[`r scales::percent(test_compare$accuracy[test_compare$model == 'Multinomial logit'], accuracy = 0.1)`]
- the tuned MLP gets about .attn[`r scales::percent(test_compare$accuracy[test_compare$model == 'Tuned MLP'], accuracy = 0.1)`]
- the simple CNN gets about .attn[`r scales::percent(test_compare$accuracy[test_compare$model == 'Simple CNN'], accuracy = 0.1)`]

So the .attn[CNN improves again] once we let the model use the image's 2D structure.

---
## Test comparison

```{r, test-compare-plot, echo = FALSE}
ggplot(test_compare, aes(x = model, y = accuracy, fill = model)) +
  geom_col(width = 0.62, show.legend = FALSE) +
  geom_text(
    aes(label = scales::percent(accuracy, accuracy = 0.1)),
    vjust = -0.35,
    family = 'Zilla Slab'
  ) +
  scale_fill_manual(values = c(
    'Multinomial logit' = slate,
    'Tuned MLP' = red_pink,
    'Simple CNN' = turquoise
  )) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = NULL, y = 'Test accuracy', title = 'A simple CNN does better still') +
  theme_minimal(base_size = 18)
```

---
## A single image, many probabilities

```{r, multi-output-example, echo = FALSE}
plot_digit(example_case, title = paste('True digit =', example_case$label[[1]])) +
  ggplot(example_prob_plot_df, aes(x = digit, y = probability, fill = digit == example_case$label[[1]])) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c('TRUE' = red_pink, 'FALSE' = slate)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    x = 'Output node',
    y = 'Predicted probability',
    title = 'Softmax output for one CNN prediction'
  ) +
  theme_minimal(base_size = 18)
```

---
## Confusion matrix for the simple CNN

```{r, cnn-cm, echo = FALSE}
ggplot(cm_cnn_test, aes(x = .pred_class, y = label, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c('') +
  coord_equal() +
  theme_minimal(base_size = 18) +
  labs(
    x = 'Predicted digit',
    y = 'True digit',
    title = 'Held-out test confusion matrix'
  )
```

---
## What this example does and does not show

It does show:

- real multiclass output
- real train / validation / test logic
- real improvement from a stronger architecture

It does not show:

- deep networks
- state-of-the-art MNIST performance

This is still a .attn[teaching example], just a better one.

---
class: inverse, middle
name: wrap

## Wrap-up

---
## Main takeaways

- Logistic regression is a neural network with no hidden layer
- Hidden layers matter only when paired with nonlinear activations
- Backpropagation is efficient chain rule
- Softmax turns multiclass scores into probabilities
- Convolutions reuse small learned filters across the image
- A small untuned MLP may only modestly improve on the baseline
- A better-specified MLP can do better, even in a simple `tidymodels` workflow
- A very simple CNN can do better still when it uses image structure

Next step: deeper architectures and more modern image models.
