---
title: 'Lecture .mono[011]'
subtitle: 'Neural networks'
author: 'Edward Rubin'
header-includes:
  - \usepackage{tikz} 
  - \usepackage{ifthen} 
  - \usetikzlibrary{positioning, shapes, arrows.meta}
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
exclude: true

```{r, setup, include = FALSE}
library(pacman)
p_load(
  magick,
  tidyverse,
  data.table,
  tidymodels,
  patchwork,
  scales,
  torch,
  coro
)

# Lecture color palette.
red_pink   = '#e64173'
turquoise  = '#20B2AA'
orange     = '#FFA500'
red        = '#fb6107'
blue       = '#181485'
navy       = '#150E37FF'
green      = '#8bb174'
yellow     = '#D8BD44'
grey_light = 'grey70'
grey_mid   = 'grey50'
grey_dark  = 'grey20'
purple     = '#6A5ACD'
slate      = '#314f4f'

knitr::opts_chunk$set(
  comment = '#>',
  fig.align = 'center',
  fig.height = 6.75,
  fig.width = 10.5,
  warning = FALSE,
  message = FALSE
)
knitr::opts_chunk$set(dev = 'svg')
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(knitr.table.format = 'html')
xaringanExtra::use_xaringan_extra(c('tile_view', 'fit_screen'))

set.seed(111)

# Load MNIST data from the local lecture folder.
train_dt = fread('train.csv') |> as_tibble()
test_dt = fread('test.csv') |> as_tibble()
pixel_cols = names(train_dt)[grepl('^pixel', names(train_dt))]

# Plot one 28 x 28 digit image from a single-row data frame.
plot_digit = function(row_df, title = NULL) {
  ggplot(
    expand_grid(y = 28:1, x = 1:28) |>
      mutate(value = as.numeric(row_df[, pixel_cols])),
    aes(x = x, y = y, fill = value)
  ) +
    geom_tile() +
    scale_fill_viridis_c('', option = 'magma', direction = -1) +
    coord_equal() +
    theme_void(base_family = 'Fira Sans Condensed', base_size = 18) +
    theme(legend.position = 'bottom') +
    labs(title = title)
}

# Activation functions for intuition and comparison plots.
activation_df = tibble(x = seq(-10, 10, length.out = 500)) |>
  mutate(
    Identity = x,
    Sigmoid = 1 / (1 + exp(-x)),
    Tanh = tanh(x),
    ReLU = pmax(0, x),
    Step = if_else(x >= 0, 1, 0)
  ) |>
  pivot_longer(-x, names_to = 'activation', values_to = 'y')

# Common derivatives that appear inside backpropagation.
activation_deriv_df = tibble(x = seq(-5, 5, length.out = 500)) |>
  mutate(
    `sigmoid''(z)` = (1 / (1 + exp(-x))) * (1 - 1 / (1 + exp(-x))),
    `tanh''(z)` = 1 - tanh(x)^2,
    `ReLU''(z)` = if_else(x > 0, 1, 0)
  ) |>
  pivot_longer(-x, names_to = 'derivative', values_to = 'y')

# Draw a simple fully connected network for lecture visuals.
make_single_neuron_plot = function() {
  nodes = tribble(
    ~name,    ~x,  ~y,  ~label, ~fill_color, ~text_color, ~size,
    'bias',   1.2, 5.0, '1',    'white',     slate,       5.0,
    'x₁',     1.2, 4.0, 'x₁',   'white',     slate,       5.0,
    'x₂',     1.2, 3.0, 'x₂',   'white',     slate,       5.0,
    'xdots',  1.2, 2.0, '⋮',   'white',     slate,       5.0,
    'xₚ',     1.2, 1.0, 'xₚ',   'white',     slate,       5.0,
    'neuron', 4.2, 3.0, 'a',    red_pink,    'white',     5.3,
    'output', 7.2, 3.0, 'ŷ',    orange,      'white',     4.7
  )

  edges =
    tribble(
      ~from,    ~to,      ~weight,
      'bias',   'neuron', 'b',
      'x₁',     'neuron', 'w₁',
      'x₂',     'neuron', 'w₂',
      # 'xdots', 'neuron', '',
      'xₚ',     'neuron', 'wₚ',
      'neuron', 'output', ''
    ) |>
    left_join(nodes |> select(name, x_from = x, y_from = y), by = c('from' = 'name')) |>
    left_join(nodes |> select(name, x_to = x, y_to = y), by = c('to' = 'name')) |>
    mutate(
      label_x = 0.58 * x_from + 0.42 * x_to,
      label_y = 0.58 * y_from + 0.42 * y_to
    )

  ggplot() +
    geom_segment(
      data = edges,
      aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
      color = grey_light,
      linewidth = 0.55,
      arrow = arrow(length = unit(0.16, 'inches'))
    ) +
    geom_point(
      data = nodes |> filter(name != 'xdots'),
      aes(x = x, y = y, fill = fill_color),
      shape = 21,
      size = 11,
      color = slate,
      stroke = 1.1
    ) +
    geom_point(
      data = nodes |> filter(name == 'xdots'),
      aes(x = x, y = y, fill = fill_color),
      shape = 21,
      size = 11,
      color = 'white',
      stroke = 1.1
    ) +
    geom_text(
      data = nodes,
      aes(x = x, y = y, label = label, color = text_color, size = size),
      family = 'Fira Sans Condensed',
      fontface = 'bold',
      show.legend = FALSE
    ) +
    geom_label(
      data = edges |> filter(weight != ''),
      aes(x = label_x, y = label_y, label = weight),
      family = 'Fira Sans Condensed',
      fill = 'white',
      color = slate,
      label.size = 0,
      size = 4.1
    ) +
    annotate('text', x = 1.2, y = 5.75, label = 'Inputs', family = 'Fira Sans Condensed', face = 'bold', color = slate, size = 6) +
    annotate('text', x = 4.2, y = 5.75, label = 'Neuron', family = 'Fira Sans Condensed', face = 'bold', color = red_pink, size = 6) +
    annotate('text', x = 7.2, y = 5.75, label = 'Output', family = 'Fira Sans Condensed', face = 'bold', color = orange, size = 6) +
    annotate('text', x = 4.2, y = 4.15, label = "z = b + w'x", family = 'Fira Sans Condensed', color = slate, size = 6) +
    annotate('text', x = 4.2, y = 1.9, label = 'a = g(z)', family = 'Fira Sans Condensed', color = red_pink, size = 6) +
    # annotate('text', x = 8.25, y = 3.0, label = 'Prediction', family = 'Fira Sans Condensed', color = slate, hjust = 0, size = 6) +
    scale_fill_identity() +
    scale_color_identity() +
    scale_size_identity() +
    coord_cartesian(xlim = c(0.4, 9.2), ylim = c(0.4, 6.1), clip = 'off') +
    theme_void(base_family = 'Fira Sans Condensed')
}

make_network_plot = function() {
  nodes = tribble(
    ~layer,   ~name, ~x,  ~y,  ~label,     ~type,
    'input',  'x1',  1.0, 4.5, 'pixel 1',  'input',
    'input',  'x2',  1.0, 3.5, 'pixel 2',  'input',
    'input',  'xdot',  1.0, 2.5, '⋮',       'input',
    'input',  'x4',  1.0, 1.5, 'pixel 784','input',
    'hidden', 'h1',  3.2, 5.0, '',         'hidden',
    'hidden', 'h2',  3.2, 4.0, '',         'hidden',
    'hidden', 'h3',  3.2, 3.0, '',         'hidden',
    'hidden', 'h4',  3.2, 2.0, '',         'hidden',
    'hidden', 'h5',  3.2, 1.0, '',         'hidden',
    'output', 'o1',  5.5, 4.8, '0',        'output',
    'output', 'o2',  5.5, 3.9, '1',        'output',
    'output', 'o3',  5.5, 3.0, '2',        'output',
    'output', 'odot',  5.5, 2.1, '⋮',       'output',
    'output', 'o5',  5.5, 1.2, '9',        'output'
  )

  edges = expand_grid(
    from = nodes |> filter(layer == 'input' & str_detect(name, 'dot', negate = TRUE)) |> pull(name),
    to = nodes |> filter(layer == 'hidden' & str_detect(name, 'dot', negate = TRUE)) |> pull(name)
  ) |>
    bind_rows(
      expand_grid(
        from = nodes |> filter(layer == 'hidden' & str_detect(name, 'dot', negate = TRUE)) |> pull(name),
        to = nodes |> filter(layer == 'output' & str_detect(name, 'dot', negate = TRUE)) |> pull(name)
      )
    ) |>
    left_join(nodes |> select(name, x_from = x, y_from = y), by = c('from' = 'name')) |>
    left_join(nodes |> select(name, x_to = x, y_to = y), by = c('to' = 'name'))

  ggplot() +
    geom_segment(
      data = edges,
      aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
      color = grey_light,
      linewidth = 0.35,
      alpha = 0.75
    ) +
    geom_point(
      data = nodes |> filter(str_detect(name, 'dot', negate = TRUE)),
      aes(x = x, y = y, fill = type),
      shape = 21,
      size = 7,
      color = slate,
      stroke = 1
    ) +
    geom_text(
      data = nodes |> filter(layer == 'input'),
      aes(x = x - 0.45, y = y, label = label),
      hjust = .5,
      family = 'Fira Sans Condensed', color = slate, size = 5
    ) +
    geom_text(
      data = nodes |> filter(layer == 'output'),
      aes(x = x + 0.35, y = y, label = label),
      hjust = .5,
      family = 'Fira Sans Condensed', color = orange, size = 5
    ) +
    annotate(
      'text', x = .8, y = 5.6, label = 'Input layer',
      family = 'Fira Sans Condensed', color = slate, size = 7, fontface = 'bold'
    ) +
    annotate(
      'text', x = 3.2, y = 5.6, label = 'Hidden layer',
      family = 'Fira Sans Condensed', color = red_pink, size = 7, fontface = 'bold'
    ) +
    annotate(
      'text', x = 5.5, y = 5.6, label = 'Output layer',
      family = 'Fira Sans Condensed', color = orange, size = 7, fontface = 'bold'
    ) +
    scale_fill_manual(values = c(input = 'white', hidden = red_pink, output = orange)) +
    coord_cartesian(xlim = c(0, 6.5), ylim = c(0.5, 6.1), clip = 'off') +
    theme_void() +
    theme(legend.position = 'none')
}

# Draw a fully connected network with two hidden layers.
make_network_plot_two_hidden = function() {
  nodes = tribble(
    ~layer,    ~name, ~x,  ~y,  ~label,      ~type,
    'input',   'x1',  1.0, 4.5, 'pixel 1',   'input',
    'input',   'x2',  1.0, 3.5, 'pixel 2',   'input',
    'input',   'xdot',1.0, 2.5, '⋮',         'input',
    'input',   'x4',  1.0, 1.5, 'pixel 784', 'input',
    'hidden1', 'h1',  3.2, 5.0, '',          'hidden',
    'hidden1', 'h2',  3.2, 4.0, '',          'hidden',
    'hidden1', 'h3',  3.2, 3.0, '',          'hidden',
    'hidden1', 'h4',  3.2, 2.0, '',          'hidden',
    'hidden1', 'h5',  3.2, 1.0, '',          'hidden',
    'hidden2', 'h6',  5.4, 5.0, '',          'hidden',
    'hidden2', 'h7',  5.4, 4.0, '',          'hidden',
    'hidden2', 'h8',  5.4, 3.0, '',          'hidden',
    'hidden2', 'h9',  5.4, 2.0, '',          'hidden',
    'hidden2', 'h10', 5.4, 1.0, '',          'hidden',
    'output',  'o1',  7.7, 4.8, '0',         'output',
    'output',  'o2',  7.7, 3.9, '1',         'output',
    'output',  'o3',  7.7, 3.0, '2',         'output',
    'output',  'odot',7.7, 2.1, '⋮',         'output',
    'output',  'o5',  7.7, 1.2, '9',         'output'
  )

  edges = expand_grid(
    from = nodes |> filter(layer == 'input' & str_detect(name, 'dot', negate = TRUE)) |> pull(name),
    to = nodes |> filter(layer == 'hidden1') |> pull(name)
  ) |>
    bind_rows(
      expand_grid(
        from = nodes |> filter(layer == 'hidden1') |> pull(name),
        to = nodes |> filter(layer == 'hidden2') |> pull(name)
      )
    ) |>
    bind_rows(
      expand_grid(
        from = nodes |> filter(layer == 'hidden2') |> pull(name),
        to = nodes |> filter(layer == 'output' & str_detect(name, 'dot', negate = TRUE)) |> pull(name)
      )
    ) |>
    left_join(nodes |> select(name, x_from = x, y_from = y), by = c('from' = 'name')) |>
    left_join(nodes |> select(name, x_to = x, y_to = y), by = c('to' = 'name'))

  ggplot() +
    geom_segment(
      data = edges,
      aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
      color = grey_light,
      linewidth = 0.35,
      alpha = 0.75
    ) +
    geom_point(
      data = nodes |> filter(str_detect(name, 'dot', negate = TRUE)),
      aes(x = x, y = y, fill = type),
      shape = 21,
      size = 7,
      color = slate,
      stroke = 1
    ) +
    geom_text(
      data = nodes |> filter(layer == 'input'),
      aes(x = x - 0.45, y = y, label = label),
      hjust = .5,
      family = 'Fira Sans Condensed', color = slate, size = 5
    ) +
    geom_text(
      data = nodes |> filter(layer == 'output'),
      aes(x = x + 0.35, y = y, label = label),
      hjust = .5,
      family = 'Fira Sans Condensed', color = orange, size = 5
    ) +
    annotate(
      'text', x = .8, y = 5.6, label = 'Input layer',
      family = 'Fira Sans Condensed', color = slate, size = 7, fontface = 'bold'
    ) +
    annotate(
      'text', x = 3.2, y = 5.6, label = 'Hidden layer 1',
      family = 'Fira Sans Condensed', color = red_pink, size = 7, fontface = 'bold'
    ) +
    annotate(
      'text', x = 5.4, y = 5.6, label = 'Hidden layer 2',
      family = 'Fira Sans Condensed', color = red_pink, size = 7, fontface = 'bold'
    ) +
    annotate(
      'text', x = 7.7, y = 5.6, label = 'Output layer',
      family = 'Fira Sans Condensed', color = orange, size = 7, fontface = 'bold'
    ) +
    scale_fill_manual(values = c(input = 'white', hidden = red_pink, output = orange)) +
    coord_cartesian(xlim = c(0, 8.7), ylim = c(0.5, 6.1), clip = 'off') +
    theme_void() +
    theme(legend.position = 'none')
}

# Draw a fully connected feedforward network with flexible layer sizes.
plot_nn = function(
  input,
  hidden,
  output,
  max_input = input,
  max_output = output,
  label_input = NULL,
  label_output = NULL,
  label_layers = TRUE
) {
  is_pos_int = function(x) {
    is.numeric(x) && length(x) == 1 && !is.na(x) && x > 0 && x %% 1 == 0
  }

  if (!is_pos_int(input)) {
    stop('`input` must be a single positive integer.', call. = FALSE)
  }
  if (!is_pos_int(output)) {
    stop('`output` must be a single positive integer.', call. = FALSE)
  }
  if (!is_pos_int(max_input)) {
    stop('`max_input` must be a single positive integer.', call. = FALSE)
  }
  if (!is_pos_int(max_output)) {
    stop('`max_output` must be a single positive integer.', call. = FALSE)
  }
  label_layers_is_logical = is.logical(label_layers) && length(label_layers) == 1 && !is.na(label_layers)
  label_layers_is_character = is.character(label_layers)
  if (!label_layers_is_logical && !label_layers_is_character) {
    stop('`label_layers` must be TRUE/FALSE or a character vector of layer labels.', call. = FALSE)
  }

  input = as.integer(input)
  output = as.integer(output)
  max_input = as.integer(max_input)
  max_output = as.integer(max_output)

  if (length(hidden) == 0 || (length(hidden) == 1 && !is.na(hidden) && hidden == 0)) {
    hidden_sizes = integer(0)
  } else {
    if (!is.numeric(hidden) || any(is.na(hidden)) || any(hidden <= 0) || any(hidden %% 1 != 0)) {
      stop('`hidden` must be 0 or a vector of positive integers.', call. = FALSE)
    }
    hidden_sizes = as.integer(hidden)
  }

  if (max_input < input && max_input < 2) {
    stop('`max_input` must be at least 2 when it is less than `input`.', call. = FALSE)
  }
  if (max_output < output && max_output < 2) {
    stop('`max_output` must be at least 2 when it is less than `output`.', call. = FALSE)
  }

  input_draw_n = as.integer(min(input, max_input))
  output_draw_n = as.integer(min(output, max_output))
  label_input_short = FALSE
  label_output_short = FALSE

  if (!is.null(label_input)) {
    allowed_input_lengths = sort(unique(c(input, input_draw_n)))
    if (!(length(label_input) %in% allowed_input_lengths)) {
      stop(
        sprintf('`label_input` must have length %s.', paste(allowed_input_lengths, collapse = ' or ')),
        call. = FALSE
      )
    }
    label_input_short = (length(label_input) == input_draw_n && input_draw_n < input)
  }
  if (!is.null(label_output)) {
    allowed_output_lengths = sort(unique(c(output, output_draw_n)))
    if (!(length(label_output) %in% allowed_output_lengths)) {
      stop(
        sprintf('`label_output` must have length %s.', paste(allowed_output_lengths, collapse = ' or ')),
        call. = FALSE
      )
    }
    label_output_short = (length(label_output) == output_draw_n && output_draw_n < output)
  }

  if (!is.null(label_input)) {
    label_input = as.character(label_input)
  }
  if (!is.null(label_output)) {
    label_output = as.character(label_output)
  }

  layer_tbl = tibble(
    layer_order = seq_len(length(c(input, hidden_sizes, output))),
    layer_id = c(
      'input',
      if (length(hidden_sizes) > 0) paste0('hidden', seq_along(hidden_sizes)) else character(0),
      'output'
    ),
    layer_type = c('input', rep('hidden', length(hidden_sizes)), 'output'),
    true_n = c(input, hidden_sizes, output)
  ) |>
    mutate(
      max_draw = case_when(
        layer_type == 'input' ~ pmin(true_n, max_input),
        layer_type == 'output' ~ pmin(true_n, max_output),
        TRUE ~ true_n
      ),
      hidden_index = cumsum(layer_type == 'hidden'),
      layer_title = case_when(
        layer_type == 'input' ~ 'Input layer',
        layer_type == 'output' ~ 'Output layer',
        TRUE ~ paste('Hidden layer', hidden_index)
      )
    )

  if (label_layers_is_logical) {
    draw_layer_titles = label_layers
    layer_tbl = layer_tbl |>
      mutate(layer_title_draw = layer_title)
  } else {
    if (length(label_layers) != nrow(layer_tbl)) {
      stop(sprintf('When `label_layers` is character, it must have length %d.', nrow(layer_tbl)), call. = FALSE)
    }
    draw_layer_titles = TRUE
    layer_tbl = layer_tbl |>
      mutate(layer_title_draw = as.character(label_layers))
  }

  build_layer_nodes = function(layer_order, layer_id, layer_type, true_n, max_draw, x) {
    truncated = max_draw < true_n

    if (truncated) {
      first_idx = if (max_draw > 1) seq_len(max_draw - 1L) else integer(0)

      layer_nodes = bind_rows(
        tibble(
          display_index = c(first_idx, max_draw + 1L),
          true_index = c(first_idx, true_n),
          is_dot = FALSE
        ),
        tibble(
          display_index = max_draw,
          true_index = NA_integer_,
          is_dot = TRUE
        )
      ) |>
        arrange(display_index)
    } else {
      layer_nodes = tibble(
        display_index = seq_len(true_n),
        true_index = seq_len(true_n),
        is_dot = FALSE
      )
    }

    display_n = nrow(layer_nodes)

    layer_nodes |>
      mutate(
        layer_order = layer_order,
        layer_id = layer_id,
        layer_type = layer_type,
        display_n = display_n,
        x = x,
        draw_edge = !is_dot,
        node_id = paste0(layer_id, '_', if_else(is_dot, 'dot', as.character(true_index))),
        side_label = NA_character_
      )
  }

  layer_gap = 2.2
  x_start = 1.0
  y_center = 3.0

  layer_tbl = layer_tbl |>
    mutate(x = x_start + (layer_order - 1) * layer_gap)

  nodes = pmap_dfr(
    layer_tbl |> select(layer_order, layer_id, layer_type, true_n, max_draw, x),
    build_layer_nodes
  )

  max_drawn = max(nodes$display_n)
  y_step = if (max_drawn <= 1) 0 else min(1, 4 / (max_drawn - 1))

  nodes = nodes |>
    mutate(y = y_center + ((display_n + 1) / 2 - display_index) * y_step)

  if (!is.null(label_input)) {
    input_rows = which(nodes$layer_type == 'input' & !nodes$is_dot)
    if (label_input_short) {
      nodes$side_label[input_rows] = label_input[seq_along(input_rows)]
    } else {
      nodes$side_label[input_rows] = label_input[nodes$true_index[input_rows]]
    }
  }

  if (!is.null(label_output)) {
    output_rows = which(nodes$layer_type == 'output' & !nodes$is_dot)
    if (label_output_short) {
      nodes$side_label[output_rows] = label_output[seq_along(output_rows)]
    } else {
      nodes$side_label[output_rows] = label_output[nodes$true_index[output_rows]]
    }
  }

  edges = map(
    seq_len(nrow(layer_tbl) - 1),
    \(i) {
      from_nodes = nodes |>
        filter(layer_order == i, draw_edge) |>
        pull(node_id)

      to_nodes = nodes |>
        filter(layer_order == i + 1, draw_edge) |>
        pull(node_id)

      if (length(from_nodes) == 0 || length(to_nodes) == 0) {
        return(tibble(from = character(0), to = character(0)))
      }

      expand_grid(from = from_nodes, to = to_nodes)
    }
  ) |>
    bind_rows() |>
    left_join(nodes |> select(node_id, x_from = x, y_from = y), by = c('from' = 'node_id')) |>
    left_join(nodes |> select(node_id, x_to = x, y_to = y), by = c('to' = 'node_id'))

  title_y = max(nodes$y) + 0.6
  title_tbl = layer_tbl |>
    select(layer_type, layer_title_draw, x) |>
    mutate(y = title_y)

  n_layers = nrow(layer_tbl)
  extra_layers = max(n_layers - 3L, 0L)
  input_label_offset = 0.45 + 0.09 * extra_layers
  output_label_offset = 0.35 + 0.09 * extra_layers

  x_left_margin = if (is.null(label_input)) 0.8 else max(1.4, input_label_offset + 0.95)
  x_right_margin = if (is.null(label_output)) 0.8 else max(1.4, output_label_offset + 1.05)
  y_min = min(nodes$y) - 0.45
  y_max = if (draw_layer_titles) title_y + 0.35 else max(nodes$y) + 0.45

  p = ggplot() +
    geom_segment(
      data = edges,
      aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
      color = grey_light,
      linewidth = 0.35,
      alpha = 0.75
    ) +
    geom_point(
      data = nodes |> filter(!is_dot),
      aes(x = x, y = y, fill = layer_type),
      shape = 21,
      size = 7,
      color = slate,
      stroke = 1
    ) +
    scale_fill_manual(values = c(input = 'white', hidden = red_pink, output = orange)) +
    coord_cartesian(
      xlim = c(min(layer_tbl$x) - x_left_margin, max(layer_tbl$x) + x_right_margin),
      ylim = c(y_min, y_max),
      clip = 'off'
    ) +
    theme_void() +
    theme(legend.position = 'none')

  if (!is.null(label_input)) {
    p = p +
      geom_text(
        data = nodes |> filter(layer_type == 'input', !is_dot),
        aes(x = x - input_label_offset, y = y, label = side_label),
        hjust = .5,
        family = 'Fira Sans Condensed',
        color = slate,
        size = 5
      )
  }

  if (!is.null(label_output)) {
    p = p +
      geom_text(
        data = nodes |> filter(layer_type == 'output', !is_dot),
        aes(x = x + output_label_offset, y = y, label = side_label),
        hjust = .5,
        family = 'Fira Sans Condensed',
        color = orange,
        size = 5
      )
  }

  if (any(nodes$is_dot & nodes$layer_type == 'input')) {
    p = p +
      geom_text(
        data = nodes |> filter(layer_type == 'input', is_dot),
        aes(x = x, y = y),
        label = '⋮',
        family = 'Fira Sans Condensed',
        color = slate,
        size = 5
      )
  }

  if (any(nodes$is_dot & nodes$layer_type == 'output')) {
    p = p +
      geom_text(
        data = nodes |> filter(layer_type == 'output', is_dot),
        aes(x = x, y = y),
        label = '⋮',
        family = 'Fira Sans Condensed',
        color = orange,
        size = 5
      )
  }

  if (draw_layer_titles) {
    p = p +
      geom_text(
        data = title_tbl |> filter(layer_type == 'input'),
        aes(x = x, y = y, label = layer_title_draw),
        family = 'Fira Sans Condensed',
        color = slate,
        size = 7,
        fontface = 'bold'
      ) +
      geom_text(
        data = title_tbl |> filter(layer_type == 'hidden'),
        aes(x = x, y = y, label = layer_title_draw),
        family = 'Fira Sans Condensed',
        color = red_pink,
        size = 7,
        fontface = 'bold'
      ) +
      geom_text(
        data = title_tbl |> filter(layer_type == 'output'),
        aes(x = x, y = y, label = layer_title_draw),
        family = 'Fira Sans Condensed',
        color = orange,
        size = 7,
        fontface = 'bold'
      )
  }

  p
}

# Draw a forward-pass / backward-pass computational graph.
make_backprop_plot = function() {
  boxes = tribble(
    ~x,  ~y,  ~label,              ~fill,
    1.0, 2.5, 'Inputs\nx',         'white',
    3.0, 2.5, 'Weighted sum\nz',   '#fde5ec',
    5.0, 2.5, 'Activation\na',     '#eef4ff',
    7.1, 2.5, 'Loss\nC',           '#fff2db'
  )

  arrows = tribble(
    ~x,   ~y,   ~xend, ~yend, ~label,
    1.65, 2.5,  2.35,  2.5,   '',
    3.65, 2.5,  4.35,  2.5,   '',
    5.65, 2.5,  6.45,  2.5,   '',
    6.45, 1.55, 5.65,  1.55,  'gradient',
    4.35, 1.55, 3.65,  1.55,  'gradient',
    2.35, 1.55, 1.65,  1.55,  'gradient'
  )

  ggplot() +
    geom_label(
      data = boxes,
      aes(x = x, y = y, label = label, fill = fill),
      size = 5,
      label.size = 0.35,
      color = slate,
      family = 'Zilla Slab'
    ) +
    geom_segment(
      data = arrows,
      aes(x = x, y = y, xend = xend, yend = yend),
      arrow = arrow(length = unit(0.18, 'inches')),
      color = grey_dark,
      linewidth = 0.5
    ) +
    geom_text(
      data = arrows |> filter(label != ''),
      aes(x = (x + xend) / 2, y = y + 0.25, label = label),
      color = red_pink,
      family = 'Zilla Slab'
    ) +
    scale_fill_identity() +
    coord_cartesian(xlim = c(0, 8), ylim = c(1, 3.5), clip = 'off') +
    theme_void()
}

# Visualize a toy convolution on a small image patch.
make_convolution_demo_plot = function() {
  input_mat = matrix(c(
    0, 0, 1, 1, 0,
    0, 0, 1, 1, 0,
    0, 0, 1, 1, 0,
    0, 0, 1, 1, 0,
    0, 0, 1, 1, 0
  ), nrow = 5, byrow = TRUE)

  kernel_mat = matrix(c(
    -1, 0, 1,
    -1, 0, 1,
    -1, 0, 1
  ), nrow = 3, byrow = TRUE)

  feature_mat = matrix(0, nrow = 3, ncol = 3)
  for (i in 1:3) {
    for (j in 1:3) {
      feature_mat[i, j] = sum(input_mat[i:(i + 2), j:(j + 2)] * kernel_mat)
    }
  }

  matrix_df = function(mat) {
    tidyr::crossing(row = seq_len(nrow(mat)), col = seq_len(ncol(mat))) |>
      mutate(
        value = purrr::map2_dbl(row, col, ~ mat[.x, .y]),
        x = col,
        y = nrow(mat) - row + 1
      )
  }

  input_df = matrix_df(input_mat)
  kernel_df = matrix_df(kernel_mat)
  feature_df = matrix_df(feature_mat)

  input_hi = input_df |> filter(row %in% 2:4, col %in% 2:4)
  feature_hi = feature_df |> filter(row == 2, col == 2)

  input_plot =
    ggplot(input_df, aes(x = x, y = y, fill = value)) +
    geom_tile(color = 'white', linewidth = 0.9) +
    geom_text(
      aes(label = value, color = value > 0.5),
      family = 'Zilla Slab',
      size = 6,
      show.legend = FALSE
    ) +
    geom_tile(
      data = input_hi,
      fill = NA,
      color = red_pink,
      linewidth = 1.2
    ) +
    scale_fill_gradient(low = 'grey95', high = slate, guide = 'none') +
    scale_color_manual(values = c('FALSE' = slate, 'TRUE' = 'white')) +
    coord_equal() +
    labs(title = 'Input image') +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5, family = 'Zilla Slab', size = 18))

  kernel_plot =
    ggplot(kernel_df, aes(x = x, y = y, fill = value)) +
    geom_tile(color = 'white', linewidth = 0.9) +
    geom_text(aes(label = value), family = 'Zilla Slab', size = 6, color = slate) +
    scale_fill_gradient2(low = blue, mid = 'white', high = orange, midpoint = 0, guide = 'none') +
    coord_equal() +
    labs(title = '3 x 3 filter') +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5, family = 'Zilla Slab', size = 18))

  feature_plot =
    ggplot(feature_df, aes(x = x, y = y, fill = value)) +
    geom_tile(color = 'white', linewidth = 0.9) +
    geom_text(aes(label = value), family = 'Zilla Slab', size = 6, color = slate) +
    geom_tile(
      data = feature_hi,
      fill = NA,
      color = turquoise,
      linewidth = 1.2
    ) +
    scale_fill_gradient(low = 'grey95', high = turquoise, guide = 'none') +
    coord_equal() +
    labs(title = 'Feature map') +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5, family = 'Zilla Slab', size = 18))

  input_plot + kernel_plot + feature_plot + plot_layout(widths = c(1.25, 1, 1))
}

# Convert flat pixel rows into image tensors for a simple CNN.
make_image_tensor = function(df) {
  n = nrow(df)
  x = array(0, dim = c(n, 1, 28, 28))
  pixel_mat = as.matrix(df[, pixel_cols]) / 255

  for (i in seq_len(n)) {
    x[i, 1, , ] = matrix(pixel_mat[i, ], nrow = 28, byrow = TRUE)
  }

  torch_tensor(x, dtype = torch_float())
}

make_label_tensor = function(df) {
  torch_tensor(as.integer(as.character(df$label)) + 1L, dtype = torch_long())
}

simple_cnn = nn_module(
  'simple_cnn',
  initialize = function() {
    self$conv1 = nn_conv2d(1, 8, kernel_size = 3, padding = 1)
    self$conv2 = nn_conv2d(8, 16, kernel_size = 3, padding = 1)
    self$pool = nn_max_pool2d(kernel_size = 2)
    self$fc1 = nn_linear(16 * 7 * 7, 64)
    self$drop = nn_dropout(p = 0.2)
    self$fc2 = nn_linear(64, 10)
  },
  forward = function(x) {
    x = self$conv1(x)
    x = nnf_relu(x)
    x = self$pool(x)
    x = self$conv2(x)
    x = nnf_relu(x)
    x = self$pool(x)
    x = torch_flatten(x, start_dim = 2)
    x = self$fc1(x)
    x = nnf_relu(x)
    x = self$drop(x)
    self$fc2(x)
  }
)

fit_simple_cnn = function(train_df, epochs = 6, batch_size = 64, learn_rate = 0.003) {
  torch_manual_seed(111)
  net = simple_cnn()
  net$to(dtype = torch_float())
  opt = optim_adam(net$parameters, lr = learn_rate)
  loss_fn = nn_cross_entropy_loss()
  train_ds = tensor_dataset(make_image_tensor(train_df), make_label_tensor(train_df))
  train_dl = dataloader(train_ds, batch_size = batch_size, shuffle = TRUE)

  for (epoch in seq_len(epochs)) {
    net$train()
    coro::loop(for (batch in train_dl) {
      x_batch = batch[[1]]$to(dtype = torch_float())
      y_batch = batch[[2]]$to(dtype = torch_long())
      opt$zero_grad()
      logits = net(x_batch)
      loss = loss_fn(logits, y_batch)
      loss$backward()
      opt$step()
    })
  }

  net
}

predict_simple_cnn = function(net, df, batch_size = 256) {
  label_levels = levels(df$label)
  ds = tensor_dataset(make_image_tensor(df), make_label_tensor(df))
  dl = dataloader(ds, batch_size = batch_size, shuffle = FALSE)
  probs_list = list()

  with_no_grad({
    net$eval()
    coro::loop(for (batch in dl) {
      logits = net(batch[[1]]$to(dtype = torch_float()))
      probs = nnf_softmax(logits, dim = 2)
      probs_list[[length(probs_list) + 1]] = as.matrix(as.array(probs))
    })
  })

  probs = do.call(rbind, probs_list)
  prob_df = as_tibble(probs, .name_repair = 'minimal')
  names(prob_df) = paste0('.pred_', 0:9)

  tibble(
    .pred_class = factor(max.col(probs, ties.method = 'first') - 1L, levels = label_levels)
  ) |>
    bind_cols(prob_df)
}

# Build a manageable MNIST subset for lecture-time fitting.
mnist_small = train_dt |>
  mutate(label = factor(label)) |>
  group_by(label) |>
  slice_sample(n = 500) |>
  ungroup()

# Keep train / validation / test separate so tuning claims are honest.
mnist_split = initial_validation_split(
  mnist_small,
  prop = c(train = 0.6, validate = 0.2),
  strata = label
)

mnist_train = training(mnist_split)
mnist_valid = validation(mnist_split)
mnist_test = testing(mnist_split)
mnist_train_valid = bind_rows(mnist_train, mnist_valid)

# Baseline recipe: modest PCA compression.
rec_base =
  recipe(label ~ ., data = mnist_train) |>
  step_zv(all_predictors()) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 40)

# Tuned recipe: keep richer PCs and append a UMAP embedding.
rec_tuned =
  recipe(label ~ ., data = mnist_train) |>
  step_zv(all_predictors()) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 60) |>
  embed::step_umap(
    all_predictors(),
    num_comp = 8,
    neighbors = 30,
    min_dist = 0.05,
    keep_original_cols = TRUE,
    prefix = 'umap'
  )

# Final refit recipes use train + validation together.
rec_base_tv =
  recipe(label ~ ., data = mnist_train_valid) |>
  step_zv(all_predictors()) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 40)

rec_tuned_tv =
  recipe(label ~ ., data = mnist_train_valid) |>
  step_zv(all_predictors()) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 60) |>
  embed::step_umap(
    all_predictors(),
    num_comp = 8,
    neighbors = 30,
    min_dist = 0.05,
    keep_original_cols = TRUE,
    prefix = 'umap'
  )

# Multinomial logit is the 'no hidden layer' baseline.
logit_spec =
  multinom_reg(penalty = 0) |>
  set_engine('nnet', trace = FALSE, MaxNWts = 20000) |>
  set_mode('classification')

# Starter neural net: explicit ReLU hidden layer.
mlp_start_spec =
  mlp(hidden_units = 15, penalty = 0.01, epochs = 50, activation = 'relu') |>
  set_engine('brulee') |>
  set_mode('classification')

# Tuned neural net: same ReLU activation with a larger network.
mlp_tuned_spec =
  mlp(hidden_units = 50, penalty = 0.001, epochs = 150, activation = 'relu') |>
  set_engine('brulee') |>
  set_mode('classification')
```

---
name: admin
# Admin
## Today

.b[Topics:] We will start an introduction to neural networks:

- anatomy of a single neuron,
- logistic regression as a one-neuron network,
- activation functions and hidden layers,
- a brief overview of training,
- the return of MNIST.

.b[Reading and resources]

- .note[Mostly:] .it[Neural Networks and Deep Learning], Ch. [1](http://neuralnetworksanddeeplearning.com/chap1.html) (some [2](http://neuralnetworksanddeeplearning.com/chap2.html) + [3](http://neuralnetworksanddeeplearning.com/chap3.html)),
- .note[Also helpful:] .it[ISL] Ch. 11,
- .note[Fun, interactive tool:] TensorFlow's [Neural Network Playground](https://playground.tensorflow.org/).

---
layout: true
# Neural networks

---
class: inverse, middle
name: intro

## (Neural) Intuition

---
layout: false
class: clear, middle

```{r, logistic-neuron, echo = FALSE, fig.height = 5, fig.width = 11}
make_single_neuron_plot()
```

.b[Anatomy] .attn[Neurons] are the building blocks of neural networks.
- take .b.slate[in] a weighted sum of inputs, $w'x = w_1 x_1 + w_2 x_2 + \cdots + w_p x_p$,
- apply a .b.slate[bias], $b$, that shifts the sum, $z = b + w'x$,
- pass $z$ through an .b.pink[activation function] to produce .b.orange[output], $a = g(z)$.

---
layout: true

# Neural networks

---
## .slab.it.grey-light[History] It all started with the perceptron

The .attn[perceptron] use a .b[hard-threshold] activiation:

$$
a =
\begin{cases}
1 & \text{if } \color{#314f4f}{w'x} \ge \color{#e64173}{T} \iff \color{#314f4f}{w'x} + \color{#e64173}{b} \ge 0 \\
0 & \text{if } \color{#314f4f}{w'x} < \color{#e64173}{T} \iff \color{#314f4f}{w'x} + \color{#e64173}{b} < 0
\end{cases}
$$

The intuition literally came from neurons firing.

When the .slate.it[weighted sum of inputs] ($\color{#314f4f}{w'x}$) exceeds a .pink[threshold] ($\color{#e64173}{T}$), .b.pink[fire!]

.note[Intuition:] Compare .slate[weighted *evidence*] against the .pink[*burden of proof*].

--

We translate .pink[threshold] into .pink[bias] ($b = -T$) to get a more "standard" form.

---
## What do the weights and bias do?

$$
z = b + \sum_{j = 1}^p w_j x_j
$$

The weighted sum does two jobs:

- the .attn[weights] say which inputs matter and in which direction,
- the .attn[bias] shifts the cutoff left or right.
  
--

Even with its simple structure, the neuron is capable of learning

- to set the weights and bias to find a good decision boundary,
- to $1$ when $z \ge 0$ and $0$ when $z < 0$,
- to separate our two classes.

---
## *Too* binary?

Binary activation may be .pink[intuitive], but it's .purple[not great for learning].

.note[Problem] We need to .purple[*learn*] the weights and bias $(w,b)$.

- When .b.purple[training], we want to .purple[adjust] $w$ and $b$ to reduce errors.
- If we make .purple[small changes] to $w$ or $b$, binary activation may change *a lot*.

$\implies$ training with binary activations can be .purple[unstable] and .purple[inefficient].

---
## *Too* binary?

.note[Solution] Keep what we like about the perceptron

- the .b.slate[weighted-sum of inputs shifted by bias],
- .b.pink[restricting the output] to a certain range (∈[0, 1] for binary classification).

but replace the hard step with a .purple[smooth, differentiable activation].

.qa[Q] Any ideas? .grey-vlight[You already know one!]

--

The .attn[sigmoid neuron] keeps the same weighted-sum idea and replaces the hard step with a differentiable activation—the .attn[sigmoid function].

---
## Logistic regression or a tiny neural network?

<!-- $\implies$ You already know how to build a (very simple) .b[neural network]! -->

.attn[Logistic regression] For a binary outcome $Y$

$$\Pr(Y = 1 \mid X = x) = \sigma(\beta_0 + x'\beta) = \sigma(w'x + b)$$

where $\sigma(z) = \dfrac{1}{1 + e^{-z}}$.

.note[Also] .attn[Sigmoid neuron] with

- a .b.slate[weighted sum of inputs] and a .b.slate[bias],
- a .b.pink[sigmoid] activation function,
- one output.

--

Neural networks are not a sharp break from (logistic) regression.
<br>
They are a .it.slab[generalization].

---
layout: false
class: clear, middle, center

.b.slate[Perceptron] .note[vs.] .b.pink[sigmoid neuron]

```{r, step-v-sigmoid, echo = FALSE}
activation_df |>
  filter(activation %in% c('Step', 'Sigmoid')) |>
  ggplot(aes(x = x, y = y, color = activation)) +
  geom_hline(yintercept = 0, linewidth = 0.25) +
  geom_vline(xintercept = 0, linewidth = 0.25) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c(Step = slate, Sigmoid = red_pink)) +
  labs(
    x = 'z',
    y = 'activation',
    color = '',
  ) +
  theme_minimal(base_size = 18, base_family = 'Fira Sans Condensed') +
  theme(legend.position = 'none')
```

---
layout: false
class: clear, middle, center

.b[Common activation functions.] .grey-vlight[(Match activation to the problem.)]

```{r, activation-functions, echo = FALSE}
activation_df |>
  filter(activation %in% c('Identity', 'Sigmoid', 'Tanh', 'ReLU')) |>
  ggplot(aes(x = x, y = y, color = activation)) +
  geom_hline(yintercept = 0, linewidth = 0.25) +
  geom_vline(xintercept = 0, linewidth = 0.25) +
  geom_line(linewidth = 1.1, show.legend = FALSE) +
  facet_wrap(~activation, scales = 'free_y') +
  scale_color_manual(values = c(
    Identity = slate,
    Sigmoid = red_pink,
    Tanh = orange,
    ReLU = blue
  )) +
  theme_minimal(base_size = 18, base_family = 'Fira Sans Condensed') +
  labs(x = 'z', y = 'g(z)')
```

---
layout: true
# Neural networks

---
## Getting flexible

Adjusting weights and bias of a .b.slate[single neuron]—regardless of the activation function—is only going to get us so far. .grey-vlight[(Flavors of regression.)]

```{r, single-neuron-plot, echo = FALSE, fig.height = 3, fig.width = 12}
plot_nn(
  input = 4,
  max_input = 3,
  hidden = 1,
  output = 1,
  label_input = c('x₁', 'x₂', 'xₚ'),
  label_output = 'ŷ',
  label_layers = c('Inputs', 'Neuron', 'Output')
)
```

.qa[Q:] What else can we do to make our model .it.pink[more flexible]?
<br>.tran.slab[Q:] .grey-vlight[(Especially for complex, highly nonlinear problems?)]

---
layout: false
class: clear, middle

Adding a .attn[hidden layer] of neurons between the .slate[input] and .orange[output] layers

```{r, plot-intro-hidden, echo = FALSE, fig.height = 4, fig.width = 11.5}
plot_nn(
  input = 4,
  max_input = 3,
  hidden = 5,
  output = 1,
  label_input = c('x₁', 'x₂', 'xₚ'),
  label_output = 'ŷ',
  label_layers = c('Inputs', 'Hidden layer', 'Output')
)
```

increases the model's ability to learn .pink[complex relationships] from .slate[x]→.orange[y].

- each .pink[hidden neuron] learns a different transformation of the input. <br> $\color{#e64173}{a_k} = a(w_k'x + b_k)$
- an .orange[output neuron] flexibly combines those transformations. <br> $\color{#FFA500}{\hat{y}} = a_y(w_y'\color{#e64173}{a} + b_y)$

---
layout: false
class: clear, middle

Let's think about building a .b.purple[feed-forward network for MNIST].

.note[Note:] .attn[Feed-forward networks] are simply neural networks where information flows in one direction—from .it.slate[input] to .it.orange[output]. .grey-vlight[(No cycles or feedback loops.)]

---
class: clear, middle, center

.b[A feedforward network] with one .pink[hidden layer.].

```{r, network-plot1, echo = FALSE, fig.height = 7.5, fig.width = 11.5}
plot_nn(
  input = 784,
  hidden = 6,
  output = 10,
  max_input = 4,
  max_output = 4,
  label_input = c('pixel₁', 'pixel₂', 'pixel₃', 'pixel₇₈₄'),
  label_output = 0:9,
  label_layers = c('Input layer', 'Hidden layer', 'Output layer')
)
```

---
class: clear, middle

There's really no reason to stop at .it[one] .pink[hidden layer]!

---
class: clear, middle, center

.b[A feedforward network] with two .pink[hidden layers]!

```{r, network-plot2, echo = FALSE, fig.height = 7.5, fig.width = 11.5}
plot_nn(
  input = 784,
  hidden = c(6, 6),
  output = 10,
  max_input = 4,
  max_output = 4,
  label_input = c('pixel₁', 'pixel₂', 'pixel₃', 'pixel₇₈₄'),
  label_output = 0:9,
  label_layers = TRUE
)
```

---
class: clear, middle, center

.b[A feedforward network] with three .pink[hidden layers]! .grey-vlight[(Getting deep!)]

```{r, network-plot3, echo = FALSE, fig.height = 7.5, fig.width = 11.5}
plot_nn(
  input = 784,
  hidden = c(6, 6, 6),
  output = 10,
  max_input = 4,
  max_output = 4,
  label_input = c('pixel₁', 'pixel₂', 'pixel₃', 'pixel₇₈₄'),
  label_output = 0:9,
  label_layers = TRUE
)
```

---
layout: true
# Neural networks

---
## Neuron stacking in hidden layers

.note[Let's be clear:] Hidden layers are not about .it[more coefficients].

The .it[power of .pink[hidden layers]] comes from their ability to<br>.it[learn increasingly complex features] built from simpler ones.

- Early neurons can respond to simple patterns;
- later neurons can combine those neurons into more complex patterns;
- every layer layers complexity on top of the previous layers.
  
Adding .attn[depth] (hidden layers) allows NNs to learn more levels of .note[abstraction].
<br>
.grey-vlight[(It also adds more parameters to learn.)]

---
layout: false
class: clear

```{r, simple-hidden1, echo = FALSE, fig.height = 3, fig.width = 12}
plot_nn(
  input = 2,
  hidden = 2,
  output = 1,
  label_input = c('x₁', 'x₂'),
  label_output = 'ŷ',
  label_layers = c('Inputs', 'Hidden layer', 'Output')
)
```

To see how .note[depth] and .note[activations] affect network flexibility and output,
<br> let's write out the math for this simple neural network.

- .slate[2 inputs],
- .pink[2 hidden neurons], and
- .orange[1 output].

---
class: clear

```{r, simple-hidden2, echo = FALSE, fig.height = 3, fig.width = 12}
plot_nn(
  input = 2,
  hidden = 2,
  output = 1,
  label_input = c('x₁', 'x₂'),
  label_output = 'ŷ',
  label_layers = c('Inputs', 'Hidden layer', 'Output')
)
```

In the .pink[hidden layer], neuron $\color{#e64173}{a_j}$ learns a transformation $g$ of the .it.slate[input]:
$$\color{#e64173}{a_j} = g\big(w^1_{j,1} \color{#314f4f}{x_1} + w^1_{j,2} \color{#314f4f}{x_2} + b^1_j\big)$$
--

The .orange[output] $\color{#FFA500}{\hat{y}}$ is a transformation $h$ of the .it.pink[hidden layer output]:
$$\color{#FFA500}{\hat{y}} = h\big(w^2_{1,1} \color{#e64173}{a_1} + w^2_{1,2} \color{#e64173}{a_2} + b^2_1\big)$$
--

Combining the preceding two steps...
\begin{align*}
\color{#FFA500}{\hat{y}} & = h\Big(w^2_{1,1} g\big(w^1_{1,1} \color{#314f4f}{x_1} + w^1_{1,2} \color{#314f4f}{x_2} + b^1_1\big) + w^2_{1,2} g\big(w^1_{2,1} \color{#314f4f}{x_1} + w^1_{2,2} \color{#314f4f}{x_2} + b^1_2\big) + b^2_1\Big)
\end{align*}

---
class: clear

\begin{align*}
\color{#FFA500}{\hat{y}} & = h\Big(w^2_{1,1} g\big(w^1_{1,1} \color{#314f4f}{x_1} + w^1_{1,2} \color{#314f4f}{x_2} + b^1_1\big) + w^2_{1,2} g\big(w^1_{2,1} \color{#314f4f}{x_1} + w^1_{2,2} \color{#314f4f}{x_2} + b^1_2\big) + b^2_1\Big)
\end{align*}
--

.note[Case 1:] Suppose $g$ and $h$ are both the .attn[identity function] $f(z) = z$.

--

$\color{#FFA500}{\hat{y}} = \Big(w^2_{1,1} \big(w^1_{1,1} \color{#314f4f}{x_1} + w^1_{1,2} \color{#314f4f}{x_2} + b^1_1\big) + w^2_{1,2} \big(w^1_{2,1} \color{#314f4f}{x_1} + w^1_{2,2} \color{#314f4f}{x_2} + b^1_2\big) + b^2_1\Big)$

--

$\color{#FFFFFF}{\hat{y}} = w^2_{1,1} w^1_{1,1} \color{#314f4f}{x_1} + w^2_{1,1} w^1_{1,2} \color{#314f4f}{x_2} + w^2_{1,1} b^1_1 + w^2_{1,2} w^1_{2,1} \color{#314f4f}{x_1} + w^2_{1,2} w^1_{2,2} \color{#314f4f}{x_2} + b^1_2 + b^2_1$

--

$\color{#FFFFFF}{\hat{y}} = \big(w^2_{1,1} w^1_{1,1} + w^2_{1,2} w^1_{2,1}\big) \color{#314f4f}{x_1} + \big(w^2_{1,1} w^1_{1,2} + w^2_{1,2} w^1_{2,2}\big) \color{#314f4f}{x_2} + \big(w^2_{1,1} b^1_1 + w^2_{1,2} b^1_2 + b^2_1\big)$

--

$\color{#FFFFFF}{\hat{y}} = \beta_{0} + \beta_{1} \color{#314f4f}{x_1} + \beta_{2} \color{#314f4f}{x_2}$

--

A lot of work to get back to a .b[linear model]!

--

.note[Case 2:] Suppose $g$ and $h$ are .attn[linear]/.attn[affine] functions.

--

.grey-vlight[(insert math)] ... Still a .b[linear model]!

---
layout: true
# Neural networks

---
## Hidden layers need nonlinearity

So hidden layers only "work" when we add .b.pink[nonlinear activations].

\begin{align*}
\color{#FFA500}{\hat{y}} & = h\Big(w^2_{1,1} g\big(w^1_{1,1} \color{#314f4f}{x_1} + w^1_{1,2} \color{#314f4f}{x_2} + b^1_1\big) + w^2_{1,2} g\big(w^1_{2,1} \color{#314f4f}{x_1} + w^1_{2,2} \color{#314f4f}{x_2} + b^1_2\big) + b^2_1\Big)
\end{align*}

If $g$ and $h$ are nonlinear, then $\color{#FFA500}{\hat{y}}$ is 

- a .orange[nonlinear combination] of the .pink[hidden-layer activations], which are
- .slate[nonlinear combinations of the inputs].

As you add more hidden layers, you *layer* more nonlinear transformations.

.note[Note] The .orange[output] activation function does not have to be nonlinear for the hidden layers to "work". .grey-vlight[(No requirement that *g* and *h* match either.)]

---
layout: false
class: clear, middle

.attn[Universal approximation theorem]

A feedforward network with one hidden layer can approximate *any* continuous function on a compact domain, given enough neurons in the hidden layer.

.grey-vlight[(It doesn't say how many neurons we need.)]

---
layout: true
# Neural networks

---
## Welcome to the playground

TensorFlow's [*Playground*](https://playground.tensorflow.org/) is a fun way to see NNs in action.

Play with the

- activation function,
- number of hidden layers and neurons,
- learning rate,
- amount of regularization.

Watch the .b[boundary] change and the .b[representation] of hidden layers.

---
class: inverse, middle
name: training

## How the network learns

---
## The loss function still runs the show

.note[As always:] We choose weights and biases to minimize a .attn[loss function].

- .note[regression]: MSE, MAE, Huber .grey-vlight.it[(linear output)]

- .note[binary]: log loss (binary cross-entropy). .grey-vlight.it[(sigmoid output)]
$$C = - \big[y \log(\hat{p}) + (1 - y) \log(1 - \hat{p})\big]$$

- .note[multiclass]: multiclass cross-entropy .grey-vlight.it[(softmax output)]
$$C = -\sum_{k = 1}^K y_k \log \hat p_k$$

--

.note[As before:] We can .attn[penalize] (.attn[regularize]) the weights to prevent overfitting.

---
## Optimization

We don't have time to cover the details of optimization; here's the *gist*:

- .attn[gradient descent] move parameters "downhill" fastest by following the negative gradient of the loss function
$$\theta \leftarrow \theta - \eta \nabla_\theta C(\theta)$$
- .attn[gradient] how each weight/bias changes the loss
- .attn[backpropagation] efficient chain rule to compute those gradients
- .attn[mini-batch] one small chunk of data used for one update
- .attn[epoch] one full pass through all training observations

It's a lot. In fact, this *gist* barely scratches the surface.

The key is efficiently computing the gradients—backpropagation. 

---
class: inverse, middle
name: mnist

## MNIST application

---
## Why MNIST again?

MNIST is a natural neural-net example

- high-dimensional inputs (784 predictors for each image),
- 10 output classes,
- nonlinear relationships between pixels and classes,
- lots of variation in handwriting.
  
Let's see how well a neural network can learn from the raw pixel data.

---
layout: false
class: clear, middle

The raw inputs are still images.

```{r, mnist-examples, echo = FALSE}
idx = c(2, 1, 17, 8, 4, 9, 22, 7, 11)
plots = lapply(idx, function(i) {
  plot_digit(train_dt[i, ])
})

wrap_plots(plots, ncol = 3) +
  plot_layout(guides = 'collect') &
  theme(
    legend.position = 'bottom',
    legend.key.width = unit(2, 'cm'),
    plot.title = element_text(size = 15, hjust = 0.5)
  )
```

---
layout: true
# Neural networks

---
## One image, many outputs

MNIST is a .attn[multi-output classification] problem.

For each image, the output layer should produce:
$$
\big(
\hat p_0,\,
\hat p_1,\,
\ldots,\,
\hat p_9
\big)
$$

If we remove the hidden layer and keep only the softmax output,
$$\sigma(z_i) = \dfrac{e^{z_i}}{\sum_{k = 0} e^{z_k}}$$
we are back at .attn[multinomial logistic regression].

---
exclude: true

## Fit the models

```{r, mnist-fit, cache = TRUE}
# Fit each workflow on the training split.
fit_logit_valid =
  workflow() |>
  add_recipe(rec_base) |>
  add_model(logit_spec) |>
  fit(mnist_train)

fit_mlp_start_valid =
  workflow() |>
  add_recipe(rec_base) |>
  add_model(mlp_start_spec) |>
  fit(mnist_train)

fit_mlp_tuned_valid =
  workflow() |>
  add_recipe(rec_tuned) |>
  add_model(mlp_tuned_spec) |>
  fit(mnist_train)

fit_cnn_valid =
  fit_simple_cnn(mnist_train, epochs = 6)

# Collect validation predictions for model comparison.
pred_logit_valid =
  bind_cols(
    mnist_valid |> select(label),
    predict(fit_logit_valid, mnist_valid)
  )

pred_mlp_start_valid =
  bind_cols(
    mnist_valid |> select(label),
    predict(fit_mlp_start_valid, mnist_valid)
  )

pred_mlp_tuned_valid =
  bind_cols(
    mnist_valid |> select(label),
    predict(fit_mlp_tuned_valid, mnist_valid)
  )

pred_cnn_valid =
  bind_cols(
    mnist_valid |> select(label),
    predict_simple_cnn(fit_cnn_valid, mnist_valid)
  )

validation_compare = tibble(
  model = c('Multinomial logit', 'Starter MLP', 'Tuned MLP', 'Simple CNN'),
  accuracy = c(
    pred_logit_valid |> metrics(truth = label, estimate = .pred_class) |> filter(.metric == 'accuracy') |> pull(.estimate),
    pred_mlp_start_valid |> metrics(truth = label, estimate = .pred_class) |> filter(.metric == 'accuracy') |> pull(.estimate),
    pred_mlp_tuned_valid |> metrics(truth = label, estimate = .pred_class) |> filter(.metric == 'accuracy') |> pull(.estimate),
    pred_cnn_valid |> metrics(truth = label, estimate = .pred_class) |> filter(.metric == 'accuracy') |> pull(.estimate)
  )
)

# Refit the strongest models on train + validation.
fit_logit_test =
  workflow() |>
  add_recipe(rec_base_tv) |>
  add_model(logit_spec) |>
  fit(mnist_train_valid)

fit_mlp_tuned_test =
  workflow() |>
  add_recipe(rec_tuned_tv) |>
  add_model(mlp_tuned_spec) |>
  fit(mnist_train_valid)

fit_cnn_test =
  fit_simple_cnn(mnist_train_valid, epochs = 6)

# Evaluate on the held-out test split.
pred_logit_test =
  bind_cols(
    mnist_test |> select(label),
    predict(fit_logit_test, mnist_test)
  )

pred_mlp_tuned_test =
  bind_cols(
    mnist_test |> select(label),
    predict(fit_mlp_tuned_test, mnist_test),
    predict(fit_mlp_tuned_test, mnist_test, type = 'prob')
  )

pred_cnn_test =
  bind_cols(
    mnist_test |> select(label),
    predict_simple_cnn(fit_cnn_test, mnist_test)
  )

test_compare = tibble(
  model = c('Multinomial logit', 'Tuned MLP', 'Simple CNN'),
  accuracy = c(
    pred_logit_test |> metrics(truth = label, estimate = .pred_class) |> filter(.metric == 'accuracy') |> pull(.estimate),
    pred_mlp_tuned_test |> metrics(truth = label, estimate = .pred_class) |> filter(.metric == 'accuracy') |> pull(.estimate),
    pred_cnn_test |> metrics(truth = label, estimate = .pred_class) |> filter(.metric == 'accuracy') |> pull(.estimate)
  )
)

# Build a confusion matrix and one probability-vector example for the CNN.
cm_cnn_test =
  pred_cnn_test |>
  count(label, .pred_class)

example_prob_row =
  pred_cnn_test |>
  mutate(case_id = row_number()) |>
  rowwise() |>
  mutate(max_prob = max(c_across(matches('^\\.pred_[0-9]+$')))) |>
  ungroup() |>
  filter(label == .pred_class) |>
  slice_max(order_by = max_prob, n = 1)

example_case =
  mnist_test |> slice(example_prob_row$case_id)

example_prob_plot_df =
  example_prob_row |>
  select(matches('^\\.pred_[0-9]+$')) |>
  pivot_longer(
    everything(),
    names_to = 'digit',
    values_to = 'probability'
  ) |>
  mutate(digit = gsub('.pred_', '', digit, fixed = TRUE))
```

---
exclude: true
## Validation results

The first pass looks like this:

- multinomial logit: .attn[`r scales::percent(validation_compare$accuracy[validation_compare$model == 'Multinomial logit'], accuracy = 0.1)`]
- starter MLP: .attn[`r scales::percent(validation_compare$accuracy[validation_compare$model == 'Starter MLP'], accuracy = 0.1)`]
- tuned MLP: .attn[`r scales::percent(validation_compare$accuracy[validation_compare$model == 'Tuned MLP'], accuracy = 0.1)`]
- simple CNN: .attn[`r scales::percent(validation_compare$accuracy[validation_compare$model == 'Simple CNN'], accuracy = 0.1)`]

So the lesson is not .it[any hidden layer wins].

The lesson is that .attn[capacity, preprocessing, and architecture all matter].

---
## Held-out test results

After tuning and fitting on the training data, here are the test accuracies

- .note[multinomial logit:] .attn[`r scales::percent(test_compare$accuracy[test_compare$model == 'Multinomial logit'], accuracy = 0.1)`]
- .note[NN (1 hidden layer with 50 units):].super.pink[†] .attn[`r scales::percent(test_compare$accuracy[test_compare$model == 'Tuned MLP'], accuracy = 0.1)`]
- .note[convolutional NN:].super.orange[††] .attn[`r scales::percent(test_compare$accuracy[test_compare$model == 'Simple CNN'], accuracy = 0.1)`]

Nonlinearity and achitectural choices matter
<br>
...even for a "simple" problem like MNIST.
<br>
.grey-vlight[(Simple is relative.)]

.footnote[.pink[†] You can use the [`brulee` package](https://brulee.tidymodels.org/reference/brulee_mlp.html) to fit a neural net using a `tidymodels` `recipe`. You'll also need to install [`torch`](https://torch.mlverse.org/).<br>.orange[††] I know, we didn't get to CNNs. They're an approach that *slides filters* across the input to detect local patterns.]

---
class: inverse, middle
name: wrap

## Wrap-up

---
## Main takeaways

NNs are a powerful class of models that extends many prior topics

- .attn[logistic regression] is a one-neuron network with sigmoid activation,
- .attn[hidden layers] allow NNs to learn complex representations of the data,
- .attn[nonlinear activations] are essential for hidden layers,
- .attn[training] involves min. loss via gradient descent and backprop.,
- .attn[penalization] mirrors bias-variance intution from regularized regression.

.note[As always] Careful tuning and construction are key to performance.