---
title: "EC524W25: Lab 03"
subtitle: "Introduction to `tidymodels`"
author: "Jose Rojas-Fallas"
format: 
  html:
    embed-resources: true
---

```{r knitr-opts, echo=FALSE, message=FALSE}
# Code chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

## Lab Goals

1. **Motivate**: Why using `tidymodels` may be good practice
2. **Understanding Preprocessing**: Learn the basics of preprocessing data using `tidymodels`

### First Step

Load our necessary packages

``` {r packages}
library(pacman)
p_load(tidyverse, modeldata, skimr, janitor, kknn, magrittr, tidymodels)
```

### Load in Our Data

We will be using credit data from the `modeldata` package 

``` {r modeldata}
# Load credit data
data(credit_data)
# Assign it a new name
credit_df <- credit_data
```

Let's see what it looks like

``` {r data-glimpse}
# Take a glimpse at the data
credit_df %>% glimpse()
```

Gross, let's clean it up a bit

``` {r clean-data}
# "Fix" the variable names
credit_df %<>% clean_names()
```

And a deeper look at the data gives us:

``` {r data-skim}
credit_df %>% skim()
```

**Notice there is missing data, that'll come up later**

### Creating Our Outcome Variable

We want to create an indicator for individuals' whose credit `status` is ***good***

We can keep it simple and use `tidyverse` to do that and remove the "old" `status` variable:

``` {r credit-good-outcome}
# Create a dummy variable for "good" credit status
# Create a dummy variable for "good" credit status
credit_df %<>% 
    # There are many ways to do this. You could also use ifelse() or case_when().
    mutate(status_good = 1 * (status == "good")) %>% 
    # Drop the old status variable
    select(-status)
```

### Introducing `tidymodels`

Imagine we want to clean **each variable**. 
Then our `tidyverse` approach will take a lot of effort for each one. 

This is where `tidymodels` helps us! 
By using `recipes` we can set the stage for our data cleaning.

#### The How of `tidymodels`

`tidymodels` workflow begins with a `recipe`, where we tell R what role each variable plays

Our first `recipe`:

``` {r recipe-01, message=TRUE}
# Define the recipe: What is an outcome and what is an explanatory variable
recipe_all <- recipe(status_good ~ ., data = credit_df)

# What is it?
recipe_all
```

R now knows what is what in the data.

So we can start cleaning/transforming the data:

We can apply whatever steps we want to clean/pre-process/manipulate our data. 
The [`tidymodels` universe](https://recipes.tidymodels.org/reference/index.html){target="_blank"} has many possible steps for:
:::

:::: {.columns}
::: {.column .small .align-center}
[Imputation]{.hp} 

- `step_impute_mean()` 
- `step_impute_mode()` 
- `step_impute_knn()` 
- `step_impute_bag()`
:::
::: {.column .small .align-center}
[Transformation]{.hp}

- `step_log()` 
- `step_poly()` 
- `step_mutate()` 
- `step_interact()`
:::
::: {.column .small .align-center}
[Dummies & Discretization]{.hp}

- `step_dummy()` 
- `step_num2factor()` 
- `step_discretize()` 
- `step_cut()`
:::
::: {.column .small .align-center}
[Many More]{.hp}

- `step_center()` 
- `step_normalize()` 
- `step_date()` 
- `step_lag()`
:::
::::

To apply any of these, you need to also tell the function `step_*()` [which variables]{.h} you want it to target. 

:::: {.columns}
::: {.column .small}
- [All:]{.hp} `all_vars()`
- [Role:]{.hp} `all_predictors(), all_outcomes(), has_role()`
:::

::: {.column .small}
- [Type:]{.hp} `all_nominal(), all_numeric()`
- [Variable Names:]{.hp} Listing the name of the variables or [using selectors](https://recipes.tidymodels.org/reference/selections.html){target="_blank"} `starts_with(), contains()`, etc.
:::
::::

Let's practice: try writing a **simple imputation for all variables that are (1) predictors AND (2) numeric**

``` {r imputation-ex, message=TRUE}
# Mean imputation for all numeric predictors

```

**Note: We haven't done much yet, we still do NOT have a dataframe**

This is where `prep()` and `juice()` enter the picture

``` {r prep-ex-01, message=TRUE}
# Add prep()
recipe_all %>%
    step_impute_mean(all_predictors() & all_numeric()) %>%
    prep()
```

Notice that the output is still a `recipe`, but now it has been **trained** the preprocessing operations on the *training data*. 
It now is aware of the number of observations and the number of incomplete rows.

Let's throw in (a dash of) `juice()` to get a cleaned dataset

``` {r prep-ex-02, message=TRUE}
# Add juice()
credit_clean <- recipe_all %>% 
    step_impute_mean(all_predictors() & all_numeric()) %>%
    prep() %>%
    juice()

# Skim it
credit_clean %>% skim()
```

Nice! There are no more **missing values for our numeric predictors**.

But there are still **some missing values in our categorical variables**

We gotta fix those too!

Let's define our **preprocessing** as: 

[_1._]{.hp} Impute the numeric predictors using the [means](https://recipes.tidymodels.org/reference/step_meanimpute.html){target="_blank"}

[_2._]{.hp} Impute the categorical predictors using [knn](https://recipes.tidymodels.org/reference/step_knnimpute.html){target="_blank"} (k = 5)

[_3._]{.hp} Create [dummy variables](https://recipes.tidymodels.org/reference/step_dummy.html){target="_blank"} for all categorical predictors

- [Note: by default, `step_dummy()` drops the original column]{.note .tiny}

[_4._]{.hp} [Interact](https://recipes.tidymodels.org/reference/step_interact.html){target="_blank"} the outputted `home_` dummies with `income`

#### Putting it All Together 

Go ahead and take a crack at it. 
Make sure you do all of the **preprocessing** steps above:

``` {r preprocessing-all}
# Start by defining the recipe again (pretend its gone)

# Preprocessing Steps
credit_clean <- recipe_all %>%
    # 1. Impute the numerical predictors using the means

    # 2. Impute the categorical variables using KNN (k = 5)

    # 3. Create dummies for categorical variables

    # 4. Interact the created dummies with `income`

    # Prep

    # And Juice
```

If you did it correctly, you should have:

- No more missing values

- Created dummies 

- Created interactions between `home` and `income`

Next, we will explore modeling using `tidymodels`